---
title: "Clustering-Based Sector Rotation Strategy"
subtitle: "Active Fund Management via K-Means Regime Detection and Kalman Signal Filtering"
author: "FIN 451 — Financial Analytics and Trading"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    theme: cosmo
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 6
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
# ── Package loading ──────────────────────────────────────────────────────────
# Packages used via explicit namespace notation throughout (pkg::fn()).
# require() is used here so a single missing package does not abort the render.

pkgs <- c(
  "tidyverse", "tidyquant", "tidymodels",
  "ggplot2", "plotly", "patchwork", "gganimate",
  "cluster", "factoextra", "moments",
  "dlm", "KFAS",
  "gt", "scales", "lubridate", "rlang", "purrr", "zoo", "TTR"
)

invisible(lapply(pkgs, function(p) {
  if (!requireNamespace(p, quietly = TRUE)) {
    message("Package not installed: ", p,
            " — install with install.packages('", p, "')")
  } else {
    suppressPackageStartupMessages(require(p, character.only = TRUE))
  }
}))
```

---

# 1. Problem Framing {#sec-framing}

<!-- ─────────────────────────────────────────────────────────────────────────
     SECTION 1: PROBLEM FRAMING
     Goal: Establish the investment mandate and motivate why an investor
           should pay for active management instead of simply holding SPY.
     Content to write:
       • Opening hook — the passive vs. active management debate
       • "Markets as ecosystems" mental model: regimes shift, sectors rotate
       • Investment mandate: long Growth, short Defensive, flat Neutral
       • Strategy objectives: superior risk-adjusted return vs. SPY benchmark
       • Brief road-map of how the rest of the document answers the question
     ──────────────────────────────────────────────────────────────────────── -->

<!-- TODO: Write narrative prose for Problem Framing -->

---

# 2. Mental Model {#sec-mental-model}

<!-- ─────────────────────────────────────────────────────────────────────────
     SECTION 2: MENTAL MODEL
     Goal: Give the reader a single, clear visual of the full decision logic
           before any code or math is introduced.
     Content to write:
       • Mermaid flowchart covering the full pipeline:
           Data → Feature Engineering → Kalman Filter → Clustering →
           Regime Label → Entry Signal (delta + adaptive theta) →
           Exit Signals (innovation z-score OR cluster flip) →
           Portfolio Construction → Performance Evaluation
       • Symmetric long/short structure must be visible in the diagram
       • All entry and exit conditions clearly labelled on diagram edges
       • Written plain-English decision rules below the diagram
         (no math yet — that is Section 7)
     ──────────────────────────────────────────────────────────────────────── -->

<!-- TODO: Insert Mermaid flowchart block -->

<!-- TODO: Write plain-English decision rules below diagram -->

---

# 3. Data {#sec-data}

## 3.1 Asset Universe & Periods

All data is sourced exclusively via API using `tidyquant::tq_get()`. No local files, CSV, or Excel imports are used anywhere in this document.

The strategy trades **11 SPDR sector ETFs**. **SPY is included as a passive benchmark only and is never traded.**

| Role | Tickers |
|------|---------|
| Benchmark (not traded) | SPY |
| Tradeable universe | XLB, XLC, XLE, XLF, XLI, XLK, XLP, XLRE, XLV, XLU, XLY |

> **Note on XLC:** The Communication Services ETF (XLC) launched in June 2018. Observations prior to that date are absent from the raw pull and are handled gracefully — affected rows are simply absent rather than filled or imputed. All downstream calculations are conditioned on data availability per ticker.

```{r data-constants}
# ── Asset universe ───────────────────────────────────────────────────────────
BENCHMARK  <- "SPY"
SECTORS    <- c("XLB", "XLC", "XLE", "XLF", "XLI",
                "XLK", "XLP", "XLRE", "XLV", "XLU", "XLY")
ALL_TICKERS <- c(BENCHMARK, SECTORS)

# ── Period boundaries ────────────────────────────────────────────────────────
TRAIN_START <- lubridate::ymd("2000-06-01")
TRAIN_END   <- lubridate::ymd("2024-12-31")
TEST_START  <- lubridate::ymd("2025-01-01")
TEST_END    <- lubridate::ymd("2026-03-31")

# ── Split boundary label (for chart annotation) ──────────────────────────────
SPLIT_DATE  <- TEST_START
```

## 3.2 Data Pull

```{r data-pull}
# ── Pull daily adjusted prices for all 12 tickers ───────────────────────────
raw_prices <- tidyquant::tq_get(
  ALL_TICKERS,
  get  = "stock.prices",
  from = TRAIN_START,
  to   = TEST_END
) |>
  dplyr::select(symbol, date, adjusted) |>
  dplyr::rename(price = adjusted) |>
  dplyr::arrange(symbol, date)

# ── Sanity check: confirm retrieval ─────────────────────────────────────────
raw_prices |>
  dplyr::group_by(symbol) |>
  dplyr::summarise(
    first_date = min(date),
    last_date  = max(date),
    n_obs      = dplyr::n(),
    .groups    = "drop"
  ) |>
  dplyr::arrange(symbol)
```

## 3.3 Train / Test Split

```{r data-split}
# ── Training window: June 2000 – December 2024 ───────────────────────────────
train_data <- raw_prices |>
  dplyr::filter(date >= TRAIN_START, date <= TRAIN_END)

# ── Test window: January 2025 – March 2026 ───────────────────────────────────
test_data <- raw_prices |>
  dplyr::filter(date >= TEST_START, date <= TEST_END)
```

## 3.4 Indexed Price Chart

Prices are indexed to **100 at each ticker's first available trading date** so that all series are visually comparable regardless of absolute price level. The dashed vertical line marks the boundary between the training and test periods.

```{r data-index}
# ── Index each ticker to 100 at its own first available date ─────────────────
prices_indexed <- raw_prices |>
  dplyr::group_by(symbol) |>
  dplyr::mutate(
    base_price    = dplyr::first(price),
    indexed_price = price / base_price * 100
  ) |>
  dplyr::ungroup() |>
  dplyr::select(symbol, date, indexed_price)
```

```{r chart-indexed, fig.alt="Interactive chart of all 12 ETFs indexed to 100 at first available date"}
# ── Base ggplot ───────────────────────────────────────────────────────────────
p_indexed <- ggplot2::ggplot(
  prices_indexed,
  ggplot2::aes(x = date, y = indexed_price, colour = symbol, group = symbol)
) +
  ggplot2::geom_line(linewidth = 0.45, alpha = 0.85) +
  ggplot2::geom_vline(
    xintercept = as.numeric(SPLIT_DATE),
    linetype   = "dashed",
    colour     = "grey30",
    linewidth  = 0.6
  ) +
  ggplot2::annotate(
    "text",
    x      = SPLIT_DATE + lubridate::days(60),
    y      = max(prices_indexed$indexed_price, na.rm = TRUE) * 0.97,
    label  = "← Train  |  Test →",
    size   = 3,
    hjust  = 0,
    colour = "grey30"
  ) +
  ggplot2::scale_colour_manual(
    values = scales::hue_pal()(length(ALL_TICKERS))
  ) +
  ggplot2::scale_x_date(
    date_breaks = "2 years",
    date_labels = "%Y"
  ) +
  ggplot2::scale_y_continuous(
    labels = scales::comma_format(suffix = "")
  ) +
  ggplot2::labs(
    title    = "SPDR Sector ETFs — Indexed to 100 at First Available Date",
    subtitle = "Daily adjusted closing prices  |  Dashed line = Train/Test split (Jan 2025)",
    x        = NULL,
    y        = "Indexed Price (base = 100)",
    colour   = "Ticker",
    caption  = "Source: Yahoo Finance via tidyquant::tq_get()"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    legend.position  = "right",
    panel.grid.minor = ggplot2::element_blank(),
    plot.title       = ggplot2::element_text(face = "bold")
  )

# ── Wrap in plotly for interactivity ─────────────────────────────────────────
plotly::ggplotly(p_indexed, tooltip = c("colour", "x", "y")) |>
  plotly::layout(
    hovermode = "x unified",
    legend    = list(title = list(text = "<b>Ticker</b>"))
  )
```

## 3.5 Summary Statistics

```{r summary-stats}
# ── Compute per-ETF summary statistics on the full raw series ────────────────
etf_summary <- raw_prices |>
  dplyr::group_by(symbol) |>
  dplyr::mutate(
    daily_return = (price / dplyr::lag(price)) - 1
  ) |>
  dplyr::summarise(
    `First Date`      = min(date),
    `Last Date`       = max(date),
    `Observations`    = dplyr::n(),
    `Mean Dly Ret`    = mean(daily_return, na.rm = TRUE),
    `Ann. Vol`        = sd(daily_return, na.rm = TRUE) * sqrt(252),
    `Min Price`       = min(price, na.rm = TRUE),
    `Max Price`       = max(price, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    `Role` = dplyr::if_else(symbol == "SPY", "Benchmark", "Tradeable")
  ) |>
  dplyr::arrange(dplyr::desc(`Role`), symbol)

# ── Render as formatted gt table ─────────────────────────────────────────────
etf_summary |>
  gt::gt() |>
  gt::tab_header(
    title    = "Asset Universe — Summary Statistics",
    subtitle = "Full sample: June 2000 – March 2026 | Adjusted closing prices"
  ) |>
  gt::fmt_date(columns = c(`First Date`, `Last Date`), date_style = "yMd") |>
  gt::fmt_number(columns = `Observations`, use_seps = TRUE, decimals = 0) |>
  gt::fmt_percent(columns = c(`Mean Dly Ret`, `Ann. Vol`), decimals = 2) |>
  gt::fmt_currency(columns = c(`Min Price`, `Max Price`), currency = "USD",
                   decimals = 2) |>
  gt::tab_style(
    style     = gt::cell_fill(color = "#f0f4ff"),
    locations = gt::cells_body(rows = `Role` == "Benchmark")
  ) |>
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::cols_label(symbol = "Ticker") |>
  gt::tab_footnote(
    footnote  = "XLC launched June 2018; prior observations are absent, not imputed.",
    locations = gt::cells_body(columns = `First Date`,
                               rows    = symbol == "XLC")
  ) |>
  gt::tab_footnote(
    footnote  = "SPY is a benchmark only and is excluded from all portfolio positions.",
    locations = gt::cells_body(columns = `Role`,
                               rows    = symbol == "SPY")
  ) |>
  gt::opt_table_font(font = gt::google_font("IBM Plex Sans")) |>
  gt::opt_stylize(style = 1, color = "blue")
```

---

# 4. Feature Engineering {#sec-features}

## 4.1 Technical Indicators

Four technical indicators are computed daily for each of the 11 tradeable sector ETFs using the `TTR` package. These are the raw observations that will subsequently be fed into the Kalman filter. SPY is excluded from feature engineering — it is never traded.

| Indicator | Formula / Window | Economic Intuition |
|-----------|------------------|--------------------|
| **RSI** | 14-day Wilder RSI | Momentum oscillator bounded [0, 100]; values above 70 indicate overbought conditions, below 30 oversold |
| **MACD** | EMA(12) − EMA(26), signal EMA(9) | Captures trend acceleration; the MACD line crossing its signal line signals directional shifts |
| **ROC** | $(P_t - P_{t-10}) / P_{t-10}$ | 10-day rate of change; measures near-term price momentum |
| **Bollinger %B** | $(P_t - \text{Lower}) / (\text{Upper} - \text{Lower})$ | Position of price within 20-day, 2-SD bands; >1 or <0 flags breakouts |

```{r compute-indicators}
# ── Compute all four indicators for every tradeable ETF ──────────────────────
# Input: train_data and test_data from Section 3 (adjusted daily prices)
# Output: raw_indicators — long-format tibble, one row per ETF × date × indicator

compute_indicators <- function(df) {
  prices <- df$price

  rsi_vals  <- TTR::RSI(prices, n = 14)

  macd_obj  <- TTR::MACD(prices, nFast = 12, nSlow = 26, nSig = 9)
  macd_vals <- macd_obj[, "macd"] - macd_obj[, "signal"]

  roc_vals  <- TTR::ROC(prices, n = 10, type = "discrete")

  bb_obj    <- TTR::BBands(prices, n = 20, sd = 2)
  pctb_vals <- bb_obj[, "pctB"]

  tibble::tibble(
    date      = df$date,
    RSI       = as.numeric(rsi_vals),
    MACD      = as.numeric(macd_vals),
    ROC       = as.numeric(roc_vals),
    BollingerB = as.numeric(pctb_vals)
  )
}

# Apply to training data for each sector ETF (not SPY)
raw_indicators_train <- train_data |>
  dplyr::filter(symbol %in% SECTORS) |>
  dplyr::group_by(symbol) |>
  dplyr::group_modify(~ compute_indicators(.x)) |>
  dplyr::ungroup() |>
  tidyr::pivot_longer(
    cols      = c(RSI, MACD, ROC, BollingerB),
    names_to  = "indicator",
    values_to = "raw_value"
  ) |>
  dplyr::filter(!is.na(raw_value))

# Apply to test data
raw_indicators_test <- test_data |>
  dplyr::filter(symbol %in% SECTORS) |>
  dplyr::group_by(symbol) |>
  dplyr::group_modify(~ compute_indicators(.x)) |>
  dplyr::ungroup() |>
  tidyr::pivot_longer(
    cols      = c(RSI, MACD, ROC, BollingerB),
    names_to  = "indicator",
    values_to = "raw_value"
  ) |>
  dplyr::filter(!is.na(raw_value))
```

## 4.2 Kalman Filter

### 4.2.1 State-Space Model

Raw technical indicators are noisy — they react to microstructure noise, bid-ask bounce, and short-lived price dislocations that carry no regime information. A Kalman filter estimates the latent *true* state of each indicator, stripping out observation noise while tracking genuine directional changes.

The model is a random-walk state space applied separately to each (ETF, indicator) pair:

$$x_t = x_{t-1} + w_t, \quad w_t \sim \mathcal{N}(0,\, W)$$

$$y_t = x_t + v_t, \quad v_t \sim \mathcal{N}(0,\, V)$$

where $x_t$ is the unobserved latent state, $y_t$ is the observed indicator value, $W$ is the process noise variance (how much the true signal drifts), and $V$ is the observation noise variance (measurement error). A higher $W/V$ ratio makes the filter more responsive; a lower ratio gives heavier smoothing.

**Critical constraint:** $W$ and $V$ are estimated via maximum likelihood on the **training period only**. The fitted parameters are then frozen and applied forward to the test period. This enforces strict temporal separation and eliminates look-ahead bias.

### 4.2.2 Parameter Estimation & Filtering

```{r kalman-fit, warning=FALSE}
# ── Fit Kalman parameters (W, V) via MLE on training data only ───────────────
# Returns a list with: filtered_state, P_t (state variance), innovation

fit_and_filter_kalman <- function(y_obs) {
  # Remove leading NAs from indicator warm-up period
  valid_idx <- which(!is.na(y_obs))
  if (length(valid_idx) < 30L) {
    return(tibble::tibble(
      filtered_state = rep(NA_real_, length(y_obs)),
      P_t            = rep(NA_real_, length(y_obs)),
      innovation     = rep(NA_real_, length(y_obs))
    ))
  }
  y_clean <- y_obs[valid_idx]

  # Build random-walk DLM: FF=1 (observation), GG=1 (state transition),
  # V (obs noise), W (process noise) — both estimated via MLE
  build_fn <- function(pars) {
    dlm::dlm(
      FF = matrix(1),
      GG = matrix(1),
      V  = exp(pars[1]),   # log-transform keeps variance positive
      W  = exp(pars[2]),
      m0 = matrix(y_clean[1]),
      C0 = matrix(100)     # diffuse prior
    )
  }

  # MLE fit on training observations
  mle_fit <- dlm::dlmMLE(
    y     = y_clean,
    parm  = c(0, 0),       # initialise log(V), log(W) at 0
    build = build_fn,
    control = list(maxit = 500)
  )

  fitted_model <- build_fn(mle_fit$par)

  # Forward Kalman filter pass
  kf <- dlm::dlmFilter(y_clean, fitted_model)

  # Extract outputs — drop t=0 prior state to align with observations
  f_state <- as.numeric(kf$m[-1])               # filtered state x̂_t
  p_vals  <- as.numeric(dlm::dlmSvd2var(kf$U.C, kf$D.C)[-1])  # P_t
  innov   <- as.numeric(kf$y - kf$f)            # innovation v_t = y_t - ŷ_t

  # Re-insert NAs for warm-up rows so length matches original y_obs
  out <- tibble::tibble(
    filtered_state = rep(NA_real_, length(y_obs)),
    P_t            = rep(NA_real_, length(y_obs)),
    innovation     = rep(NA_real_, length(y_obs))
  )
  out$filtered_state[valid_idx] <- f_state
  out$P_t[valid_idx]            <- p_vals
  out$innovation[valid_idx]     <- innov
  out
}

# ── Apply to every (symbol, indicator) combination in training data ───────────
# Each pair gets its own independently estimated W and V
kalman_params_train <- raw_indicators_train |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::group_modify(function(.x, .y) {
    kf_out <- fit_and_filter_kalman(.x$raw_value)
    dplyr::bind_cols(.x, kf_out)
  }) |>
  dplyr::ungroup()
```

### 4.2.3 Applying Frozen Parameters to the Test Period

The MLE-estimated model parameters are derived entirely from training data. To apply them to the test period without refitting, we re-run the filter forward using the final training state as the initial condition.

```{r kalman-apply-test, warning=FALSE}
# ── Helper: apply a pre-fit DLM to new observations ──────────────────────────
apply_frozen_kalman <- function(train_y, test_y) {
  valid_train <- which(!is.na(train_y))
  if (length(valid_train) < 30L) {
    return(tibble::tibble(
      filtered_state = rep(NA_real_, length(test_y)),
      P_t            = rep(NA_real_, length(test_y)),
      innovation     = rep(NA_real_, length(test_y))
    ))
  }
  y_train_clean <- train_y[valid_train]

  build_fn <- function(pars) {
    dlm::dlm(FF = matrix(1), GG = matrix(1),
              V = exp(pars[1]), W = exp(pars[2]),
              m0 = matrix(y_train_clean[1]), C0 = matrix(100))
  }

  mle_fit     <- dlm::dlmMLE(y_train_clean, c(0, 0), build_fn,
                              control = list(maxit = 500))
  fitted_mod  <- build_fn(mle_fit$par)

  # Filter training data to get terminal state for initialisation
  kf_train    <- dlm::dlmFilter(y_train_clean, fitted_mod)
  n_tr        <- length(y_train_clean)
  m_terminal  <- as.numeric(kf_train$m[n_tr + 1])
  C_terminal  <- as.numeric(dlm::dlmSvd2var(kf_train$U.C, kf_train$D.C)[n_tr + 1])

  # Re-initialise model at terminal training state — NO refitting
  fitted_mod$m0 <- matrix(m_terminal)
  fitted_mod$C0 <- matrix(C_terminal)

  valid_test <- which(!is.na(test_y))
  out <- tibble::tibble(
    filtered_state = rep(NA_real_, length(test_y)),
    P_t            = rep(NA_real_, length(test_y)),
    innovation     = rep(NA_real_, length(test_y))
  )

  if (length(valid_test) == 0L) return(out)

  y_test_clean <- test_y[valid_test]
  kf_test      <- dlm::dlmFilter(y_test_clean, fitted_mod)

  out$filtered_state[valid_test] <- as.numeric(kf_test$m[-1])
  out$P_t[valid_test]            <- as.numeric(
    dlm::dlmSvd2var(kf_test$U.C, kf_test$D.C)[-1])
  out$innovation[valid_test]     <- as.numeric(kf_test$y - kf_test$f)
  out
}

# ── Apply frozen parameters to test data ────────────────────────────────────
kalman_params_test <- raw_indicators_test |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::group_modify(function(.x, .y) {
    # Retrieve matching training series for this (symbol, indicator) pair
    train_series <- raw_indicators_train |>
      dplyr::filter(symbol == .y$symbol, indicator == .y$indicator) |>
      dplyr::pull(raw_value)

    kf_out <- apply_frozen_kalman(train_series, .x$raw_value)
    dplyr::bind_cols(.x, kf_out)
  }) |>
  dplyr::ungroup()

# ── Combine training and test Kalman outputs into a single tidy tibble ────────
kalman_all <- dplyr::bind_rows(
  dplyr::mutate(kalman_params_train, period = "train"),
  dplyr::mutate(kalman_params_test,  period = "test")
) |>
  dplyr::select(symbol, date, indicator, period,
                raw_value, filtered_state, P_t, innovation) |>
  dplyr::arrange(symbol, indicator, date)
```

## 4.3 Delta Signal & Adaptive Threshold

### 4.3.1 Delta: First Difference of Filtered State

The filtered state $\hat{x}_t$ tells us the Kalman filter's best estimate of the indicator's current level. The **delta** is its first difference:

$$\delta_t = \hat{x}_t - \hat{x}_{t-1}$$

Delta measures the *rate of change* of the smoothed signal — analogous to velocity in a physical system. A positive $\delta_t$ means the filtered signal is accelerating upward; negative means it is decelerating or reversing. This is the entry trigger quantity: we want to enter positions when the smoothed signal has genuine directional momentum, not just a level reading.

### 4.3.2 Adaptive Threshold Scaled by State Covariance

The Kalman filter outputs the state covariance $P_t$ alongside the filtered state. $P_t$ measures the filter's own uncertainty about its state estimate — it rises when the filter is being surprised by observations (regime transition) and falls when the model fits well (stable regime).

The adaptive entry threshold uses $P_t$ directly:

$$\theta_{\text{adaptive},t} = \theta_{\text{base}} \cdot \sqrt{P_t}$$

where $\theta_{\text{base}}$ is estimated as the median absolute delta over the **training period only**. The intuition:

- **High $P_t$ (high uncertainty):** the filter is confused — require a larger $|\delta_t|$ to enter, avoiding false signals during regime transitions
- **Low $P_t$ (high certainty):** the filter is confident — allow smaller $|\delta_t|$ to trigger entries, capturing early momentum

```{r compute-delta}
# ── Compute delta and adaptive threshold from Kalman outputs ─────────────────
kalman_all <- kalman_all |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::arrange(date, .by_group = TRUE) |>
  dplyr::mutate(
    delta = filtered_state - dplyr::lag(filtered_state)
  ) |>
  dplyr::ungroup()

# ── Estimate theta_base per (symbol, indicator) on TRAINING data only ─────────
theta_base_tbl <- kalman_all |>
  dplyr::filter(period == "train", !is.na(delta)) |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::summarise(
    theta_base = stats::median(abs(delta), na.rm = TRUE),
    .groups    = "drop"
  )

# ── Join theta_base back and compute adaptive threshold ──────────────────────
kalman_all <- kalman_all |>
  dplyr::left_join(theta_base_tbl, by = c("symbol", "indicator")) |>
  dplyr::mutate(
    theta_adaptive = theta_base * sqrt(P_t)
  )
```

## 4.4 Innovation Sequence & Rolling Z-Score

The Kalman innovation at time $t$ is defined as the one-step-ahead prediction error:

$$v_t = y_t - \hat{y}_{t|t-1}$$

Under a correctly specified model, innovations should be approximately white noise with zero mean. When the innovation z-score spikes — i.e., the filter is being repeatedly surprised — this is statistical evidence of a regime break. This forms the primary exit signal (Section 6).

```{r compute-innovation-zscore}
# ── Rolling z-score of innovation sequence (60-day window) ───────────────────
kalman_all <- kalman_all |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::arrange(date, .by_group = TRUE) |>
  dplyr::mutate(
    innov_roll_mean = zoo::rollapply(
      innovation, width = 60, FUN = mean,
      fill = NA, align = "right", na.rm = TRUE
    ),
    innov_roll_sd = zoo::rollapply(
      innovation, width = 60, FUN = sd,
      fill = NA, align = "right", na.rm = TRUE
    ),
    innovation_zscore = dplyr::if_else(
      innov_roll_sd > 0,
      (innovation - innov_roll_mean) / innov_roll_sd,
      NA_real_
    )
  ) |>
  dplyr::ungroup()
```

## 4.5 Visualisation: XLK — Raw vs Filtered Signals

XLK (Technology) is used as the illustrative ETF throughout this section. Its high volatility and distinct trend cycles make it the most visually informative sector for demonstrating Kalman smoothing.

The 2×2 grid below shows all four indicators over the training period. In each panel the grey line is the raw TTR indicator; the coloured line is the Kalman filtered state $\hat{x}_t$.

```{r chart-raw-vs-filtered, fig.height=10, fig.alt="2x2 grid of raw vs Kalman filtered signals for XLK across all four indicators"}
# ── Prepare XLK training data for plotting ───────────────────────────────────
xlk_plot_data <- kalman_all |>
  dplyr::filter(symbol == "XLK", period == "train",
                !is.na(filtered_state)) |>
  dplyr::select(date, indicator, raw_value, filtered_state)

indicator_labels <- c(
  RSI        = "RSI (14-day)",
  MACD       = "MACD Line − Signal",
  ROC        = "Rate of Change (10-day)",
  BollingerB = "Bollinger Band %B"
)

# ── Build one panel per indicator ────────────────────────────────────────────
make_panel <- function(ind_name) {
  df <- xlk_plot_data |> dplyr::filter(indicator == ind_name)

  ggplot2::ggplot(df, ggplot2::aes(x = date)) +
    ggplot2::geom_line(
      ggplot2::aes(y = raw_value),
      colour = "grey70", linewidth = 0.35, alpha = 0.8
    ) +
    ggplot2::geom_line(
      ggplot2::aes(y = filtered_state),
      colour = "#2563EB", linewidth = 0.65
    ) +
    ggplot2::scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
    ggplot2::labs(
      title    = indicator_labels[ind_name],
      subtitle = "Grey = raw  |  Blue = Kalman filtered state",
      x        = NULL,
      y        = ind_name
    ) +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      plot.title       = ggplot2::element_text(face = "bold", size = 10)
    )
}

p_rsi  <- make_panel("RSI")
p_macd <- make_panel("MACD")
p_roc  <- make_panel("ROC")
p_bb   <- make_panel("BollingerB")

# ── 2x2 patchwork layout ─────────────────────────────────────────────────────
(p_rsi | p_macd) / (p_roc | p_bb) +
  patchwork::plot_annotation(
    title   = "XLK — Raw vs Kalman Filtered Technical Indicators (Training Period)",
    caption = "Training period: June 2000 – December 2024",
    theme   = ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = 12)
    )
  )
```

```{r chart-delta-threshold, fig.alt="XLK RSI delta signal with adaptive threshold bands"}
# ── Delta + adaptive threshold chart for XLK RSI ─────────────────────────────
xlk_delta_data <- kalman_all |>
  dplyr::filter(symbol == "XLK", indicator == "RSI",
                period == "train", !is.na(delta))

ggplot2::ggplot(xlk_delta_data, ggplot2::aes(x = date)) +
  ggplot2::geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.4) +
  ggplot2::geom_line(
    ggplot2::aes(y = delta),
    colour = "#1D4ED8", linewidth = 0.45, alpha = 0.85
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y =  theta_adaptive),
    colour = "#DC2626", linewidth = 0.5, linetype = "dashed"
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = -theta_adaptive),
    colour = "#DC2626", linewidth = 0.5, linetype = "dashed"
  ) +
  ggplot2::scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  ggplot2::labs(
    title    = "XLK RSI — Delta Signal with Adaptive Entry Threshold",
    subtitle = "Blue = δ_t  |  Red dashed = ±θ_adaptive = θ_base · √P_t",
    x        = NULL,
    y        = "Delta (filtered state first difference)",
    caption  = "Entry triggered when |δ_t| exceeds the adaptive threshold"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    plot.title       = ggplot2::element_text(face = "bold")
  )
```

```{r chart-innovation-zscore, fig.alt="XLK RSI Kalman innovation z-score with exit threshold lines"}
# ── Innovation z-score chart for XLK RSI ────────────────────────────────────
xlk_innov_data <- kalman_all |>
  dplyr::filter(symbol == "XLK", indicator == "RSI",
                period == "train", !is.na(innovation_zscore))

ggplot2::ggplot(xlk_innov_data, ggplot2::aes(x = date)) +
  ggplot2::geom_hline(yintercept =  0, colour = "grey50",  linewidth = 0.4) +
  ggplot2::geom_hline(yintercept =  2, colour = "#DC2626", linewidth = 0.55,
                      linetype = "dashed") +
  ggplot2::geom_hline(yintercept = -2, colour = "#DC2626", linewidth = 0.55,
                      linetype = "dashed") +
  ggplot2::geom_line(
    ggplot2::aes(y = innovation_zscore),
    colour = "#059669", linewidth = 0.45, alpha = 0.85
  ) +
  ggplot2::annotate("text", x = min(xlk_innov_data$date) + lubridate::days(200),
                    y = 2.3, label = "Exit threshold (+2)", colour = "#DC2626",
                    size = 3, hjust = 0) +
  ggplot2::annotate("text", x = min(xlk_innov_data$date) + lubridate::days(200),
                    y = -2.3, label = "Exit threshold (−2)", colour = "#DC2626",
                    size = 3, hjust = 0) +
  ggplot2::scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  ggplot2::labs(
    title    = "XLK RSI — Kalman Innovation Z-Score (60-Day Rolling)",
    subtitle = "Spikes beyond ±2 signal filter surprise — potential regime break → exit trigger",
    x        = NULL,
    y        = "Innovation Z-Score",
    caption  = "Innovation v_t = observed_t − predicted_t  |  Z-score computed on 60-day rolling window"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    plot.title       = ggplot2::element_text(face = "bold")
  )
```

---

# 5. Clustering {#sec-clustering}

<!-- ─────────────────────────────────────────────────────────────────────────
     SECTION 5: CLUSTERING
     Goal: Identify market regimes via K-means; validate and label clusters.
     Content to write:
       A. Elbow Plot
          • Within-cluster sum of squares (WCSS) for k = 2..10
          • Identify optimal k visually and justify choice

       B. K-Means via tidymodels
          • recipe() → step_scale() → step_center() → prep() → bake()
          • kmeans() on baked training features
          • Monthly cluster assignment: cluster at each end-of-month date
          • Use lubridate::rollback() for end-of-month dates
          • Never use future data in cluster assignment (no look-ahead)

       C. Silhouette Analysis
          • Compute silhouette width for each observation
          • Average silhouette width as cluster quality metric
          • Plot silhouette widths, flag low-quality assignments

       D. Cluster Labelling
          • Interpret clusters as Growth / Neutral / Defensive
          • Show cluster centroid values per feature
          • Map numeric cluster IDs to regime labels consistently

       E. Visualisation
          • Static ggplot: PCA-reduced cluster scatter (training period)
          • gganimate: animated cluster membership month-by-month

       F. Stability Analysis
          • Re-cluster on pre-COVID (to 2019-12-31) and post-COVID
            (2020-01-01 onward) sub-periods separately
          • Compare centroids and label assignments
          • Discuss regime stability implications for out-of-sample use
     ──────────────────────────────────────────────────────────────────────── -->

<!-- TODO: Elbow plot for k = 2..10 -->

<!-- TODO: tidymodels K-means pipeline -->

<!-- TODO: Monthly cluster assignment at rollback() dates -->

<!-- TODO: Silhouette analysis and plot -->

<!-- TODO: Cluster centroid table and label mapping -->

<!-- TODO: Static ggplot PCA scatter -->

<!-- TODO: gganimate animated cluster membership chart -->

<!-- TODO: Stability analysis — pre vs post COVID sub-periods -->

---

# 6. Signal Generation {#sec-signals}

<!-- ─────────────────────────────────────────────────────────────────────────
     SECTION 6: SIGNAL GENERATION
     Goal: Combine cluster regime label with Kalman delta to produce
           actionable long/short entry signals, and define exit triggers.
     Content to write:
       A. Entry Logic
          • Condition: cluster != "Neutral" AND |delta_t| > theta_adaptive
          • Growth cluster  → LONG  (delta must be positive)
          • Defensive cluster → SHORT (delta must be negative)
          • Neutral cluster  → FLAT  (no position taken)

       B. Exit Logic — Primary (Innovation Z-Score)
          • Compute rolling z-score of Kalman innovation sequence
          • Exit when |z_score(innovation_t)| > 2
          • Applies symmetrically to long and short positions
          • Rationale: filter is "surprised" — regime may have broken

       C. Exit Logic — Secondary (Cluster Flip)
          • Exit long  when cluster flips from Growth to Neutral/Defensive
          • Exit short when cluster flips from Defensive to Neutral/Growth
          • Evaluated only at monthly rebalancing dates
          • Provides orthogonal regime-level risk management

       D. Visualisation
          • Time series chart: filtered signal + delta + entry/exit markers
          • Show both a long-entry example and a short-entry example
          • Annotate innovation z-score spikes that triggered exits
     ──────────────────────────────────────────────────────────────────────── -->

<!-- TODO: Implement entry signal logic (delta + cluster) -->

<!-- TODO: Implement primary exit: innovation z-score > 2 -->

<!-- TODO: Implement secondary exit: cluster flip at rebalance date -->

<!-- TODO: Chart: filtered signal + delta + entry/exit markers (long example) -->

<!-- TODO: Chart: filtered signal + delta + entry/exit markers (short example) -->

---

# 7. Decision Rules (Formal) {#sec-rules}

<!-- ─────────────────────────────────────────────────────────────────────────
     SECTION 7: DECISION RULES — FORMAL STATEMENT
     Goal: State every decision rule mathematically, unambiguously.
           Professor grades explicitly on clarity here.
     Content to write:
       • Numbered list of all rules using LaTeX math blocks
       1. Kalman state-space model equations
       2. Delta definition: δ_t = x̂_t − x̂_{t−1}
       3. Adaptive threshold: θ_t = θ_base · √P_t
       4. Entry rule — Long:
            c_t = "Growth" AND δ_t > +θ_t  →  w_i > 0
       5. Entry rule — Short:
            c_t = "Defensive" AND δ_t < −θ_t  →  w_i < 0
       6. No-trade rule — Neutral:
            c_t = "Neutral"  →  w_i = 0
       7. Primary exit — Innovation spike:
            |z(v_t)| > 2  →  close position immediately
       8. Secondary exit — Cluster flip:
            c_t ≠ c_{t−1} at rebalance date  →  close position
       9. Inverse volatility weighting formula:
            w_i = (1/σ_i) / Σ_j (1/σ_j)   [long book, positive]
            w_i = −(1/σ_i) / Σ_j (1/σ_j)  [short book, negative]
      10. Monthly rebalancing timing:
            t* = lubridate::rollback(first day of next month)
     ──────────────────────────────────────────────────────────────────────── -->

<!-- TODO: Write all formal decision rules with LaTeX math notation -->

---

# 8. Training Period Backtest {#sec-training}

<!-- ─────────────────────────────────────────────────────────────────────────
     SECTION 8: TRAINING PERIOD BACKTEST — June 2000 to December 2024
     Goal: Evaluate strategy performance over in-sample period vs SPY.
     Content to write:
       A. Portfolio Construction
          • Implement rebalance_portfolio() function
          • Inverse volatility weighting within active cluster
          • Negative weights for Defensive (short) book
          • Flat (cash) during Neutral regime
          • Monthly rebalancing at rollback() end-of-month dates
          • Track: portfolio NAV, per-ETF allocations, monthly turnover

       B. Performance Metrics
          • Cumulative return vs SPY (both on same chart)
          • Annualised return, annualised volatility
          • Sharpe ratio (annualised, risk-free = 0 or 3-month T-bill)
          • Maximum drawdown and drawdown duration
          • Information ratio: active return / tracking error vs SPY
          • Win rate of individual monthly positions
          • Long book contribution vs Short book contribution

       C. Stress Period Analysis
          • 2008 Global Financial Crisis
          • 2020 COVID crash and recovery
          • 2022 rate-hike bear market
          • Annotate cumulative return chart with these periods

       D. Visualisation
          • plotly: cumulative return — strategy vs SPY (interactive)
          • ggplot: monthly returns heatmap (year × month grid)
          • ggplot: drawdown chart
          • ggplot: long/short contribution stacked bar (monthly)
          • patchwork: combined layout of above charts
     ──────────────────────────────────────────────────────────────────────── -->

<!-- TODO: Implement rebalance_portfolio() with inverse vol weighting -->

<!-- TODO: Run backtest loop over training period -->

<!-- TODO: Compute performance metrics table (gt()) -->

<!-- TODO: plotly cumulative return chart vs SPY -->

<!-- TODO: Monthly returns heatmap -->

<!-- TODO: Drawdown chart -->

<!-- TODO: Long/short contribution breakdown -->

<!-- TODO: Stress period annotations -->

<!-- TODO: patchwork combined layout -->

---

# 9. Out-of-Sample Test {#sec-oos}

<!-- ─────────────────────────────────────────────────────────────────────────
     SECTION 9: OUT-OF-SAMPLE TEST — January 2025 to March 2026
     Goal: Honest assessment of strategy generalisation beyond training data.
     NOTE: Kalman W/V parameters and theta_base estimated on training only.
           Scaling parameters (mean, sd) from training applied to test.
           Cluster model fit on training, applied forward without refitting.
     Content to write:
       A. OOS Execution
          • Apply pre-fit Kalman filter parameters to test data
          • Apply pre-fit scaling parameters to test features
          • Apply pre-fit cluster centroids to assign test month labels
          • Generate signals and positions identically to training logic

       B. Performance Metrics (same set as Section 8)
          • Cumulative return vs SPY
          • Sharpe ratio, max drawdown, information ratio
          • Win rate

       C. Long-Only vs Long-Short Comparison
          • Run long-only version (no short book, flat when Defensive)
          • Compare: long-short vs long-only vs SPY
          • Quantify value added by shorting the Defensive cluster

       D. Honest Assessment
          • If OOS underperforms, explain using cluster stability analysis
          • Did regime labels shift in 2025? Compare centroids
          • When does the strategy fail: VIX > 40, silhouette < 0.3?
          • Regime detection lag vs rapid market moves

       E. Visualisation
          • plotly: OOS cumulative return — strategy vs SPY (interactive)
          • ggplot: OOS monthly returns heatmap
          • ggplot: long-short vs long-only vs SPY comparison chart
     ──────────────────────────────────────────────────────────────────────── -->

<!-- TODO: Apply pre-fit model to test data (no refitting) -->

<!-- TODO: Generate OOS positions and NAV -->

<!-- TODO: Compute OOS performance metrics table (gt()) -->

<!-- TODO: Long-only vs long-short comparison -->

<!-- TODO: plotly OOS cumulative return chart vs SPY -->

<!-- TODO: OOS monthly returns heatmap -->

<!-- TODO: Long-short vs long-only vs SPY comparison chart -->

<!-- TODO: Honest narrative assessment of OOS performance -->

---

# 10. Circle Back {#sec-circle-back}

<!-- ─────────────────────────────────────────────────────────────────────────
     SECTION 10: CIRCLE BACK
     Goal: Return to the original question — does this strategy justify
           paying active management fees over holding SPY? Provide a
           balanced, intellectually honest answer.
     Content to write:
       A. What the Strategy Learned
          • Which regime signals were most predictive?
          • Which ETFs drove long-book returns? Short-book returns?
          • Did the Kalman innovation exit genuinely add value vs
            holding through the regime change?

       B. When Clustering Fails
          • VIX > 40: panic regimes don't cluster cleanly
          • Silhouette width < 0.3: regime ambiguity, signals unreliable
          • Rapid regime transitions (days, not months): monthly
            rebalancing introduces dangerous lag
          • Extended Neutral periods: strategy flat, opportunity cost

       C. Limitations
          • No transaction costs or slippage modelled
          • Short-selling constraints ignored (borrow cost, availability)
          • Margin call constraints not modelled (per professor confirmation)
          • Single-model risk: K-means is sensitive to initialisation
          • Look-ahead audit: state any remaining risks of data leakage

       D. Inversion: What Would Kill This Strategy?
          • Persistent low-volatility regime (all clusters merge)
          • Structural break in sector correlations (e.g. AI disruption)
          • Central bank intervention compressing cross-sector dispersion
          • ETF proliferation reducing signal in individual sector ETFs

       E. Active Management Value — Final Answer
          • Does the Information Ratio justify active fees?
          • Under what market conditions does active add value here?
          • Honest recommendation: when to use this vs hold SPY
     ──────────────────────────────────────────────────────────────────────── -->

<!-- TODO: Write "What the strategy learned" narrative -->

<!-- TODO: Write "When clustering fails" with quantitative thresholds -->

<!-- TODO: Write limitations section -->

<!-- TODO: Write inversion / failure mode analysis -->

<!-- TODO: Write final active management value assessment -->

---

# References {#sec-references}

<!-- ─────────────────────────────────────────────────────────────────────────
     REFERENCES
     Required citations:
       • Kalman (1960) — original Kalman filter paper
       • Harvey (1989) — Time Series Models, state space in econometrics
       • Durbin & Koopman (2012) — State Space Methods by State Space
       • Hartigan & Wong (1979) — K-means algorithm
       • Avellaneda & Lee (2010) — stat arb, state space in finance
     ──────────────────────────────────────────────────────────────────────── -->

<!-- TODO: Create references.bib with all required citations -->
<!-- TODO: Re-add  bibliography: references.bib  to YAML header once .bib file exists -->
<!-- TODO: Restore ::: {#refs} ::: div once bibliography is wired up -->
