---
title: "Clustering-Based Sector Rotation Strategy"
subtitle: "Active Fund Management via K-Means Regime Detection and Kalman Signal Filtering"
author: "Cai Naidoo"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    theme: flatly
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 6
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
# ── Package loading ──────────────────────────────────────────────────────────
# Packages used via explicit namespace notation throughout (pkg::fn()).
# require() is used here so a single missing package does not abort the render.

pkgs <- c(
  "tidyverse", "tidyquant", "tidymodels",
  "ggplot2", "plotly", "patchwork",
  "cluster", "factoextra", "moments",
  "dlm", "KFAS",
  "gt", "scales", "lubridate", "rlang", "purrr", "zoo", "TTR",
  "stringr"
)

# ── Named constants for signal parameters (avoid magic numbers) ──────────────
LOOKBACK_DAYS    <- 21L   # rolling window for features and vol
INNOV_WINDOW     <- 63L   # rolling window for innovation z-score (1 quarter)
INNOV_Z_THRESH   <- 2     # innovation z-score exit threshold
K_CLUSTERS       <- 3L    # number of k-means clusters
KMEANS_NSTART    <- 50L   # k-means random restarts
KMEANS_SEED      <- 451L  # reproducible seed

invisible(lapply(pkgs, function(p) {
  if (!requireNamespace(p, quietly = TRUE)) {
    message("Package not installed: ", p,
            " — install with install.packages('", p, "')")
  } else {
    suppressPackageStartupMessages(require(p, character.only = TRUE))
  }
}))
```

::: {.callout-tip icon=false}
## Executive Summary

**Strategy:** Clustering-based sector rotation across 11 SPDR sector ETFs, using K-means regime detection (k=3) and Kalman-filtered signal generation with adaptive entry thresholds.

| Component | Detail |
|-----------|--------|
| **Universe** | 11 sector ETFs (XLB, XLC, XLE, XLF, XLI, XLK, XLP, XLRE, XLV, XLU, XLY) |
| **Benchmark** | SPY (not traded) |
| **Regimes** | Growth (long) · Neutral (cash) · Defensive (short) |
| **Entry signal** | Kalman delta > adaptive threshold ($\theta_{\text{base}} \cdot \sqrt{P_t}$) |
| **Exit signals** | Innovation z-score \|z\| > 2 (daily) · Cluster flip (monthly) |
| **Weighting** | Inverse volatility within each book |
| **Rebalancing** | Monthly (end of month) |
| **Training** | June 2000 – December 2024 |
| **Testing** | January 2025 – March 2026 |

The core thesis: markets have regimes, regimes are detectable, and a disciplined filter can time entry and exit within them. The Kalman filter's own uncertainty ($P_t$) and surprise (innovation z-score) drive both the confidence to enter and the urgency to exit.
:::

---

# 1. Problem Framing {#sec-framing}

## The Question

SPY returned roughly 10.5% annualised over the past century — through crashes, pandemics, and bear markets. Most active managers failed to match it. This document addresses a single question: can a systematic, rules-based sector rotation strategy justify its cost over simply holding SPY?

The answer is not "yes by construction." It is "yes under specific, identifiable conditions, and no when those conditions break down."

## Markets Have Regimes

The central intuition is ecological. Markets are not machines with fixed input-output relationships — they are ecosystems with regimes. During expansions, cyclical sectors attract capital. During contractions, defensive sectors hold value. During crises, correlations collapse toward one and the distinction between sectors disappears. These are qualitatively different states, each with its own logic. A strategy that cannot tell them apart will be right on average and catastrophically wrong in the tails.

## The Investment Mandate

The strategy operates in three states. In a **Growth regime**, it goes long sector ETFs weighted by inverse volatility. In a **Defensive regime**, it goes short those same defensive names with negative inverse-volatility weights — capturing underperformance, not just avoiding it. In a **Neutral regime**, it holds cash — a deliberate acknowledgement of model uncertainty.

Rebalancing occurs at month-end. Intra-month, only the Kalman filter's innovation z-score can trigger an exit. Every position is rule-determined — no discretion, no overrides.

## Two Tools, Two Frequencies

K-means clustering identifies the **regime** at monthly frequency — which direction to trade. The Kalman filter operates at **daily frequency** — whether the evidence is strong enough to act. The filter's first difference (delta) provides a momentum signal, and its state covariance $P_t$ scales the entry threshold adaptively: high uncertainty demands stronger evidence, high confidence permits faster entries.

The regime tells you what to do. The filter tells you when.

The benchmark throughout is SPY — never traded, never assigned a cluster label, never gamed.

---

# 2. Mental Model {#sec-mental-model}

Before any code, here is the complete decision pipeline — every rule, every branch, every exit condition.

```{mermaid}
%%| fig-width: 9
flowchart TD
    A([Daily Price Data\nSPY + 11 Sector ETFs]) --> B[Feature Engineering\nRSI · MACD · ROC · Bollinger %B\nRolling Return · Vol · Sharpe]

    B --> C[Kalman Filter\nper ETF × per Indicator\nEstimate W and V on training only]

    C --> D[Filtered State x̂_t\nState Covariance P_t\nInnovation v_t]

    D --> E[Delta Signal\nδ_t = x̂_t − x̂_t-1]
    D --> F[Innovation Z-Score\nz_t = rolling z-score of v_t]

    E --> G[Monthly K-Means Clustering\nk = 3  ·  tidymodels workflow\nFeatures scaled on training only]

    G --> H{Regime Label\nat rebal date τ}

    H -->|Growth| I[Entry Check\nδ_t > +θ_adaptive ?]
    H -->|Neutral| J[FLAT\nHold Cash]
    H -->|Defensive| K[Entry Check\nδ_t < −θ_adaptive ?]

    I -->|Yes| L[LONG\nInverse Vol Weighted\n+w_i]
    I -->|No| J
    K -->|Yes| M[SHORT\nInverse Vol Weighted\n−w_i]
    K -->|No| J

    L --> N{Exit Check\nDaily}
    M --> N

    F --> N
    N -->|z_t > 2 | O[EXIT\nInnovation Spike\nRegime Break Signal]
    N -->|Cluster flips\nat month end| O
    N -->|No exit| P[Hold Position\nuntil next rebal]

    O --> Q([Portfolio NAV\nMonthly Rebalance\nlubridate::rollback])
    P --> Q
    J --> Q

    Q --> R[Performance vs SPY\nSharpe · MaxDD · Info Ratio]

    style A fill:#e0f2fe,stroke:#0284c7
    style J fill:#fef9c3,stroke:#ca8a04
    style L fill:#dcfce7,stroke:#16a34a
    style M fill:#fee2e2,stroke:#dc2626
    style O fill:#fed7aa,stroke:#ea580c
    style R fill:#f3e8ff,stroke:#9333ea
    style Q fill:#e0f2fe,stroke:#0284c7
```

::: {.callout-note icon=false}
## Poker Analogy
Clustering is reading the table before you bet. Growth is a loose aggressive game — press your edge. Defensive is tight passive — fold and preserve chips. Neutral means you are still reading the room and stay out of big pots.
:::

Data enters at the top. The cluster label sets direction (Growth = long, Defensive = short, Neutral = cash), the delta must clear the adaptive threshold to enter, and two independent exits can close any position: an innovation z-score spike (daily) or a cluster flip (monthly). Long weights sum to +1, short weights sum to −1.

With the full pipeline visible, the next step is the data that feeds it.

---

# 3. Data {#sec-data}

## 3.1 Asset Universe & Periods

These twelve tickers span the entire US equity market through a sector lens — every dollar in the S&P 500 maps to one of these eleven sectors. Together they make regime rotation visible. All data is sourced via API (`tidyquant::tq_get()`). **SPY is a benchmark only and is never traded.**

| Role | Tickers |
|------|---------|
| Benchmark (not traded) | SPY |
| Tradeable universe | XLB, XLC, XLE, XLF, XLI, XLK, XLP, XLRE, XLV, XLU, XLY |

> **Note:** XLC launched June 2018. Prior observations are absent, not imputed.

```{r data-constants}
#| echo: false
# ── Asset universe ───────────────────────────────────────────────────────────
BENCHMARK  <- "SPY"
SECTORS    <- c("XLB", "XLC", "XLE", "XLF", "XLI",
                "XLK", "XLP", "XLRE", "XLV", "XLU", "XLY")
ALL_TICKERS <- c(BENCHMARK, SECTORS)

# ── Period boundaries ────────────────────────────────────────────────────────
TRAIN_START <- lubridate::ymd("2000-06-01")
TRAIN_END   <- lubridate::ymd("2024-12-31")
TEST_START  <- lubridate::ymd("2025-01-01")
TEST_END    <- lubridate::ymd("2026-03-31")

# ── Split boundary label (for chart annotation) ──────────────────────────────
SPLIT_DATE  <- TEST_START
```

## 3.2 Data Pull

```{r data-pull}
#| echo: false
# ── Pull daily adjusted prices for all 12 tickers ───────────────────────────
raw_prices <- tidyquant::tq_get(
  ALL_TICKERS,
  get  = "stock.prices",
  from = TRAIN_START,
  to   = TEST_END
) |>
  dplyr::select(symbol, date, adjusted) |>
  dplyr::rename(price = adjusted) |>
  dplyr::arrange(symbol, date)

# ── Sanity check: confirm retrieval ─────────────────────────────────────────
raw_prices |>
  dplyr::group_by(symbol) |>
  dplyr::summarise(
    first_date = min(date),
    last_date  = max(date),
    n_obs      = dplyr::n(),
    .groups    = "drop"
  ) |>
  dplyr::arrange(symbol) |>
  gt::gt() |>
  gt::tab_header(title = "Data Coverage by Ticker") |>
  gt::fmt_date(columns = c(first_date, last_date), date_style = "yMd") |>
  gt::fmt_integer(columns = n_obs) |>
  gt::cols_label(symbol = "Ticker", first_date = "First Date",
                 last_date = "Last Date", n_obs = "Observations") |>
  gt::opt_table_font(font = gt::google_font("IBM Plex Sans")) |>
  gt::opt_stylize(style = 1, color = "blue")
```

```{r data-split}
#| echo: false
train_data <- raw_prices |>
  dplyr::filter(date >= TRAIN_START, date <= TRAIN_END)
test_data <- raw_prices |>
  dplyr::filter(date >= TEST_START, date <= TEST_END)
```

## 3.3 Indexed Price Chart

Prices indexed to 100 at each ticker's first available date. The dashed line separates the training period (June 2000 – December 2024) from the test period (January 2025 – March 2026).

```{r data-index}
#| echo: false
# ── Index each ticker to 100 at its own first available date ─────────────────
prices_indexed <- raw_prices |>
  dplyr::group_by(symbol) |>
  dplyr::mutate(
    base_price    = dplyr::first(price),
    indexed_price = price / base_price * 100
  ) |>
  dplyr::ungroup() |>
  dplyr::select(symbol, date, indexed_price)
```

```{r chart-indexed, fig.alt="Interactive chart of all 12 ETFs indexed to 100 at first available date"}
# ── Base ggplot ───────────────────────────────────────────────────────────────
p_indexed <- ggplot2::ggplot(
  prices_indexed,
  ggplot2::aes(x = date, y = indexed_price, colour = symbol, group = symbol)
) +
  ggplot2::geom_line(linewidth = 0.45, alpha = 0.85) +
  ggplot2::geom_vline(
    xintercept = as.numeric(SPLIT_DATE),
    linetype   = "dashed",
    colour     = "grey30",
    linewidth  = 0.6
  ) +
  ggplot2::annotate(
    "text",
    x      = SPLIT_DATE + lubridate::days(60),
    y      = max(prices_indexed$indexed_price, na.rm = TRUE) * 0.97,
    label  = "← Train  |  Test →",
    size   = 3,
    hjust  = 0,
    colour = "grey30"
  ) +
  ggplot2::scale_colour_manual(
    values = scales::hue_pal()(length(ALL_TICKERS))
  ) +
  ggplot2::scale_x_date(
    date_breaks = "2 years",
    date_labels = "%Y"
  ) +
  ggplot2::scale_y_continuous(
    labels = scales::comma_format(suffix = "")
  ) +
  ggplot2::labs(
    title    = "SPDR Sector ETFs — Indexed to 100 at First Available Date",
    subtitle = "Daily adjusted closing prices  |  Dashed line = Train/Test split (Jan 2025)",
    x        = NULL,
    y        = "Indexed Price (base = 100)",
    colour   = "Ticker",
    caption  = "Source: Yahoo Finance via tidyquant::tq_get()"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    legend.position  = "right",
    panel.grid.minor = ggplot2::element_blank(),
    plot.title       = ggplot2::element_text(face = "bold")
  )

# ── Wrap in plotly for interactivity ─────────────────────────────────────────
plotly::ggplotly(p_indexed, tooltip = c("colour", "x", "y")) |>
  plotly::layout(
    hovermode = "x unified",
    legend    = list(title = list(text = "<b>Ticker</b>"))
  )
```

## 3.4 Summary Statistics

```{r summary-stats}
#| echo: false
# ── Compute per-ETF summary statistics on the full raw series ────────────────
etf_summary <- raw_prices |>
  dplyr::group_by(symbol) |>
  dplyr::mutate(
    daily_return = (price / dplyr::lag(price)) - 1
  ) |>
  dplyr::summarise(
    `First Date`      = min(date),
    `Last Date`       = max(date),
    `Observations`    = dplyr::n(),
    `Mean Dly Ret`    = mean(daily_return, na.rm = TRUE),
    `Ann. Vol`        = sd(daily_return, na.rm = TRUE) * sqrt(252),
    `Min Price`       = min(price, na.rm = TRUE),
    `Max Price`       = max(price, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    `Role` = dplyr::if_else(symbol == "SPY", "Benchmark", "Tradeable")
  ) |>
  dplyr::arrange(dplyr::desc(`Role`), symbol)

# ── Render as formatted gt table ─────────────────────────────────────────────
etf_summary |>
  gt::gt() |>
  gt::tab_header(
    title    = "Asset Universe — Summary Statistics",
    subtitle = "Full sample: June 2000 – March 2026 | Adjusted closing prices"
  ) |>
  gt::fmt_date(columns = c(`First Date`, `Last Date`), date_style = "yMd") |>
  gt::fmt_number(columns = `Observations`, use_seps = TRUE, decimals = 0) |>
  gt::fmt_percent(columns = c(`Mean Dly Ret`, `Ann. Vol`), decimals = 2) |>
  gt::fmt_currency(columns = c(`Min Price`, `Max Price`), currency = "USD",
                   decimals = 2) |>
  gt::tab_style(
    style     = gt::cell_fill(color = "#f0f4ff"),
    locations = gt::cells_body(rows = `Role` == "Benchmark")
  ) |>
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::cols_label(symbol = "Ticker") |>
  gt::tab_footnote(
    footnote  = "XLC launched June 2018; prior observations are absent, not imputed.",
    locations = gt::cells_body(columns = `First Date`,
                               rows    = symbol == "XLC")
  ) |>
  gt::tab_footnote(
    footnote  = "SPY is a benchmark only and is excluded from all portfolio positions.",
    locations = gt::cells_body(columns = `Role`,
                               rows    = symbol == "SPY")
  ) |>
  gt::opt_table_font(font = gt::google_font("IBM Plex Sans")) |>
  gt::opt_stylize(style = 1, color = "blue")
```

---

# 4. Feature Engineering {#sec-features}

## 4.1 Technical Indicators

Raw technical indicators are the problem this section solves — they react to noise as much as signal. Four indicators are computed daily for each sector ETF, then fed through a Kalman filter to extract what matters.

| Indicator | Window | Purpose |
|-----------|--------|---------|
| **RSI** | 14-day | Momentum oscillator |
| **MACD** | 12/26/9 | Trend acceleration |
| **ROC** | 10-day | Near-term price momentum |
| **Bollinger %B** | 20-day, 2 SD | Breakout detection |

```{r compute-indicators}
#| echo: false
# ── Compute all four indicators for every tradeable ETF ──────────────────────
# Input: train_data and test_data from Section 3 (adjusted daily prices)
# Output: raw_indicators — long-format tibble, one row per ETF × date × indicator

compute_indicators <- function(df) {
  prices <- df$price

  rsi_vals  <- TTR::RSI(prices, n = 14)

  macd_obj  <- TTR::MACD(prices, nFast = 12, nSlow = 26, nSig = 9)
  macd_vals <- macd_obj[, "macd"] - macd_obj[, "signal"]

  roc_vals  <- TTR::ROC(prices, n = 10, type = "discrete")

  bb_obj    <- TTR::BBands(prices, n = 20, sd = 2)
  pctb_vals <- bb_obj[, "pctB"]

  tibble::tibble(
    date      = df$date,
    RSI       = as.numeric(rsi_vals),
    MACD      = as.numeric(macd_vals),
    ROC       = as.numeric(roc_vals),
    BollingerB = as.numeric(pctb_vals)
  )
}

# Apply to training data for each sector ETF (not SPY)
raw_indicators_train <- train_data |>
  dplyr::filter(symbol %in% SECTORS) |>
  dplyr::group_by(symbol) |>
  dplyr::group_modify(~ compute_indicators(.x)) |>
  dplyr::ungroup() |>
  tidyr::pivot_longer(
    cols      = c(RSI, MACD, ROC, BollingerB),
    names_to  = "indicator",
    values_to = "raw_value"
  ) |>
  dplyr::filter(!is.na(raw_value))

# Apply to test data
raw_indicators_test <- test_data |>
  dplyr::filter(symbol %in% SECTORS) |>
  dplyr::group_by(symbol) |>
  dplyr::group_modify(~ compute_indicators(.x)) |>
  dplyr::ungroup() |>
  tidyr::pivot_longer(
    cols      = c(RSI, MACD, ROC, BollingerB),
    names_to  = "indicator",
    values_to = "raw_value"
  ) |>
  dplyr::filter(!is.na(raw_value))
```

## 4.2 Kalman Filter

### 4.2.1 State-Space Model

A random-walk state space model separates latent signal from observation noise, applied independently to each (ETF, indicator) pair:

$$x_t = x_{t-1} + w_t, \quad w_t \sim \mathcal{N}(0,\, W)$$

$$y_t = x_t + v_t, \quad v_t \sim \mathcal{N}(0,\, V)$$

$W$ controls how much the true signal drifts; $V$ captures measurement noise. A higher $W/V$ ratio makes the filter more responsive. **$W$ and $V$ are estimated via MLE on training data only and frozen for the test period** — no look-ahead.

```{r kalman-fit, warning=FALSE}
#| echo: false
# ── Fit Kalman parameters (W, V) via MLE on training data only ───────────────
# Returns a list with: filtered_state, P_t (state variance), innovation

fit_and_filter_kalman <- function(y_obs) {
  # Remove leading NAs from indicator warm-up period
  valid_idx <- which(!is.na(y_obs))
  if (length(valid_idx) < 30L) {
    return(tibble::tibble(
      filtered_state = rep(NA_real_, length(y_obs)),
      P_t            = rep(NA_real_, length(y_obs)),
      innovation     = rep(NA_real_, length(y_obs))
    ))
  }
  y_clean <- y_obs[valid_idx]

  # Build random-walk DLM: FF=1 (observation), GG=1 (state transition),
  # V (obs noise), W (process noise) — both estimated via MLE
  build_fn <- function(pars) {
    dlm::dlm(
      FF = matrix(1),
      GG = matrix(1),
      V  = exp(pars[1]),   # log-transform keeps variance positive
      W  = exp(pars[2]),
      m0 = matrix(y_clean[1]),
      C0 = matrix(100)     # diffuse prior
    )
  }

  # MLE fit on training observations
  mle_fit <- dlm::dlmMLE(
    y     = y_clean,
    parm  = c(0, 0),       # initialise log(V), log(W) at 0
    build = build_fn,
    control = list(maxit = 500)
  )

  fitted_model <- build_fn(mle_fit$par)

  # Forward Kalman filter pass
  kf <- dlm::dlmFilter(y_clean, fitted_model)

  # Extract outputs — drop t=0 prior state to align with observations
  f_state <- as.numeric(kf$m[-1])               # filtered state x̂_t
  p_vals  <- as.numeric(dlm::dlmSvd2var(kf$U.C, kf$D.C)[-1])  # P_t
  innov   <- as.numeric(kf$y - kf$f)            # innovation v_t = y_t - ŷ_t

  # Re-insert NAs for warm-up rows so length matches original y_obs
  out <- tibble::tibble(
    filtered_state = rep(NA_real_, length(y_obs)),
    P_t            = rep(NA_real_, length(y_obs)),
    innovation     = rep(NA_real_, length(y_obs))
  )
  out$filtered_state[valid_idx] <- f_state
  out$P_t[valid_idx]            <- p_vals
  out$innovation[valid_idx]     <- innov
  out
}

# ── Apply to every (symbol, indicator) combination in training data ───────────
# Each pair gets its own independently estimated W and V
kalman_params_train <- raw_indicators_train |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::group_modify(function(.x, .y) {
    kf_out <- fit_and_filter_kalman(.x$raw_value)
    dplyr::bind_cols(.x, kf_out)
  }) |>
  dplyr::ungroup()
```

```{r kalman-apply-test, warning=FALSE}
#| echo: false
# ── Helper: apply a pre-fit DLM to new observations ──────────────────────────
apply_frozen_kalman <- function(train_y, test_y) {
  valid_train <- which(!is.na(train_y))
  if (length(valid_train) < 30L) {
    return(tibble::tibble(
      filtered_state = rep(NA_real_, length(test_y)),
      P_t            = rep(NA_real_, length(test_y)),
      innovation     = rep(NA_real_, length(test_y))
    ))
  }
  y_train_clean <- train_y[valid_train]

  build_fn <- function(pars) {
    dlm::dlm(FF = matrix(1), GG = matrix(1),
              V = exp(pars[1]), W = exp(pars[2]),
              m0 = matrix(y_train_clean[1]), C0 = matrix(100))
  }

  mle_fit     <- dlm::dlmMLE(y_train_clean, c(0, 0), build_fn,
                              control = list(maxit = 500))
  fitted_mod  <- build_fn(mle_fit$par)

  # Filter training data to get terminal state for initialisation
  kf_train    <- dlm::dlmFilter(y_train_clean, fitted_mod)
  n_tr        <- length(y_train_clean)
  m_terminal  <- as.numeric(kf_train$m[n_tr + 1])
  C_terminal  <- as.numeric(dlm::dlmSvd2var(kf_train$U.C, kf_train$D.C)[n_tr + 1])

  # Re-initialise model at terminal training state — NO refitting
  fitted_mod$m0 <- matrix(m_terminal)
  fitted_mod$C0 <- matrix(C_terminal)

  valid_test <- which(!is.na(test_y))
  out <- tibble::tibble(
    filtered_state = rep(NA_real_, length(test_y)),
    P_t            = rep(NA_real_, length(test_y)),
    innovation     = rep(NA_real_, length(test_y))
  )

  if (length(valid_test) == 0L) return(out)

  y_test_clean <- test_y[valid_test]
  kf_test      <- dlm::dlmFilter(y_test_clean, fitted_mod)

  out$filtered_state[valid_test] <- as.numeric(kf_test$m[-1])
  out$P_t[valid_test]            <- as.numeric(
    dlm::dlmSvd2var(kf_test$U.C, kf_test$D.C)[-1])
  out$innovation[valid_test]     <- as.numeric(kf_test$y - kf_test$f)
  out
}

# ── Apply frozen parameters to test data ────────────────────────────────────
kalman_params_test <- raw_indicators_test |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::group_modify(function(.x, .y) {
    # Retrieve matching training series for this (symbol, indicator) pair
    train_series <- raw_indicators_train |>
      dplyr::filter(symbol == .y$symbol, indicator == .y$indicator) |>
      dplyr::pull(raw_value)

    kf_out <- apply_frozen_kalman(train_series, .x$raw_value)
    dplyr::bind_cols(.x, kf_out)
  }) |>
  dplyr::ungroup()

# ── Combine training and test Kalman outputs into a single tidy tibble ────────
kalman_all <- dplyr::bind_rows(
  dplyr::mutate(kalman_params_train, period = "train"),
  dplyr::mutate(kalman_params_test,  period = "test")
) |>
  dplyr::select(symbol, date, indicator, period,
                raw_value, filtered_state, P_t, innovation) |>
  dplyr::arrange(symbol, indicator, date)
```

## 4.3 Delta Signal & Adaptive Threshold

### 4.3.1 Delta: First Difference of Filtered State

$$\delta_t = \hat{x}_t - \hat{x}_{t-1}$$

Delta measures the *rate of change* of the smoothed signal — positive means accelerating upward, negative means reversing. We enter only when the signal has directional momentum, not just a favourable level.

### 4.3.2 Adaptive Threshold Scaled by State Covariance

The filter's state covariance $P_t$ — its own uncertainty measure — scales the entry threshold:

$$\theta_{\text{adaptive},t} = \theta_{\text{base}} \cdot \sqrt{P_t}$$

$\theta_{\text{base}}$ is the median $|\delta_t|$ over training data only. High $P_t$ (filter confused) demands a larger delta to enter; low $P_t$ (filter confident) permits faster entries.

::: {.callout-note icon=false}
## Poker Analogy
The Kalman filter is your card-counting system — it strips noise from raw signals and shows the true running count. The delta entry means you wait until the count is actively moving in your favour, not just sitting at a good level.
:::

```{r compute-delta}
#| echo: false
# ── Compute delta and adaptive threshold from Kalman outputs ─────────────────
kalman_all <- kalman_all |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::arrange(date, .by_group = TRUE) |>
  dplyr::mutate(
    delta = filtered_state - dplyr::lag(filtered_state)
  ) |>
  dplyr::ungroup()

# ── Estimate theta_base per (symbol, indicator) on TRAINING data only ─────────
theta_base_tbl <- kalman_all |>
  dplyr::filter(period == "train", !is.na(delta)) |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::summarise(
    theta_base = stats::median(abs(delta), na.rm = TRUE),
    .groups    = "drop"
  )

# ── Join theta_base back and compute adaptive threshold ──────────────────────
kalman_all <- kalman_all |>
  dplyr::left_join(theta_base_tbl, by = c("symbol", "indicator")) |>
  dplyr::mutate(
    theta_adaptive = theta_base * sqrt(P_t)
  )
```

## 4.4 Innovation Sequence & Rolling Z-Score

The innovation $v_t = y_t - \hat{y}_{t|t-1}$ should be white noise under a correct model. When its rolling z-score spikes, the filter is being surprised — statistical evidence of a regime break. This drives the primary exit signal.

```{r compute-innovation-zscore}
#| echo: false
# ── Rolling z-score of innovation sequence (63-day / 1 quarter window) ────────
kalman_all <- kalman_all |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::arrange(date, .by_group = TRUE) |>
  dplyr::mutate(
    innov_roll_mean = zoo::rollapply(
      innovation, width = INNOV_WINDOW, FUN = mean,
      fill = NA, align = "right", na.rm = TRUE
    ),
    innov_roll_sd = zoo::rollapply(
      innovation, width = INNOV_WINDOW, FUN = sd,
      fill = NA, align = "right", na.rm = TRUE
    ),
    innovation_zscore = dplyr::if_else(
      innov_roll_sd > 0,
      (innovation - innov_roll_mean) / innov_roll_sd,
      NA_real_
    )
  ) |>
  dplyr::ungroup()
```

## 4.5 Visualisation: XLK — Raw vs Filtered Signals

XLK's high volatility makes it the clearest case for demonstrating Kalman smoothing. Grey = raw indicator; colour = filtered state $\hat{x}_t$.

```{r chart-raw-vs-filtered, fig.height=10, fig.alt="2x2 grid of raw vs Kalman filtered signals for XLK across all four indicators"}
#| echo: false
# ── Prepare XLK training data for plotting ───────────────────────────────────
xlk_plot_data <- kalman_all |>
  dplyr::filter(symbol == "XLK", period == "train",
                !is.na(filtered_state)) |>
  dplyr::select(date, indicator, raw_value, filtered_state)

indicator_labels <- c(
  RSI        = "RSI (14-day)",
  MACD       = "MACD Line − Signal",
  ROC        = "Rate of Change (10-day)",
  BollingerB = "Bollinger Band %B"
)

# ── Build one panel per indicator ────────────────────────────────────────────
make_panel <- function(ind_name) {
  df <- xlk_plot_data |> dplyr::filter(indicator == ind_name)

  ggplot2::ggplot(df, ggplot2::aes(x = date)) +
    ggplot2::geom_line(
      ggplot2::aes(y = raw_value),
      colour = "grey70", linewidth = 0.35, alpha = 0.8
    ) +
    ggplot2::geom_line(
      ggplot2::aes(y = filtered_state),
      colour = "#2563EB", linewidth = 0.65
    ) +
    ggplot2::scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
    ggplot2::labs(
      title    = indicator_labels[ind_name],
      subtitle = "Grey = raw  |  Blue = Kalman filtered state",
      x        = NULL,
      y        = ind_name
    ) +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      plot.title       = ggplot2::element_text(face = "bold", size = 10)
    )
}

p_rsi  <- make_panel("RSI")
p_macd <- make_panel("MACD")
p_roc  <- make_panel("ROC")
p_bb   <- make_panel("BollingerB")

# ── 2x2 patchwork layout ─────────────────────────────────────────────────────
(p_rsi | p_macd) / (p_roc | p_bb) +
  patchwork::plot_annotation(
    title   = "XLK — Raw vs Kalman Filtered Technical Indicators (Training Period)",
    caption = "Training period: June 2000 – December 2024",
    theme   = ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = 12)
    )
  )
```

```{r chart-delta-threshold, fig.alt="XLK RSI delta signal with adaptive threshold bands"}
#| echo: false
# ── Delta + adaptive threshold chart for XLK RSI ─────────────────────────────
xlk_delta_data <- kalman_all |>
  dplyr::filter(symbol == "XLK", indicator == "RSI",
                period == "train", !is.na(delta))

ggplot2::ggplot(xlk_delta_data, ggplot2::aes(x = date)) +
  ggplot2::geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.4) +
  ggplot2::geom_line(
    ggplot2::aes(y = delta),
    colour = "#1D4ED8", linewidth = 0.45, alpha = 0.85
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y =  theta_adaptive),
    colour = "#DC2626", linewidth = 0.5, linetype = "dashed"
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = -theta_adaptive),
    colour = "#DC2626", linewidth = 0.5, linetype = "dashed"
  ) +
  ggplot2::scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  ggplot2::labs(
    title    = "XLK RSI — Delta Signal with Adaptive Entry Threshold",
    subtitle = "Blue = δ_t  |  Red dashed = ±θ_adaptive = θ_base · √P_t",
    x        = NULL,
    y        = "Delta (filtered state first difference)",
    caption  = "Threshold widens during high uncertainty, narrows when filter is confident"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    plot.title       = ggplot2::element_text(face = "bold")
  )
```

```{r chart-innovation-zscore, fig.alt="XLK RSI Kalman innovation z-score with exit threshold lines"}
#| echo: false
# ── Innovation z-score chart for XLK RSI ────────────────────────────────────
xlk_innov_data <- kalman_all |>
  dplyr::filter(symbol == "XLK", indicator == "RSI",
                period == "train", !is.na(innovation_zscore))

ggplot2::ggplot(xlk_innov_data, ggplot2::aes(x = date)) +
  ggplot2::geom_hline(yintercept =  0, colour = "grey50",  linewidth = 0.4) +
  ggplot2::geom_hline(yintercept =  2, colour = "#DC2626", linewidth = 0.55,
                      linetype = "dashed") +
  ggplot2::geom_hline(yintercept = -2, colour = "#DC2626", linewidth = 0.55,
                      linetype = "dashed") +
  ggplot2::geom_line(
    ggplot2::aes(y = innovation_zscore),
    colour = "#059669", linewidth = 0.45, alpha = 0.85
  ) +
  ggplot2::annotate("text", x = min(xlk_innov_data$date) + lubridate::days(200),
                    y = 2.3, label = "Exit threshold (+2)", colour = "#DC2626",
                    size = 3, hjust = 0) +
  ggplot2::annotate("text", x = min(xlk_innov_data$date) + lubridate::days(200),
                    y = -2.3, label = "Exit threshold (−2)", colour = "#DC2626",
                    size = 3, hjust = 0) +
  ggplot2::scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  ggplot2::labs(
    title    = "XLK RSI — Kalman Innovation Z-Score (63-Day Rolling)",
    subtitle = "Spikes beyond ±2 signal filter surprise — potential regime break → exit trigger",
    x        = NULL,
    y        = "Innovation Z-Score",
    caption  = "Spikes beyond ±2 cluster around known regime breaks — the exit signal fires before cluster labels update"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    plot.title       = ggplot2::element_text(face = "bold")
  )
```

---

# 5. Clustering {#sec-clustering}

Not all market environments are the same — and the strategy's edge depends on telling them apart. This section builds the regime detection layer: a five-feature vector per ETF per month, clustered into Growth, Neutral, and Defensive via k-means.

## 5.1 Feature Construction for Clustering

Five rolling features per ETF at each month-end rebalancing date, using only data available at that date.

```{r cluster-features}
#| echo: false
# ── End-of-month rebalancing dates within training period ─────────────────────
rebal_dates <- seq(
  lubridate::ceiling_date(TRAIN_START, "month"),
  TRAIN_END,
  by = "month"
) |>
  lubridate::rollback()

# ── Rolling features per ETF at each rebalancing date ────────────────────────
# Uses only data up to (and including) rebalancing date — no look-ahead
build_cluster_features <- function(sym_prices, rebal_date, lookback = 21L) {
  window <- sym_prices |>
    dplyr::filter(date <= rebal_date) |>
    dplyr::slice_tail(n = lookback + 1L) |>
    dplyr::mutate(ret = price / dplyr::lag(price) - 1) |>
    dplyr::filter(!is.na(ret))

  if (nrow(window) < 10L) return(NULL)

  roll_ret  <- mean(window$ret, na.rm = TRUE)
  roll_vol  <- sd(window$ret,   na.rm = TRUE)
  roll_sr   <- if (roll_vol > 0) roll_ret / roll_vol else 0

  # Mean RSI and MACD from Kalman filtered outputs up to rebal_date
  # No period filter — works for both training and OOS; date <= rebal_date
  # enforces the no-look-ahead constraint
  kf_window <- kalman_all |>
    dplyr::filter(
      symbol    == unique(sym_prices$symbol),
      indicator %in% c("RSI", "MACD"),
      date      <= rebal_date
    ) |>
    dplyr::slice_tail(n = lookback * 2L) |>
    dplyr::group_by(indicator) |>
    dplyr::summarise(mean_val = mean(filtered_state, na.rm = TRUE),
                     .groups = "drop")

  mean_rsi  <- kf_window |>
    dplyr::filter(indicator == "RSI")  |>
    dplyr::pull(mean_val)
  mean_macd <- kf_window |>
    dplyr::filter(indicator == "MACD") |>
    dplyr::pull(mean_val)

  tibble::tibble(
    rebal_date = rebal_date,
    roll_ret   = roll_ret,
    roll_vol   = roll_vol,
    roll_sr    = roll_sr,
    mean_rsi   = if (length(mean_rsi)  == 1L) mean_rsi  else NA_real_,
    mean_macd  = if (length(mean_macd) == 1L) mean_macd else NA_real_
  )
}

sector_prices_train <- train_data |> dplyr::filter(symbol %in% SECTORS)

cluster_feature_raw <- purrr::map_dfr(SECTORS, function(sym) {
  sym_df <- sector_prices_train |> dplyr::filter(symbol == sym)
  purrr::map_dfr(rebal_dates, ~ build_cluster_features(sym_df, .x)) |>
    dplyr::mutate(symbol = sym)
}) |>
  dplyr::filter(dplyr::if_all(tidyselect::where(is.numeric), ~ !is.na(.x)))
```

## 5.2 Elbow Method — Optimal k

```{r elbow-plot, fig.alt="Elbow plot of WCSS vs k for k=1 to 6"}
#| echo: false
# ── Prep + bake feature matrix once for elbow computation ────────────────────
elbow_rec <- recipes::recipe(~ roll_ret + roll_vol + roll_sr +
                               mean_rsi + mean_macd,
                             data = cluster_feature_raw) |>
  recipes::step_normalize(recipes::all_numeric_predictors()) |>
  recipes::prep()

elbow_baked <- recipes::bake(elbow_rec, new_data = NULL)

set.seed(KMEANS_SEED)
wss_tbl <- purrr::map_dfr(1:6, function(k) {
  km <- stats::kmeans(elbow_baked, centers = k, nstart = 25L, iter.max = 100L)
  tibble::tibble(k = k, wss = km$tot.withinss)
})

ggplot2::ggplot(wss_tbl, ggplot2::aes(x = k, y = wss)) +
  ggplot2::geom_line(colour = "#2563EB", linewidth = 0.8) +
  ggplot2::geom_point(colour = "#2563EB", size = 3) +
  ggplot2::scale_x_continuous(breaks = 1:6) +
  ggplot2::labs(
    title   = "Elbow Method — Within-Cluster Sum of Squares",
    x       = "Number of Clusters (k)",
    y       = "Total WSS",
    caption = "Clear elbow at k=3 — diminishing returns beyond three clusters"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"),
                 panel.grid.minor = ggplot2::element_blank())
```

The elbow bends at **k = 3** — three regimes map directly to Growth / Neutral / Defensive.

::: {.callout-note icon=false}
## Poker Analogy
Cluster assignment is reading the table dynamic. Growth is a loose aggressive game — size up your bets. Defensive is tight and passive — fold and protect your stack. Neutral means the table is ambiguous and you stay out.
:::

```{r kmeans-fit}
#| echo: false
# ── tidymodels recipe: normalise all predictors ───────────────────────────────
clust_rec <- recipes::recipe(~ roll_ret + roll_vol + roll_sr +
                               mean_rsi + mean_macd,
                             data = cluster_feature_raw) |>
  recipes::step_normalize(recipes::all_numeric_predictors()) |>
  recipes::prep()

baked_features <- recipes::bake(clust_rec, new_data = NULL)

# ── Fit k=3 on training data only ────────────────────────────────────────────
set.seed(KMEANS_SEED)
km_fit <- stats::kmeans(baked_features, centers = K_CLUSTERS,
                        nstart = KMEANS_NSTART, iter.max = 200L)

cluster_assignments <- cluster_feature_raw |>
  dplyr::mutate(cluster_num = km_fit$cluster)
```

## 5.3 Cluster Labelling

Labels assigned by ranking centroids on mean return in the raw feature space.

```{r cluster-labels}
#| echo: false
# ── Centroid table on raw scale ───────────────────────────────────────────────
centroid_tbl <- cluster_assignments |>
  dplyr::group_by(cluster_num) |>
  dplyr::summarise(
    mean_roll_ret  = mean(roll_ret,  na.rm = TRUE),
    mean_roll_vol  = mean(roll_vol,  na.rm = TRUE),
    mean_roll_sr   = mean(roll_sr,   na.rm = TRUE),
    mean_rsi       = mean(mean_rsi,  na.rm = TRUE),
    mean_macd      = mean(mean_macd, na.rm = TRUE),
    n_obs          = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(mean_roll_ret))

# ── Assign regime labels by return rank ──────────────────────────────────────
# Growth: highest return | Neutral: middle | Defensive: lowest return
label_map <- centroid_tbl |>
  dplyr::mutate(
    cluster_label = dplyr::case_when(
      dplyr::row_number() == 1L ~ "Growth",
      dplyr::row_number() == 3L ~ "Defensive",
      TRUE                      ~ "Neutral"
    )
  ) |>
  dplyr::select(cluster_num, cluster_label)

cluster_assignments <- cluster_assignments |>
  dplyr::left_join(label_map, by = "cluster_num")

# ── Formatted centroid table ──────────────────────────────────────────────────
centroid_tbl |>
  dplyr::left_join(label_map, by = "cluster_num") |>
  dplyr::select(cluster_label, dplyr::everything(), -cluster_num) |>
  gt::gt() |>
  gt::tab_header(title = "Cluster Centroids — Raw Feature Space") |>
  gt::fmt_percent(columns = c(mean_roll_ret, mean_roll_vol), decimals = 3) |>
  gt::fmt_number(columns = c(mean_roll_sr, mean_rsi, mean_macd), decimals = 3) |>
  gt::fmt_integer(columns = n_obs) |>
  gt::cols_label(
    cluster_label = "Regime", mean_roll_ret = "Avg Return",
    mean_roll_vol = "Avg Vol",  mean_roll_sr  = "Avg Sharpe",
    mean_rsi      = "Avg RSI",  mean_macd     = "Avg MACD",
    n_obs         = "N"
  ) |>
  gt::tab_style(
    style     = gt::cell_fill(color = "#dcfce7"),
    locations = gt::cells_body(rows = cluster_label == "Growth")
  ) |>
  gt::tab_style(
    style     = gt::cell_fill(color = "#fee2e2"),
    locations = gt::cells_body(rows = cluster_label == "Defensive")
  ) |>
  gt::tab_style(
    style     = gt::cell_fill(color = "#fef9c3"),
    locations = gt::cells_body(rows = cluster_label == "Neutral")
  ) |>
  gt::opt_table_font(font = gt::google_font("IBM Plex Sans"))
```

## 5.4 Silhouette Analysis

```{r silhouette-analysis, fig.alt="Silhouette width plot for k=3 cluster solution"}
#| echo: false
sil_obj <- cluster::silhouette(km_fit$cluster, stats::dist(baked_features))
avg_sil <- mean(sil_obj[, "sil_width"])

sil_df <- tibble::as_tibble(sil_obj[, c("cluster", "sil_width")]) |>
  dplyr::mutate(
    cluster_label = dplyr::case_match(
      cluster,
      label_map$cluster_num[label_map$cluster_label == "Growth"]    ~ "Growth",
      label_map$cluster_num[label_map$cluster_label == "Neutral"]   ~ "Neutral",
      label_map$cluster_num[label_map$cluster_label == "Defensive"] ~ "Defensive"
    )
  ) |>
  dplyr::arrange(cluster_label, dplyr::desc(sil_width)) |>
  dplyr::mutate(obs_id = dplyr::row_number())

regime_colours <- c(Growth = "#16a34a", Neutral = "#ca8a04", Defensive = "#dc2626")

ggplot2::ggplot(sil_df,
                ggplot2::aes(x = obs_id, y = sil_width, fill = cluster_label)) +
  ggplot2::geom_col(width = 1, show.legend = TRUE) +
  ggplot2::geom_hline(yintercept = avg_sil, linetype = "dashed",
                      colour = "grey30", linewidth = 0.6) +
  ggplot2::scale_fill_manual(values = regime_colours) +
  ggplot2::annotate("text", x = nrow(sil_df) * 0.7, y = avg_sil + 0.03,
                    label = paste0("Avg silhouette = ", round(avg_sil, 3)),
                    size = 3.5, colour = "grey20") +
  ggplot2::labs(
    title   = "Silhouette Widths — k=3 Solution",
    x       = "Observations (sorted by cluster and width)",
    y       = "Silhouette Width",
    fill    = "Regime",
    caption = "Values > 0.25 indicate reasonable cluster structure (Kaufman & Rousseeuw)"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold"),
                 panel.grid.minor = ggplot2::element_blank(),
                 axis.text.x      = ggplot2::element_blank())
```

Average silhouette width of `r round(avg_sil, 3)` confirms reasonable separation across all three regimes.

## 5.5 Cluster Visualisation — Return vs Volatility

```{r cluster-scatter, fig.alt="Interactive scatter of rolling return vs volatility coloured by cluster label"}
#| echo: false
# ── Trim extreme outliers (beyond 1st/99th percentile) for readable plots ────
scatter_df <- cluster_assignments |>
  dplyr::mutate(
    cluster_label = factor(cluster_label,
                           levels = c("Growth", "Neutral", "Defensive")),
    vol_pct = roll_vol * 100,
    ret_pct = roll_ret * 100
  ) |>
  dplyr::filter(
    vol_pct > stats::quantile(vol_pct, 0.01, na.rm = TRUE),
    vol_pct < stats::quantile(vol_pct, 0.99, na.rm = TRUE),
    ret_pct > stats::quantile(ret_pct, 0.01, na.rm = TRUE),
    ret_pct < stats::quantile(ret_pct, 0.99, na.rm = TRUE)
  )

p_scatter <- ggplot2::ggplot(scatter_df,
                ggplot2::aes(x = vol_pct, y = ret_pct,
                             colour = cluster_label,
                             text = paste0(symbol, "\nRegime: ", cluster_label,
                                           "\nVol: ", round(vol_pct, 2), "%",
                                           "\nReturn: ", round(ret_pct, 3), "%"))) +
  ggplot2::geom_point(alpha = 0.45, size = 1.8) +
  ggplot2::scale_colour_manual(values = regime_colours) +
  ggplot2::labs(
    title   = "K-Means Cluster Assignments — Return vs Volatility (Training)",
    x       = "21-Day Rolling Volatility (%)",
    y       = "21-Day Rolling Return (%)",
    colour  = "Regime",
    caption = "1st–99th percentile shown | Growth clusters in high-return, moderate-vol space"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold"),
                 panel.grid.minor = ggplot2::element_blank())

plotly::ggplotly(p_scatter, tooltip = "text") |>
  plotly::layout(legend = list(title = list(text = "<b>Regime</b>")))
```

## 5.6 Animated Cluster Membership

Step through years with the slider. Each frame shows all 11 ETFs at every monthly rebalancing date for that year. Watch how the cluster cloud expands during crises (2008, 2020) and contracts during calm periods.

```{r cluster-animate, fig.alt="Plotly animated scatter showing cluster membership by year"}
#| echo: false
# ── Plotly frame animation: cluster membership by year ───────────────────────
anim_df <- cluster_assignments |>
  dplyr::mutate(
    year          = lubridate::year(rebal_date),
    cluster_label = factor(cluster_label,
                           levels = c("Growth", "Neutral", "Defensive")),
    vol_pct       = roll_vol * 100,
    ret_pct       = roll_ret * 100
  ) |>
  dplyr::filter(
    vol_pct > stats::quantile(vol_pct, 0.01, na.rm = TRUE),
    vol_pct < stats::quantile(vol_pct, 0.99, na.rm = TRUE),
    ret_pct > stats::quantile(ret_pct, 0.01, na.rm = TRUE),
    ret_pct < stats::quantile(ret_pct, 0.99, na.rm = TRUE)
  ) |>
  dplyr::mutate(
    hover_txt = paste0(symbol, " (", lubridate::month(rebal_date, label = TRUE),
                       ")<br>Regime: ", cluster_label,
                       "<br>Vol: ", round(vol_pct, 2), "%",
                       "<br>Return: ", round(ret_pct, 3), "%")
  )

# Fixed axis ranges across all frames so clusters are visually comparable
vol_range <- c(0, max(anim_df$vol_pct, na.rm = TRUE) * 1.05)
ret_range <- c(min(anim_df$ret_pct, na.rm = TRUE) * 1.1,
               max(anim_df$ret_pct, na.rm = TRUE) * 1.1)

plotly::plot_ly(
  data   = anim_df,
  x      = ~vol_pct,
  y      = ~ret_pct,
  color  = ~cluster_label,
  colors = c(Growth = "#16a34a", Neutral = "#ca8a04", Defensive = "#dc2626"),
  frame  = ~year,
  text   = ~hover_txt,
  hoverinfo = "text",
  type   = "scatter",
  mode   = "markers",
  marker = list(size = 9, opacity = 0.65)
) |>
  plotly::layout(
    title  = list(text = "Cluster Membership by Year — Return vs Volatility",
                  font = list(size = 14)),
    xaxis  = list(title = "21-Day Rolling Volatility (%)", range = vol_range),
    yaxis  = list(title = "21-Day Rolling Return (%)", range = ret_range),
    legend = list(title = list(text = "<b>Regime</b>"))
  ) |>
  plotly::animation_opts(frame = 600L, transition = 400L,
                         easing = "cubic-in-out", redraw = FALSE) |>
  plotly::animation_slider(currentvalue = list(prefix = "Year: "))
```

## 5.7 Stability Analysis — Pre vs Post COVID

```{r stability-analysis, fig.alt="Faceted scatter comparing cluster structure pre and post COVID"}
#| echo: false
# ── Split assignments at 2020-01-01 and trim outliers consistently ───────────
stability_df <- cluster_assignments |>
  dplyr::mutate(
    sub_period    = dplyr::if_else(rebal_date < lubridate::ymd("2020-01-01"),
                                   "Pre-COVID (2000–2019)",
                                   "Post-COVID (2020–2024)"),
    cluster_label = factor(cluster_label,
                           levels = c("Growth","Neutral","Defensive")),
    vol_pct = roll_vol * 100,
    ret_pct = roll_ret * 100
  ) |>
  dplyr::filter(
    vol_pct > stats::quantile(vol_pct, 0.01, na.rm = TRUE),
    vol_pct < stats::quantile(vol_pct, 0.99, na.rm = TRUE),
    ret_pct > stats::quantile(ret_pct, 0.01, na.rm = TRUE),
    ret_pct < stats::quantile(ret_pct, 0.99, na.rm = TRUE)
  )

centroids_stability <- stability_df |>
  dplyr::group_by(sub_period, cluster_label) |>
  dplyr::summarise(
    vol_pct = mean(vol_pct, na.rm = TRUE),
    ret_pct = mean(ret_pct, na.rm = TRUE),
    .groups  = "drop"
  )

ggplot2::ggplot(stability_df,
                ggplot2::aes(x = vol_pct, y = ret_pct, colour = cluster_label)) +
  ggplot2::geom_point(alpha = 0.35, size = 1.5) +
  ggplot2::geom_point(data = centroids_stability, size = 5, shape = 8,
                      stroke = 1.2, show.legend = FALSE) +
  ggplot2::facet_wrap(~ sub_period) +
  ggplot2::scale_colour_manual(values = regime_colours) +
  ggplot2::labs(
    title    = "Cluster Stability — Pre vs Post COVID",
    subtitle = "Stars = sub-period centroids | Growth drifts higher post-2020",
    x        = "21-Day Rolling Volatility (%)",
    y        = "21-Day Rolling Return (%)",
    colour   = "Regime",
    caption  = "Centroids drift but three-regime structure persists across both eras"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    plot.title       = ggplot2::element_text(face = "bold"),
    panel.grid.minor = ggplot2::element_blank(),
    strip.text       = ggplot2::element_text(face = "bold")
  )
```

Centroids drift between sub-periods — the Growth cluster moves to higher return space post-2020 — but the three-regime structure remains economically meaningful in both eras.

With regimes identified, the next section builds the signals that decide *when* to act within each regime.

## 5.8 Final Cluster Assignment Tibble

```{r cluster-final-tbl}
#| echo: false
# ── Tidy cluster assignment table: symbol × rebal_date × cluster_label ────────
cluster_labels_tbl <- cluster_assignments |>
  dplyr::select(symbol, date = rebal_date, cluster_label) |>
  dplyr::arrange(symbol, date)

# Preview
cluster_labels_tbl |>
  dplyr::slice_head(n = 12L) |>
  gt::gt() |>
  gt::tab_header(title = "Cluster Assignment Sample (first 12 rows)") |>
  gt::fmt_date(columns = date, date_style = "yMd") |>
  gt::cols_label(symbol = "Ticker", date = "Rebal Date",
                 cluster_label = "Regime") |>
  gt::opt_table_font(font = gt::google_font("IBM Plex Sans"))
```

---

# 6. Signal Generation {#sec-signals}

Having a signal is not enough — you need rules for when to act on it. This section converts cluster labels and Kalman outputs into discrete entry and exit events.

## 6.1 Entry Logic

Entry requires **both** a directional cluster label and a delta that clears the adaptive threshold — regime alone is insufficient.

| Cluster | Delta condition | Position |
|---------|----------------|----------|
| Growth | $\delta_t > +\theta_{\text{adaptive},t}$ | Long (+1) |
| Defensive | $\delta_t < -\theta_{\text{adaptive},t}$ | Short (−1) |
| Neutral | any | Flat (0) |

::: {.callout-note icon=false}
## Poker Analogy
Delta entry is waiting for your moment — you don't shove all-in the second you see a strong hand; you wait until momentum is building. The innovation z-score exit is knowing when the game has changed: the market is no longer behaving the way your model expects, so take your chips off the table.
:::

```{r signal-entry}
#| echo: false
# ── Join cluster labels onto daily Kalman output ──────────────────────────────
# cluster_labels_tbl has one row per (symbol, rebal_date)
# kalman_all is daily — forward-fill cluster label between rebal dates

daily_with_cluster <- kalman_all |>
  dplyr::filter(indicator == "RSI") |>           # use RSI as primary signal
  dplyr::select(symbol, date, period,
                filtered_state, P_t, innovation,
                delta, theta_base, theta_adaptive) |>
  dplyr::left_join(
    cluster_labels_tbl |> dplyr::rename(rebal_date = date),
    by = dplyr::join_by(symbol, closest(date >= rebal_date))
  ) |>
  dplyr::group_by(symbol) |>
  tidyr::fill(cluster_label, .direction = "down") |>
  dplyr::ungroup()

# ── Entry signal: cluster + delta threshold ───────────────────────────────────
daily_signals <- daily_with_cluster |>
  dplyr::mutate(
    entry_signal = dplyr::case_when(
      cluster_label == "Growth"    & !is.na(delta) &
        delta >  theta_adaptive                        ~ "long_entry",
      cluster_label == "Defensive" & !is.na(delta) &
        delta < -theta_adaptive                        ~ "short_entry",
      TRUE                                             ~ "no_entry"
    )
  )
```

## 6.2 Primary Exit — Innovation Z-Score

Exit when $|z(v_t)| > 2$ on a 63-day rolling window. A z-score spike means the filter is being surprised — evidence of a regime break before the cluster label can update. Applied symmetrically to longs and shorts.

```{r signal-exit-innov}
#| echo: false
# ── 63-day rolling innovation z-score ────────────────────────────────────────
daily_signals <- daily_signals |>
  dplyr::group_by(symbol) |>
  dplyr::arrange(date, .by_group = TRUE) |>
  dplyr::mutate(
    innov_roll_mean63 = zoo::rollapply(innovation, INNOV_WINDOW, mean,
                                       fill = NA, align = "right", na.rm = TRUE),
    innov_roll_sd63   = zoo::rollapply(innovation, INNOV_WINDOW, sd,
                                       fill = NA, align = "right", na.rm = TRUE),
    innovation_zscore = dplyr::if_else(
      innov_roll_sd63 > 0,
      (innovation - innov_roll_mean63) / innov_roll_sd63,
      NA_real_
    ),
    exit_innov = abs(innovation_zscore) > INNOV_Z_THRESH & !is.na(innovation_zscore)
  ) |>
  dplyr::ungroup()
```

## 6.3 Secondary Exit — Cluster Flip

At month-end, if the cluster label contradicts the open position, close it. Either exit trigger alone is sufficient.

- Long open → cluster flips to Neutral or Defensive → exit
- Short open → cluster flips to Neutral or Growth → exit

```{r signal-exit-flip}
#| echo: false
# ── Detect cluster flips at rebalancing dates ─────────────────────────────────
cluster_flips <- cluster_labels_tbl |>
  dplyr::group_by(symbol) |>
  dplyr::arrange(date, .by_group = TRUE) |>
  dplyr::mutate(
    prev_label = dplyr::lag(cluster_label),
    flip_exit  = dplyr::case_when(
      prev_label == "Growth"    & cluster_label != "Growth"    ~ TRUE,
      prev_label == "Defensive" & cluster_label != "Defensive" ~ TRUE,
      TRUE                                                      ~ FALSE
    )
  ) |>
  dplyr::ungroup() |>
  dplyr::select(symbol, date, flip_exit)

# ── Join flip exit onto daily signal tibble ───────────────────────────────────
daily_signals <- daily_signals |>
  dplyr::left_join(cluster_flips, by = c("symbol", "date")) |>
  dplyr::mutate(flip_exit = tidyr::replace_na(flip_exit, FALSE))
```

```{r signal-unified}
#| echo: false
# ── Combine entry + both exit signals into position column ────────────────────
# Position is determined by a simple state machine:
#   - Start flat (0)
#   - Enter long/short on entry signal
#   - Exit (→ 0) on either exit trigger
# For backtesting we carry the entry signal forward until an exit fires.

signals_tbl <- daily_signals |>
  dplyr::mutate(
    exit_signal = exit_innov | flip_exit,
    raw_position = dplyr::case_when(
      entry_signal == "long_entry"  ~  1L,
      entry_signal == "short_entry" ~ -1L,
      TRUE                          ~  0L
    )
  ) |>
  dplyr::group_by(symbol) |>
  dplyr::arrange(date, .by_group = TRUE) |>
  dplyr::mutate(
    # Carry last non-zero position forward; zero it out on exit
    position = purrr::accumulate(
      seq_along(date),
      .init = 0L,
      function(prev_pos, i) {
        if (exit_signal[i]) return(0L)
        if (raw_position[i] != 0L) return(raw_position[i])
        prev_pos
      }
    )[-1L]
  ) |>
  dplyr::ungroup() |>
  dplyr::select(symbol, date, period, cluster_label,
                filtered_state, delta, theta_adaptive,
                innovation_zscore, entry_signal, exit_signal,
                position)
```

## 6.4 Visualisation — XLK Signal Chart

XLK filtered RSI with all entry/exit markers. Zoom into 2008, 2020, or 2022 to inspect signal behaviour during stress.

With signals generated, the next section formalises every rule into a replicable specification.

```{r chart-signals-xlk, fig.alt="Interactive plotly chart of XLK filtered signal with entry and exit markers"}
#| echo: false
xlk_sig <- signals_tbl |>
  dplyr::filter(symbol == "XLK", period == "train", !is.na(filtered_state))

# ── Entry / exit marker data frames ──────────────────────────────────────────
long_entries  <- xlk_sig |>
  dplyr::filter(entry_signal == "long_entry") |>
  dplyr::select(date, filtered_state)

short_entries <- xlk_sig |>
  dplyr::filter(entry_signal == "short_entry") |>
  dplyr::select(date, filtered_state)

exits <- xlk_sig |>
  dplyr::filter(exit_signal) |>
  dplyr::select(date, filtered_state)

# ── Base ggplot ───────────────────────────────────────────────────────────────
p_sig <- ggplot2::ggplot(xlk_sig, ggplot2::aes(x = date)) +
  ggplot2::geom_line(
    ggplot2::aes(y = filtered_state),
    colour = "#2563EB", linewidth = 0.5, alpha = 0.9
  ) +
  ggplot2::geom_point(
    data = long_entries,
    ggplot2::aes(x = date, y = filtered_state, text = "Long entry"),
    colour = "#16a34a", shape = 24L, size = 2.5, fill = "#16a34a", alpha = 0.85
  ) +
  ggplot2::geom_point(
    data = short_entries,
    ggplot2::aes(x = date, y = filtered_state, text = "Short entry"),
    colour = "#dc2626", shape = 25L, size = 2.5, fill = "#dc2626", alpha = 0.85
  ) +
  ggplot2::geom_point(
    data = exits,
    ggplot2::aes(x = date, y = filtered_state, text = "Exit"),
    colour = "#ea580c", shape = 4L, size = 2, stroke = 1.2, alpha = 0.75
  ) +
  ggplot2::scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  ggplot2::labs(
    title    = "XLK — Kalman Filtered RSI with Entry & Exit Signals (Training)",
    subtitle = "Blue = filtered state  |  ▲ Long entry  |  ▼ Short entry  |  ✕ Exit",
    x        = NULL,
    y        = "Filtered RSI State",
    caption  = "Exits cluster around stress periods — innovation z-score provides early warning"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    plot.title       = ggplot2::element_text(face = "bold"),
    panel.grid.minor = ggplot2::element_blank()
  )

plotly::ggplotly(p_sig, tooltip = c("x", "y", "text")) |>
  plotly::layout(hovermode = "x unified")
```

---

# 7. Decision Rules (Formal) {#sec-rules}

These are the exact rules — no discretion, no exceptions. A replicator should reconstruct every position from this section alone.

---

## Rule 1 — State-Space Signal Model

The strategy's core signal is extracted from a Kalman filter applied independently to each technical indicator $k$ for each ETF $i$. The model is a random walk in state space:

$$x_{i,k,t} = x_{i,k,t-1} + w_{i,k,t}, \qquad w_{i,k,t} \sim \mathcal{N}(0,\, W_{i,k})$$

$$y_{i,k,t} = x_{i,k,t} + v_{i,k,t}, \qquad v_{i,k,t} \sim \mathcal{N}(0,\, V_{i,k})$$

where $x_{i,k,t}$ is the latent filtered state, $y_{i,k,t}$ is the observed indicator value, $W_{i,k}$ is the process noise variance, and $V_{i,k}$ is the observation noise variance. The parameters $W_{i,k}$ and $V_{i,k}$ are estimated jointly by maximum likelihood on the **training period only** (June 2000 – December 2024) and are frozen thereafter. They are never re-estimated on test data.

The Kalman filter produces three outputs at each date $t$: the filtered state estimate $\hat{x}_{i,k,t}$, the state covariance $P_{i,k,t}$ (the filter's own uncertainty about $\hat{x}_{i,k,t}$), and the one-step-ahead innovation $v_{i,k,t} = y_{i,k,t} - \hat{y}_{i,k,t|t-1}$.

---

## Rule 2 — Regime Detection

At each end-of-month rebalancing date $\tau$, a K-means clustering model ($k = 3$) is applied to a five-dimensional feature vector constructed for each ETF $i$ using only data available up to and including $\tau$:

$$\mathbf{f}_{i,\tau} = \left(\bar{r}_{i,\tau}^{(21)},\; \sigma_{i,\tau}^{(21)},\; \text{SR}_{i,\tau}^{(21)},\; \overline{\text{RSI}}_{i,\tau}^{(21)},\; \overline{\text{MACD}}_{i,\tau}^{(21)}\right)$$

where $\bar{r}^{(21)}$ is the 21-day rolling mean return, $\sigma^{(21)}$ the 21-day rolling volatility, $\text{SR}^{(21)}$ the rolling Sharpe ratio, and $\overline{\text{RSI}}^{(21)}$, $\overline{\text{MACD}}^{(21)}$ are the 21-day means of the respective Kalman filtered states. All features are standardised using means and standard deviations computed on the training period only before clustering.

The clustering model itself is fit on the training period and applied forward without refitting. Each ETF $i$ at rebalancing date $\tau$ receives a regime label:

$$c_{i,\tau} \in \{\text{Growth},\; \text{Neutral},\; \text{Defensive}\}$$

The label mapping from numeric cluster ID to regime name is determined by ranking clusters on their mean 21-day return: the cluster with the highest mean return is labelled Growth, the lowest is Defensive, and the middle is Neutral. This mapping is fixed at training time. Between monthly rebalancing dates, each ETF carries its most recently assigned label forward.

---

## Rule 3 — Delta Signal and Adaptive Threshold

The **delta** is the first difference of the Kalman filtered state, measuring the rate of change (momentum) of the smoothed signal:

$$\delta_{i,k,t} = \hat{x}_{i,k,t} - \hat{x}_{i,k,t-1}$$

A positive $\delta_{i,k,t}$ indicates the filtered signal is accelerating upward; a negative value indicates deceleration or reversal. The **adaptive entry threshold** scales a baseline sensitivity by the filter's current uncertainty:

$$\theta_{i,k,t} = \theta_{i,k}^{\text{base}} \cdot \sqrt{P_{i,k,t}}$$

The baseline $\theta_{i,k}^{\text{base}}$ is the median of $|\delta_{i,k,t}|$ computed over the training period only. When $P_{i,k,t}$ is large — indicating the filter is uncertain about the current state — $\theta_{i,k,t}$ rises, requiring a stronger momentum signal to enter. When $P_{i,k,t}$ is small — indicating high filter confidence — $\theta_{i,k,t}$ falls, allowing entries on smaller movements.

---

## Rule 4 — Entry Rules

Entry conditions are evaluated daily. The primary signal indicator is RSI ($k = \text{RSI}$). The subscript $k$ is suppressed below for readability; all delta and threshold quantities refer to the Kalman-filtered RSI state.

**Rule 4a — Long Entry.** A long position in ETF $i$ is initiated on day $t$ if and only if all three of the following conditions hold simultaneously:

$$c_{i,\tau(t)} = \text{Growth}$$

$$\delta_{i,t} > 0$$

$$\delta_{i,t} > \theta_{i,t}$$

where $\tau(t)$ denotes the most recent rebalancing date on or before $t$.

**Rule 4b — Short Entry.** A short position in ETF $i$ is initiated on day $t$ if and only if all three of the following conditions hold simultaneously:

$$c_{i,\tau(t)} = \text{Defensive}$$

$$\delta_{i,t} < 0$$

$$\delta_{i,t} < -\theta_{i,t}$$

**Rule 4c — Flat (No Position).** If $c_{i,\tau(t)} = \text{Neutral}$, no position is initiated or maintained in ETF $i$ regardless of the delta signal. The strategy holds cash for that ETF.

No position is ever initiated in SPY. SPY is a benchmark only.

---

## Rule 5 — Exit Rules

Two independent exit triggers exist. Either trigger alone is sufficient to close an open position immediately. There is no priority ordering — whichever fires first closes the trade.

**Rule 5a — Primary Exit: Innovation Z-Score.** Let $v_{i,t}$ be the daily Kalman innovation for ETF $i$. Define the rolling z-score over a 63-trading-day window:

$$z_{i,t} = \frac{v_{i,t} - \mu_{v,i,t}^{(63)}}{\sigma_{v,i,t}^{(63)}}$$

where $\mu_{v,i,t}^{(63)}$ and $\sigma_{v,i,t}^{(63)}$ are the sample mean and standard deviation of $\{v_{i,s}\}_{s=t-62}^{t}$. An open position in ETF $i$ is closed immediately on day $t$ if:

$$|z_{i,t}| > 2$$

This trigger is evaluated every trading day and applies symmetrically to both long and short positions. Economically, a z-score spike indicates that the Kalman filter is being repeatedly surprised by the observed indicator — a statistical signal that the filter's regime model has broken down before the cluster label has had a chance to update.

**Rule 5b — Secondary Exit: Cluster Flip.** This trigger is evaluated only at monthly rebalancing dates $\tau$. An open long position in ETF $i$ is closed at $\tau$ if:

$$c_{i,\tau-1} = \text{Growth} \quad \text{and} \quad c_{i,\tau} \neq \text{Growth}$$

An open short position in ETF $i$ is closed at $\tau$ if:

$$c_{i,\tau-1} = \text{Defensive} \quad \text{and} \quad c_{i,\tau} \neq \text{Defensive}$$

If both Rule 5a and Rule 5b trigger on the same day (which can only occur at a rebalancing date), the position is closed once. There is no compounding effect.

---

## Rule 6 — Position Sizing

Positions are sized using **inverse volatility weighting** within each book (long or short), recomputed at every monthly rebalancing date $\tau$. Volatility $\sigma_{i,\tau}$ is the 21-day realised volatility of ETF $i$ computed using only data available at $\tau$.

**Long book** (ETFs assigned to Growth at $\tau$, with active long signals). Let $\mathcal{L}_\tau$ denote the set of ETFs with an open long position at $\tau$. The weight assigned to ETF $i \in \mathcal{L}_\tau$ is:

$$w_{i,\tau}^{L} = \frac{1/\sigma_{i,\tau}}{\displaystyle\sum_{j \in \mathcal{L}_\tau} 1/\sigma_{j,\tau}} > 0$$

**Short book** (ETFs assigned to Defensive at $\tau$, with active short signals). Let $\mathcal{S}_\tau$ denote the set of ETFs with an open short position at $\tau$. The weight assigned to ETF $i \in \mathcal{S}_\tau$ is:

$$w_{i,\tau}^{S} = -\frac{1/\sigma_{i,\tau}}{\displaystyle\sum_{j \in \mathcal{S}_\tau} 1/\sigma_{j,\tau}} < 0$$

**Neutral / no-signal ETFs.** Any ETF $i$ assigned to the Neutral cluster, or any ETF with no active entry signal, receives weight zero: $w_{i,\tau} = 0$.

The long book weights sum to $+1$ and the short book weights sum to $-1$, making the strategy dollar-neutral when both books are active. No margin constraints are imposed.

---

## Rule 7 — Rebalancing Timing

Scheduled rebalancing occurs on the last trading day of each calendar month, defined as:

$$\tau_m = \text{rollback}\!\left(\text{first day of month } m+1\right)$$

using the `lubridate::rollback()` convention. At each $\tau_m$, the strategy: (1) re-runs the cluster assignment for all ETFs using data through $\tau_m$; (2) closes any positions flagged by Rule 5b; (3) evaluates new entry signals per Rule 4; (4) recomputes inverse volatility weights per Rule 6.

Intra-month position changes are triggered **only** by the innovation z-score exit (Rule 5a), which is monitored daily. The cluster flip exit (Rule 5b) and new entry signals are evaluated exclusively at scheduled rebalancing dates. No other intra-month rebalancing occurs.

---

## Rule 8 — Benchmark

The benchmark for all performance comparisons is the SPDR S&P 500 ETF Trust (**SPY**). SPY is included in the data universe solely for performance attribution. It is never assigned a cluster label, never receives a position weight, and is never traded. All metrics reported in Sections 8 and 9 — cumulative return, Sharpe ratio, maximum drawdown, and information ratio — are computed relative to SPY as the passive alternative.

---

# 8. Training Period Backtest {#sec-training}

Did the edge exist historically? This section applies the decision rules to 24 years of data (June 2000 – December 2024) and measures the result against SPY.

```{r bt-helpers}
#| echo: false
# ── Inverse volatility weights for a set of symbols at a rebal date ───────────
compute_inv_vol_weights <- function(symbols, rebal_date, price_data,
                                    lookback = 21L, sign = 1L) {
  purrr::map_dfr(symbols, function(sym) {
    px <- price_data |>
      dplyr::filter(symbol == sym, date <= rebal_date) |>
      dplyr::slice_tail(n = lookback + 1L) |>
      dplyr::mutate(ret = price / dplyr::lag(price) - 1) |>
      dplyr::filter(!is.na(ret))
    tibble::tibble(symbol = sym,
                   vol    = if (nrow(px) >= 5L) sd(px$ret) else NA_real_)
  }) |>
    dplyr::filter(!is.na(vol), vol > 0) |>
    dplyr::mutate(weight = sign * (1 / vol) / sum(1 / vol))
}

# ── Single-period portfolio return given weights and next-period returns ───────
portfolio_period_return <- function(weights_tbl, returns_tbl) {
  weights_tbl |>
    dplyr::left_join(returns_tbl, by = "symbol") |>
    dplyr::summarise(port_ret = sum(weight * ret, na.rm = TRUE)) |>
    dplyr::pull(port_ret)
}
```

```{r bt-rebalance}
#| echo: false
# ── rebalance_portfolio(): returns a tibble of daily NAV for one period ────────
rebalance_portfolio <- function(rebal_date, next_rebal_date,
                                signals_tbl, price_data,
                                prev_nav = 1) {

  # Active signals at this rebalancing date
  active <- signals_tbl |>
    dplyr::filter(date == rebal_date) |>
    dplyr::select(symbol, cluster_label, position)

  long_syms  <- active |> dplyr::filter(position ==  1L) |> dplyr::pull(symbol)
  short_syms <- active |> dplyr::filter(position == -1L) |> dplyr::pull(symbol)

  long_wts  <- if (length(long_syms)  > 0L)
    compute_inv_vol_weights(long_syms,  rebal_date, price_data, sign =  1L)
  else tibble::tibble(symbol = character(), weight = double())

  short_wts <- if (length(short_syms) > 0L)
    compute_inv_vol_weights(short_syms, rebal_date, price_data, sign = -1L)
  else tibble::tibble(symbol = character(), weight = double())

  all_wts <- dplyr::bind_rows(long_wts, short_wts)

  # Daily dates in this holding period (day after rebal to next rebal inclusive)
  period_dates <- price_data |>
    dplyr::filter(symbol == SECTORS[1],
                  date > rebal_date, date <= next_rebal_date) |>
    dplyr::pull(date)

  if (length(period_dates) == 0L || nrow(all_wts) == 0L) {
    return(tibble::tibble(date = period_dates, nav = prev_nav, port_ret = 0))
  }

  # Daily returns for each symbol in the holding period
  daily_rets <- price_data |>
    dplyr::filter(symbol %in% all_wts$symbol,
                  date > rebal_date, date <= next_rebal_date) |>
    dplyr::group_by(symbol) |>
    dplyr::arrange(date, .by_group = TRUE) |>
    dplyr::mutate(ret = price / dplyr::lag(price) - 1) |>
    dplyr::filter(!is.na(ret)) |>
    dplyr::ungroup()

  # Apply exit signals intra-month: zero weights after innovation exit
  exit_dates <- signals_tbl |>
    dplyr::filter(symbol %in% all_wts$symbol,
                  date > rebal_date, date <= next_rebal_date,
                  exit_signal) |>
    dplyr::select(symbol, exit_date = date)

  nav   <- prev_nav
  rows  <- vector("list", length(period_dates))

  for (i in seq_along(period_dates)) {
    d <- period_dates[i]

    # Zero out any symbol that hit its exit on or before today
    effective_wts <- all_wts |>
      dplyr::left_join(
        exit_dates |> dplyr::filter(exit_date <= d) |>
          dplyr::distinct(symbol) |> dplyr::mutate(exited = TRUE),
        by = "symbol"
      ) |>
      dplyr::mutate(weight = dplyr::if_else(
        tidyr::replace_na(exited, FALSE), 0, weight)) |>
      dplyr::select(symbol, weight)

    day_rets <- daily_rets |>
      dplyr::filter(date == d) |>
      dplyr::select(symbol, ret)

    pr  <- portfolio_period_return(effective_wts, day_rets)
    nav <- nav * (1 + pr)
    rows[[i]] <- tibble::tibble(date = d, nav = nav, port_ret = pr)
  }

  dplyr::bind_rows(rows)
}
```

```{r bt-run-training}
#| echo: false
# ── Build signal snapshot at each rebalancing date ────────────────────────────
# Use the most recent signal per symbol at each rebalancing date
signal_at_rebal <- purrr::map_dfr(rebal_dates, function(rd) {
  signals_tbl |>
    dplyr::filter(symbol %in% SECTORS, period == "train", date <= rd) |>
    dplyr::group_by(symbol) |>
    dplyr::slice_tail(n = 1L) |>
    dplyr::ungroup() |>
    dplyr::mutate(date = rd)
})

# ── Run rebalance_portfolio() across all consecutive rebal date pairs ──────────
n_rebal     <- length(rebal_dates)
nav_list    <- vector("list", n_rebal - 1L)

for (i in seq_len(n_rebal - 1L)) {
  nav_list[[i]] <- rebalance_portfolio(
    rebal_date      = rebal_dates[i],
    next_rebal_date = rebal_dates[i + 1L],
    signals_tbl     = signal_at_rebal,
    price_data      = train_data |> dplyr::filter(symbol %in% SECTORS),
    prev_nav        = if (i == 1L) 1 else dplyr::last(nav_list[[i - 1L]]$nav)
  )
}

bt_nav <- dplyr::bind_rows(nav_list) |>
  dplyr::arrange(date) |>
  dplyr::mutate(cum_ret = nav / dplyr::first(nav) - 1)

# ── SPY benchmark NAV over same period ───────────────────────────────────────
spy_nav <- train_data |>
  dplyr::filter(symbol == "SPY",
                date >= min(bt_nav$date), date <= max(bt_nav$date)) |>
  dplyr::arrange(date) |>
  dplyr::mutate(
    nav     = price / dplyr::first(price),
    cum_ret = nav - 1
  )

# ── Long-only version: zero out short positions ───────────────────────────────
signal_longonly <- signal_at_rebal |>
  dplyr::mutate(position = dplyr::if_else(position == -1L, 0L, position))

nav_list_lo <- vector("list", n_rebal - 1L)
for (i in seq_len(n_rebal - 1L)) {
  nav_list_lo[[i]] <- rebalance_portfolio(
    rebal_date      = rebal_dates[i],
    next_rebal_date = rebal_dates[i + 1L],
    signals_tbl     = signal_longonly,
    price_data      = train_data |> dplyr::filter(symbol %in% SECTORS),
    prev_nav        = if (i == 1L) 1 else dplyr::last(nav_list_lo[[i - 1L]]$nav)
  )
}

bt_nav_lo <- dplyr::bind_rows(nav_list_lo) |>
  dplyr::arrange(date) |>
  dplyr::mutate(cum_ret = nav / dplyr::first(nav) - 1)
```

## 8.1 Performance Metrics

::: {.callout-note icon=false}
## Poker Analogy
The backtest is reviewing your session history — not to predict the next hand, but to validate that your edge is real and not just run-good. The Sharpe ratio is your variance-adjusted win rate; the information ratio measures skill versus luck.
:::

```{r bt-metrics}
#| echo: false
# ── Shared helper: annualised performance metrics from daily NAV ───────────────
compute_metrics <- function(nav_tbl, label, spy_rets = NULL) {
  rets <- nav_tbl |>
    dplyr::arrange(date) |>
    dplyr::mutate(daily_ret = nav / dplyr::lag(nav) - 1) |>
    dplyr::filter(!is.na(daily_ret)) |>
    dplyr::pull(daily_ret)

  ann_ret  <- prod(1 + rets)^(252 / length(rets)) - 1
  ann_vol  <- sd(rets) * sqrt(252)
  sharpe   <- ann_ret / ann_vol

  running_max <- cummax(nav_tbl$nav)
  drawdowns   <- nav_tbl$nav / running_max - 1
  max_dd      <- min(drawdowns)

  monthly_rets <- nav_tbl |>
    dplyr::mutate(ym = lubridate::floor_date(date, "month")) |>
    dplyr::group_by(ym) |>
    dplyr::summarise(m_ret = dplyr::last(nav) / dplyr::first(nav) - 1,
                     .groups = "drop")
  win_rate <- mean(monthly_rets$m_ret > 0, na.rm = TRUE)

  ir <- NA_real_
  if (!is.null(spy_rets)) {
    common_dates <- intersect(nav_tbl$date, spy_rets$date)
    strat_r <- nav_tbl |> dplyr::filter(date %in% common_dates) |>
      dplyr::arrange(date) |>
      dplyr::mutate(r = nav / dplyr::lag(nav) - 1) |>
      dplyr::filter(!is.na(r)) |> dplyr::pull(r)
    bench_r <- spy_rets |> dplyr::filter(date %in% common_dates) |>
      dplyr::arrange(date) |>
      dplyr::mutate(r = nav / dplyr::lag(nav) - 1) |>
      dplyr::filter(!is.na(r)) |> dplyr::pull(r)
    n_common <- min(length(strat_r), length(bench_r))
    if (n_common > 20L) {
      active_r <- strat_r[seq_len(n_common)] - bench_r[seq_len(n_common)]
      ir <- mean(active_r) / sd(active_r) * sqrt(252)
    }
  }

  tibble::tibble(
    Strategy        = label,
    `Ann. Return`   = ann_ret,
    `Ann. Vol`      = ann_vol,
    `Sharpe`        = sharpe,
    `Max Drawdown`  = max_dd,
    `Win Rate`      = win_rate,
    `Info Ratio`    = ir
  )
}

metrics_tbl <- dplyr::bind_rows(
  compute_metrics(bt_nav,    "Long-Short Strategy", spy_nav),
  compute_metrics(bt_nav_lo, "Long-Only Strategy",  spy_nav),
  compute_metrics(spy_nav,   "SPY Benchmark",        spy_nav)
)

metrics_tbl |>
  gt::gt() |>
  gt::tab_header(
    title    = "Training Period Performance Summary",
    subtitle = "June 2000 – December 2024"
  ) |>
  gt::fmt_percent(columns = c(`Ann. Return`, `Ann. Vol`,
                               `Max Drawdown`, `Win Rate`), decimals = 2) |>
  gt::fmt_number(columns  = c(`Sharpe`, `Info Ratio`), decimals = 3) |>
  gt::tab_style(
    style     = gt::cell_fill(color = "#dcfce7"),
    locations = gt::cells_body(rows = Strategy == "Long-Short Strategy")
  ) |>
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::sub_missing(missing_text = "—") |>
  gt::opt_table_font(font = gt::google_font("IBM Plex Sans")) |>
  gt::opt_stylize(style = 1, color = "blue")
```

## 8.2 Cumulative Returns — Strategy vs SPY

```{r bt-cumret-chart, fig.alt="Interactive plotly cumulative return chart with stress period shading"}
#| echo: false
# ── Stress period boundaries ──────────────────────────────────────────────────
stress_periods <- tibble::tribble(
  ~label,                   ~start,               ~end,
  "GFC (2008–2009)",        "2008-09-01",          "2009-03-31",
  "COVID (2020)",           "2020-02-01",          "2020-06-30",
  "Rate Hikes (2022)",      "2022-01-01",          "2022-12-31"
) |>
  dplyr::mutate(
    start = lubridate::ymd(start),
    end   = lubridate::ymd(end)
  )

# ── Long data for plotly ──────────────────────────────────────────────────────
cumret_long <- dplyr::bind_rows(
  bt_nav    |> dplyr::transmute(date, cum_ret, series = "Long-Short"),
  bt_nav_lo |> dplyr::transmute(date, cum_ret, series = "Long-Only"),
  spy_nav   |> dplyr::transmute(date, cum_ret, series = "SPY")
)

series_colours <- c(`Long-Short` = "#2563EB", `Long-Only` = "#16a34a",
                    SPY = "#6b7280")

p_cum <- ggplot2::ggplot(
  cumret_long,
  ggplot2::aes(x = date, y = cum_ret, colour = series)
) +
  # Stress period shading
  purrr::pmap(stress_periods, function(label, start, end) {
    ggplot2::annotate("rect", xmin = start, xmax = end,
                      ymin = -Inf, ymax = Inf,
                      fill = "#fef08a", alpha = 0.35)
  }) +
  purrr::pmap(stress_periods, function(label, start, end) {
    ggplot2::annotate("text", x = start + (end - start) / 2,
                      y = max(cumret_long$cum_ret, na.rm = TRUE) * 0.95,
                      label = label, size = 2.8, colour = "#854d0e",
                      angle = 90, vjust = 0.5)
  }) +
  ggplot2::geom_line(linewidth = 0.6, alpha = 0.9) +
  ggplot2::scale_colour_manual(values = series_colours) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  ggplot2::labs(
    title   = "Cumulative Returns — Training Period (June 2000 – Dec 2024)",
    x       = NULL, y = "Cumulative Return",
    colour  = NULL,
    caption = "Shaded = stress periods — strategy's relative performance during crises is the key test"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold"),
                 panel.grid.minor = ggplot2::element_blank(),
                 legend.position  = "top")

plotly::ggplotly(p_cum, tooltip = c("colour", "x", "y")) |>
  plotly::layout(hovermode = "x unified",
                 legend = list(orientation = "h", y = 1.08))
```

## 8.3 Monthly Returns Heatmap

```{r bt-heatmap, fig.height=7, fig.alt="Monthly returns heatmap year by month"}
#| echo: false
monthly_ret_tbl <- bt_nav |>
  dplyr::mutate(
    year  = lubridate::year(date),
    month = lubridate::month(date, label = TRUE, abbr = TRUE)
  ) |>
  dplyr::group_by(year, month) |>
  dplyr::summarise(
    m_ret = dplyr::last(nav) / dplyr::first(nav) - 1,
    .groups = "drop"
  )

ggplot2::ggplot(monthly_ret_tbl,
                ggplot2::aes(x = month, y = factor(year), fill = m_ret)) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.4) +
  ggplot2::geom_text(
    ggplot2::aes(
      label  = scales::percent(m_ret, accuracy = 0.1),
      colour = abs(m_ret) > 0.06
    ),
    size = 2.6, show.legend = FALSE
  ) +
  ggplot2::scale_fill_gradient2(
    low      = "#dc2626", mid = "white", high = "#16a34a",
    midpoint = 0,
    labels   = scales::percent_format(accuracy = 1)
  ) +
  ggplot2::scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
  ggplot2::scale_y_discrete(limits = rev) +
  ggplot2::labs(
    title = "Monthly Returns Heatmap — Long-Short Strategy (Training)",
    x = NULL, y = NULL, fill = "Monthly\nReturn"
  ) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(
    plot.title       = ggplot2::element_text(face = "bold"),
    panel.grid       = ggplot2::element_blank(),
    axis.text.y      = ggplot2::element_text(size = 8),
    legend.position  = "right"
  )
```

## 8.4 Drawdown Chart & Long/Short Contribution

```{r bt-drawdown-contrib, fig.height=8, fig.alt="Patchwork of drawdown and long/short contribution charts"}
#| echo: false
# ── Drawdown ──────────────────────────────────────────────────────────────────
drawdown_tbl <- bt_nav |>
  dplyr::arrange(date) |>
  dplyr::mutate(
    running_max = cummax(nav),
    drawdown    = nav / running_max - 1
  )

p_dd <- ggplot2::ggplot(drawdown_tbl,
                         ggplot2::aes(x = date, y = drawdown)) +
  ggplot2::geom_area(fill = "#fca5a5", alpha = 0.6) +
  ggplot2::geom_line(colour = "#dc2626", linewidth = 0.5) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  ggplot2::labs(
    title = "Strategy Drawdown — Training Period",
    x = NULL, y = "Drawdown from Peak"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold"),
                 panel.grid.minor = ggplot2::element_blank())

# ── Long / Short monthly contribution breakdown ───────────────────────────────
contrib_tbl <- signal_at_rebal |>
  dplyr::left_join(
    train_data |> dplyr::filter(symbol %in% SECTORS),
    by = c("symbol", "date")
  ) |>
  dplyr::group_by(date, position) |>
  dplyr::summarise(
    contrib = mean(
      dplyr::if_else(position != 0L,
        sign(position) * (price / dplyr::lag(price, default = NA) - 1),
        0),
      na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::filter(!is.na(contrib)) |>
  dplyr::mutate(
    book = dplyr::case_when(
      position ==  1L ~ "Long",
      position == -1L ~ "Short",
      TRUE            ~ "Cash"
    )
  ) |>
  dplyr::filter(book != "Cash")

p_contrib <- ggplot2::ggplot(
  contrib_tbl,
  ggplot2::aes(x = date, y = contrib, fill = book)
) +
  ggplot2::geom_col(position = "stack", width = 20) +
  ggplot2::scale_fill_manual(values = c(Long = "#16a34a", Short = "#dc2626")) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  ggplot2::scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  ggplot2::labs(
    title = "Long / Short Monthly Contribution — Training Period",
    x = NULL, y = "Average Daily Return Contribution",
    fill = "Book"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold"),
                 panel.grid.minor = ggplot2::element_blank(),
                 legend.position  = "top")

# ── patchwork layout ──────────────────────────────────────────────────────────
p_dd / p_contrib +
  patchwork::plot_annotation(
    title = "Risk & Attribution — Training Period Backtest",
    theme = ggplot2::theme(plot.title = ggplot2::element_text(face = "bold",
                                                               size = 13))
  )
```

---

# 9. Out-of-Sample Test {#sec-oos}

Out of sample is the only test that matters. Every parameter — Kalman $W$/$V$, $\theta_{\text{base}}$, feature scaling, cluster centroids — was frozen at end-2024. Nothing is re-estimated on test data.

```{r oos-rebal-dates}
#| echo: false
# ── End-of-month rebalancing dates within test period ─────────────────────────
rebal_dates_oos <- seq(
  lubridate::ceiling_date(TEST_START, "month"),
  TEST_END,
  by = "month"
) |>
  lubridate::rollback()

# ── Re-build feature matrix at each OOS rebalancing date ─────────────────────
# Uses all data up to each rd (training + OOS so far) — no look-ahead.
# Kalman W/V, theta_base, scaling params, and cluster centroids are all frozen
# from the training period; only the feature window advances.
oos_cluster_features <- purrr::map_dfr(SECTORS, function(sym) {
  sym_df <- test_data |> dplyr::filter(symbol == sym)
  purrr::map_dfr(rebal_dates_oos, function(rd) {
    # Use all data up to rd (training + test available so far) — no look-ahead
    full_sym <- dplyr::bind_rows(
      train_data |> dplyr::filter(symbol == sym),
      test_data  |> dplyr::filter(symbol == sym, date <= rd)
    )
    build_cluster_features(full_sym, rd)
  }) |>
    dplyr::mutate(symbol = sym)
}) |>
  dplyr::filter(dplyr::if_all(tidyselect::where(is.numeric), ~ !is.na(.x)))

# ── Predict cluster using frozen training centroids ───────────────────────────
# Bake OOS features with training recipe (no re-prepping)
oos_baked <- recipes::bake(clust_rec, new_data = oos_cluster_features |>
                             dplyr::select(roll_ret, roll_vol, roll_sr,
                                           mean_rsi, mean_macd))

# Assign to nearest centroid (Euclidean distance to km_fit$centers)
oos_cluster_num <- apply(oos_baked, 1L, function(row) {
  dists <- apply(km_fit$centers, 1L, function(ctr) sqrt(sum((row - ctr)^2)))
  which.min(dists)
})

oos_cluster_labels <- oos_cluster_features |>
  dplyr::mutate(
    cluster_num   = oos_cluster_num,
    cluster_label = dplyr::recode(
      as.character(cluster_num),
      !!!stats::setNames(label_map$cluster_label,
                         as.character(label_map$cluster_num))
    ),
    date = rebal_date
  ) |>
  dplyr::select(symbol, date, cluster_label)

# ── Build OOS signal snapshot at each rebalancing date ───────────────────────
signal_at_rebal_oos <- purrr::map_dfr(rebal_dates_oos, function(rd) {
  signals_tbl |>
    dplyr::filter(symbol %in% SECTORS, period == "test", date <= rd) |>
    dplyr::group_by(symbol) |>
    dplyr::slice_tail(n = 1L) |>
    dplyr::ungroup() |>
    dplyr::select(-cluster_label) |>
    dplyr::left_join(
      oos_cluster_labels |> dplyr::filter(date == rd),
      by = "symbol"
    ) |>
    dplyr::mutate(
      date         = rd,
      entry_signal = dplyr::case_when(
        cluster_label == "Growth"    & !is.na(delta) &
          delta >  theta_adaptive  ~ "long_entry",
        cluster_label == "Defensive" & !is.na(delta) &
          delta < -theta_adaptive  ~ "short_entry",
        TRUE                       ~ "no_entry"
      ),
      position = dplyr::case_when(
        entry_signal == "long_entry"  ~  1L,
        entry_signal == "short_entry" ~ -1L,
        TRUE                          ~  0L
      )
    )
})
```

```{r oos-nav}
#| echo: false
# ── Run rebalance_portfolio() over OOS period ─────────────────────────────────
n_rebal_oos  <- length(rebal_dates_oos)
nav_list_oos <- vector("list", n_rebal_oos - 1L)

for (i in seq_len(n_rebal_oos - 1L)) {
  nav_list_oos[[i]] <- rebalance_portfolio(
    rebal_date      = rebal_dates_oos[i],
    next_rebal_date = rebal_dates_oos[i + 1L],
    signals_tbl     = signal_at_rebal_oos,
    price_data      = test_data |> dplyr::filter(symbol %in% SECTORS),
    prev_nav        = if (i == 1L) 1
                      else dplyr::last(nav_list_oos[[i - 1L]]$nav)
  )
}

oos_nav <- dplyr::bind_rows(nav_list_oos) |>
  dplyr::arrange(date) |>
  dplyr::mutate(cum_ret = nav / dplyr::first(nav) - 1)

# ── Long-only OOS ─────────────────────────────────────────────────────────────
signal_lo_oos <- signal_at_rebal_oos |>
  dplyr::mutate(position = dplyr::if_else(position == -1L, 0L, position))

nav_list_oos_lo <- vector("list", n_rebal_oos - 1L)
for (i in seq_len(n_rebal_oos - 1L)) {
  nav_list_oos_lo[[i]] <- rebalance_portfolio(
    rebal_date      = rebal_dates_oos[i],
    next_rebal_date = rebal_dates_oos[i + 1L],
    signals_tbl     = signal_lo_oos,
    price_data      = test_data |> dplyr::filter(symbol %in% SECTORS),
    prev_nav        = if (i == 1L) 1
                      else dplyr::last(nav_list_oos_lo[[i - 1L]]$nav)
  )
}

oos_nav_lo <- dplyr::bind_rows(nav_list_oos_lo) |>
  dplyr::arrange(date) |>
  dplyr::mutate(cum_ret = nav / dplyr::first(nav) - 1)

# ── SPY OOS benchmark ─────────────────────────────────────────────────────────
spy_nav_oos <- test_data |>
  dplyr::filter(symbol == "SPY",
                date >= min(oos_nav$date), date <= max(oos_nav$date)) |>
  dplyr::arrange(date) |>
  dplyr::mutate(nav = price / dplyr::first(price), cum_ret = nav - 1)
```

## 9.1 Performance Metrics

```{r oos-metrics}
#| echo: false
# ── OOS metrics using same compute_metrics() helper from Section 8 ─────────────
oos_metrics <- dplyr::bind_rows(
  compute_metrics(oos_nav,    "Long-Short (OOS)", spy_nav_oos),
  compute_metrics(oos_nav_lo, "Long-Only (OOS)",  spy_nav_oos),
  compute_metrics(spy_nav_oos,"SPY (OOS)",         spy_nav_oos)
)

# ── Side-by-side training vs OOS ─────────────────────────────────────────────
train_metrics_compare <- dplyr::bind_rows(
  compute_metrics(bt_nav,    "Long-Short", spy_nav),
  compute_metrics(spy_nav,   "SPY",        spy_nav)
) |>
  dplyr::rename_with(~ paste0(.x, " (Train)"), -Strategy)

oos_metrics_compare <- dplyr::bind_rows(
  compute_metrics(oos_nav,    "Long-Short", spy_nav_oos),
  compute_metrics(spy_nav_oos,"SPY",        spy_nav_oos)
) |>
  dplyr::rename_with(~ paste0(.x, " (OOS)"), -Strategy)

dplyr::left_join(train_metrics_compare, oos_metrics_compare, by = "Strategy") |>
  gt::gt() |>
  gt::tab_header(
    title    = "Training vs Out-of-Sample Performance Comparison",
    subtitle = "Train: Jun 2000–Dec 2024  |  OOS: Jan 2025–Mar 2026"
  ) |>
  gt::fmt_percent(
    columns = tidyselect::contains(c("Return", "Vol", "Drawdown", "Win")),
    decimals = 2
  ) |>
  gt::fmt_number(
    columns = tidyselect::contains(c("Sharpe", "Info")),
    decimals = 3
  ) |>
  gt::tab_style(
    style     = gt::cell_fill(color = "#dcfce7"),
    locations = gt::cells_body(rows = Strategy == "Long-Short")
  ) |>
  gt::tab_spanner(label = "Training Period",
                  columns = tidyselect::ends_with("(Train)")) |>
  gt::tab_spanner(label = "Out-of-Sample Period",
                  columns = tidyselect::ends_with("(OOS)")) |>
  gt::sub_missing(missing_text = "—") |>
  gt::opt_table_font(font = gt::google_font("IBM Plex Sans")) |>
  gt::opt_stylize(style = 1, color = "blue")
```

## 9.2 OOS Cumulative Returns

```{r oos-cumret, fig.alt="Interactive OOS cumulative return chart vs SPY"}
#| echo: false
oos_long <- dplyr::bind_rows(
  oos_nav    |> dplyr::transmute(date, cum_ret, series = "Long-Short (OOS)"),
  oos_nav_lo |> dplyr::transmute(date, cum_ret, series = "Long-Only (OOS)"),
  spy_nav_oos|> dplyr::transmute(date, cum_ret, series = "SPY (OOS)")
)

p_oos_cum <- ggplot2::ggplot(
  oos_long,
  ggplot2::aes(x = date, y = cum_ret, colour = series)
) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::scale_colour_manual(
    values = c(`Long-Short (OOS)` = "#2563EB",
               `Long-Only (OOS)`  = "#16a34a",
               `SPY (OOS)`        = "#6b7280")
  ) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  ggplot2::labs(
    title   = "OOS Cumulative Returns — Jan 2025 to Mar 2026",
    x = NULL, y = "Cumulative Return", colour = NULL
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold"),
                 panel.grid.minor = ggplot2::element_blank(),
                 legend.position  = "top")

plotly::ggplotly(p_oos_cum, tooltip = c("colour", "x", "y")) |>
  plotly::layout(hovermode = "x unified",
                 legend = list(orientation = "h", y = 1.1))
```

## 9.3 Training vs OOS Side-by-Side

```{r oos-vs-train, fig.height=5, fig.alt="Patchwork of training and OOS cumulative return charts side by side"}
#| echo: false
# ── Normalise training returns for visual comparability ───────────────────────
train_long <- dplyr::bind_rows(
  bt_nav  |> dplyr::transmute(date, cum_ret, series = "Long-Short"),
  spy_nav |> dplyr::transmute(date, cum_ret, series = "SPY")
)

make_cumret_panel <- function(df, title_str) {
  ggplot2::ggplot(df, ggplot2::aes(x = date, y = cum_ret, colour = series)) +
    ggplot2::geom_line(linewidth = 0.6) +
    ggplot2::scale_colour_manual(
      values = c(`Long-Short` = "#2563EB", SPY = "#6b7280",
                 `Long-Short (OOS)` = "#2563EB", `SPY (OOS)` = "#6b7280")
    ) +
    ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    ggplot2::scale_x_date(date_labels = "%Y") +
    ggplot2::labs(title = title_str, x = NULL, y = "Cumulative Return",
                  colour = NULL) +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold"),
                   panel.grid.minor = ggplot2::element_blank(),
                   legend.position  = "bottom")
}

p_train_panel <- make_cumret_panel(train_long,
                                   "Training (Jun 2000 – Dec 2024)")
p_oos_panel   <- make_cumret_panel(
  oos_long |>
    dplyr::filter(series %in% c("Long-Short (OOS)", "SPY (OOS)")) |>
    dplyr::mutate(series = stringr::str_remove(series, " \\(OOS\\)")),
  "Out-of-Sample (Jan 2025 – Mar 2026)"
)

p_train_panel | p_oos_panel
```



```{r oos-heatmap, fig.height=3, fig.alt="OOS monthly returns heatmap"}

#NOT RENDERING PROPERLY JUST CUT OUT
#| include: false
oos_monthly <- oos_nav |>
  dplyr::mutate(
    year  = lubridate::year(date),
    month = lubridate::month(date, label = TRUE, abbr = TRUE)
  ) |>
  dplyr::group_by(year, month) |>
  dplyr::summarise(m_ret = dplyr::last(nav) / dplyr::first(nav) - 1,
                   .groups = "drop")

# Ensure all 12 months appear on x-axis for consistent layout
all_months <- factor(month.abb, levels = month.abb)

ggplot2::ggplot(oos_monthly,
                ggplot2::aes(x = month, y = factor(year), fill = m_ret)) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.5) +
  ggplot2::geom_text(
    ggplot2::aes(
      label  = scales::percent(m_ret, accuracy = 0.1),
      colour = abs(m_ret) > 0.04
    ),
    size = 3.5, fontface = "bold", show.legend = FALSE
  ) +
  ggplot2::scale_fill_gradient2(
    low = "#dc2626", mid = "white", high = "#16a34a", midpoint = 0,
    labels = scales::percent_format(accuracy = 1)
  ) +
  ggplot2::scale_colour_manual(values = c(`TRUE` = "white",
                                          `FALSE` = "grey20")) +
  ggplot2::scale_x_discrete(limits = month.abb, drop = FALSE) +
  ggplot2::scale_y_discrete(limits = rev) +
  ggplot2::labs(
    title = "Monthly Returns Heatmap — OOS Period (Jan 2025 – Mar 2026)",
    x = NULL, y = NULL, fill = "Monthly\nReturn",
    caption = "Empty cells = months not yet in OOS window"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title    = ggplot2::element_text(face = "bold"),
                 panel.grid    = ggplot2::element_blank())
```

## 9.5 Honest Assessment

Fifteen months of OOS data limits statistical power. Four structural factors deserve honest examination.

**Cluster stability.** Frozen centroids will mis-classify if sector dispersion compressed or inverted in 2025. Any month with average silhouette width below 0.25 is a low-confidence assignment.

**Rebalancing lag.** The strategy is structurally slow — entries are monthly, and the innovation z-score requires 63 days of history. In fast-moving markets this lag is costly.

**Short book value.** The long-short vs long-only comparison directly tests whether shorting Defensive sectors added value. If XLU, XLP, XLRE outperformed in a risk-off environment, the short book detracted.

**Conditions for the strategy to work OOS.** Three things must hold: (1) sufficient sector dispersion for meaningful clusters, (2) the Kalman signal leads rather than lags price, (3) Defensive ETFs genuinely underperform during Growth regimes. All three are empirical — not maintained by assumption.

---

# 10. Circle Back {#sec-circle-back}

What a good strategy knows about itself matters more than its best backtest.

## 10.1 What the Strategy Learned

The strategy earns its justification in some market conditions and not others — and understanding that boundary is itself the most valuable output.

The three-cluster structure was not imposed; it emerged from the data. K-means consistently separated high-return cyclicals from low-return defensives with a neutral group in between, and this structure held across both pre- and post-COVID sub-periods (though centroids drifted). Sectors do rotate across regimes, and that rotation is partially predictable from lagged behaviour.

The Kalman filter's innovation sequence carried information beyond smoothing. Innovation z-score spikes preceded or coincided with cluster flips in the training period, capturing regime transitions earlier than monthly re-assignment could. Using the filter's internal mechanics — not just its output — as a risk signal is this strategy's most technically differentiated feature.

## 10.2 When Clustering Fails

K-means on sector ETFs is fragile under three identifiable conditions.

**Extreme volatility (VIX > 40).** Cross-sector correlations converge toward 1.0 — every sector falls simultaneously, the feature space collapses, and silhouette widths drop below 0.25. Cluster assignments become essentially random.

**Structural breaks.** The AI revolution altered XLK's return dynamics and its correlations with every other sector. If technology becomes a macro factor rather than a sector, frozen centroids are anchored to a world that no longer exists.

**Regime speed.** A transition that unfolds in days (COVID: 23 trading days peak-to-trough) generates a cluster flip signal only after the damage is done. The innovation exit provides daily protection but requires 63 days of history — the first two months of any new regime are partially blind.

## 10.3 Limitations

**Transaction costs** are absent. At ~5 bps per one-way trade across monthly rebalancing of eleven ETFs over 24 years, the cumulative drag is not negligible.

**Short selling is treated as costless.** Borrow costs on Defensive ETFs rise precisely when the strategy wants to be short them. The information ratio should be read as an upper bound.

**Single-model concentration.** Different distance metrics, seeds, or $k$ values would produce different regime assignments. The strategy commits to one solution without quantifying model uncertainty — a more robust version would average across clustering specifications.

**Look-ahead audit.** Kalman $W$/$V$, feature scaling, and centroids are all frozen from training — correct. The one remaining risk: `theta_base` assumes delta distributions are stationary across periods. If test-period signals are substantially more volatile, the threshold is too low and generates excess false entries.

## 10.4 Inversion — Five Ways This Strategy Fails

::: {.callout-note icon=false}
## Poker Analogy
The inversion exercise is studying your losing sessions. The best players study when their edge broke down, not just hands they won. Instead of asking how to win more, ask what conditions would guarantee losses.
:::

What would have to be true for this strategy to reliably destroy capital?

1. **Persistent low dispersion.** All eleven sectors move together — cluster assignments become random and transaction costs erode capital.
2. **Inverted return relationship.** Defensive sectors outperform Growth for an extended period (partially occurred in 2022). The strategy loses on both books if labels don't update fast enough.
3. **Kalman structural break.** If HFT permanently alters the autocorrelation of RSI/MACD, innovations become persistently non-zero, triggering constant false exits and churning the portfolio.
4. **Crowding.** If enough quant managers trade the same eleven ETFs with similar signals, the edge is arbitraged away.
5. **Short-selling restrictions.** A borrow ban on Defensive ETFs — as imposed on financials in 2008 — removes the hedge and leaves a naked long book.

## 10.5 What Would Change in a Rebuild

A few changes would improve robustness without sacrificing the core thesis.

**Regime uncertainty via GMM.** Replace hard k-means assignments with Gaussian Mixture Model membership probabilities. Scale position sizes by cluster confidence — 0.9 Growth probability warrants larger exposure than 0.6. Averaging across multiple clustering approaches (k-means, GMM, hierarchical) reduces signal variance.

**Mixed-frequency signals.** Retain monthly cluster assignment as the regime prior but evaluate delta entry/exit at weekly frequency. This reduces the rebalancing lag identified in Section 9 without abandoning the low-turnover character.

**Regime-quality filter.** Suppress all trading when silhouette width falls below 0.25 — go to cash regardless of cluster label. Knowing when the model is inapplicable is more valuable than any specific model improvement.

**Making the clustering explanation part better.** I found a lot of issues when it came to making the clustering seciton especially the use of gganimate as I had long render times and then when I swapped to plotly it still was not producing what I wanted

**Account for transaction costs at the start** I made the mistake of not accounting for transaction costs and margin calls, when shorting it could be quite vital and I think I definenitly missed this part which could really improve the accuracy of this model.


