---
title: "Clustering-Based Sector Rotation Strategy"
subtitle: "Active Fund Management via K-Means Regime Detection and Kalman Signal Filtering"
author: "FIN 451 — Financial Analytics and Trading"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    theme: flatly
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 6
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
# ── Package loading ──────────────────────────────────────────────────────────
# Packages used via explicit namespace notation throughout (pkg::fn()).
# require() is used here so a single missing package does not abort the render.

pkgs <- c(
  "tidyverse", "tidyquant", "tidymodels",
  "ggplot2", "plotly", "patchwork", "gganimate",
  "cluster", "factoextra", "moments",
  "dlm", "KFAS",
  "gt", "scales", "lubridate", "rlang", "purrr", "zoo", "TTR"
)

invisible(lapply(pkgs, function(p) {
  if (!requireNamespace(p, quietly = TRUE)) {
    message("Package not installed: ", p,
            " — install with install.packages('", p, "')")
  } else {
    suppressPackageStartupMessages(require(p, character.only = TRUE))
  }
}))
```

---

# 1. Problem Framing {#sec-framing}

## The Question Every Active Manager Must Answer

The S&P 500 index returned approximately 10.5% annualised over the past century. An investor who bought SPY in June 2000 and held it through December 2024 would have compounded their capital through the dot-com collapse, the Global Financial Crisis, a pandemic, and two bear markets — and still come out ahead of the majority of actively managed funds. The passive case is not a straw man. It is the empirical reality that active management has to beat, after fees, consistently, and on a risk-adjusted basis.

This document addresses a single question: is there a systematic, rules-based active strategy applied to the eleven SPDR sector ETFs that justifies its cost over simply holding SPY? The answer this document pursues is not "yes by construction" — it is "yes under specific, identifiable market conditions, and no when those conditions break down." That distinction matters. Overclaiming is the most common failure mode in quantitative finance, and the most damaging to an investor who acts on it.

## Markets as Ecosystems, Not Machines

The central intuition behind this strategy is ecological rather than mechanical. A machine has a fixed input-output relationship — the same lever always produces the same result. An ecosystem is different: it has regimes. A forest in spring behaves differently from the same forest in drought. The species relationships, the energy flows, the feedback loops all shift. Attempting to apply the optimal spring strategy during a drought destroys value.

Financial markets share this character. The relationship between sector returns is not fixed. During expansions, cyclical sectors — technology, consumer discretionary, industrials — attract capital as investors reach for growth. During contractions, defensive sectors — utilities, consumer staples, real estate — hold value as investors seek yield and stability. During genuine crises, correlations collapse toward one and the distinction between sectors temporarily disappears. These are not random fluctuations around a stable mean; they are qualitatively different states of the system, each with its own internal logic.

A strategy that treats all market environments identically will be right on average and disastrously wrong in the tails. A strategy that can identify which regime is currently active, and adjust accordingly, has a structural advantage — provided the regime identification is reliable, timely, and not already priced in by the time the strategy acts.

## The Investment Mandate

This strategy operates as a fully active fund manager with three states:

**Growth regime:** The strategy takes long positions in sector ETFs assigned to the Growth cluster by the k-means regime detection model. These ETFs are weighted by inverse volatility — lower-volatility sectors receive proportionally larger allocations, stabilising the portfolio's risk contribution across holdings.

**Defensive regime:** The strategy takes short positions in sector ETFs assigned to the Defensive cluster. Short positions are expressed as negative inverse-volatility weights. The thesis is that Defensive sectors underperform when the Growth regime is active, and the strategy should capture this underperformance rather than simply avoiding it.

**Neutral regime:** The strategy holds cash. No sector ETFs are held long or short. This is not a failure mode — it is a deliberate design choice that acknowledges the model's uncertainty during regime transitions.

The strategy rebalances at the end of every calendar month. Intra-month exits are triggered only by a statistical signal from the Kalman filter, described in Section 4. No discretionary overrides exist; every position is determined by a rule that could be specified before the fact and verified after it.

## Why This Combination of Tools

Two components drive the strategy's potential edge, and they operate at different frequencies.

The k-means clustering model identifies the **regime** — the coarse-grained state of the market — at monthly frequency. It does this by grouping the eleven sector ETFs into clusters based on their recent return, volatility, Sharpe ratio, and momentum characteristics. The cluster structure is not imposed; it emerges from the data and is validated by silhouette analysis and sub-period stability tests. The regime label tells the strategy which direction to trade.

The Kalman filter operates at **daily frequency** and solves a different problem: distinguishing genuine signal from noise in technical indicators. Raw RSI, MACD, rate-of-change, and Bollinger Band readings are contaminated by microstructure noise that carries no predictive information. The Kalman filter estimates the latent true state of each indicator, and its first difference — the delta — provides a momentum signal that is less prone to false triggers than the raw indicator. Critically, the filter also produces a state covariance $P_t$ that measures its own uncertainty, which is used to scale the entry threshold adaptively. When the filter is uncertain, it requires stronger evidence to enter. When it is confident, it allows entry on smaller movements.

Together, the two components address a core challenge in systematic trading: the regime tells you what to do, and the filter tells you when the evidence is strong enough to do it.

## What This Document Demonstrates

The document is organised to build the case for this strategy incrementally, section by section, before delivering an honest verdict in Section 10.

Section 2 presents the full decision logic as a visual flowchart and written rules before any code is shown — a commitment to explaining the strategy in plain terms before implementing it. Section 3 pulls all data via API. Sections 4 and 5 construct the features and the clustering model, with explicit validation of cluster quality and stability. Section 6 assembles the entry and exit signals. Section 7 states every trading rule formally in mathematical notation. Sections 8 and 9 run the backtest on the training period and the out-of-sample test period respectively, reporting the same set of performance metrics for both. Section 10 returns to this opening question — honestly, including the failure modes and the conditions under which the strategy should not be used.

The benchmark throughout is SPY. It is never traded, never assigned a cluster label, and never gamed. It is the standard against which every performance claim is measured.

---

# 2. Mental Model {#sec-mental-model}

The diagram below shows the complete decision pipeline from raw price data to final portfolio position. Every arrow represents a conditional rule; every node represents a transformation or decision point. The symmetric long/short structure is visible in the centre branch.

```{mermaid}
%%| fig-width: 9
flowchart TD
    A([Daily Price Data\nSPY + 11 Sector ETFs]) --> B[Feature Engineering\nRSI · MACD · ROC · Bollinger %B\nRolling Return · Vol · Sharpe]

    B --> C[Kalman Filter\nper ETF × per Indicator\nEstimate W and V on training only]

    C --> D[Filtered State x̂_t\nState Covariance P_t\nInnovation v_t]

    D --> E[Delta Signal\nδ_t = x̂_t − x̂_t-1]
    D --> F[Innovation Z-Score\nz_t = rolling z-score of v_t]

    E --> G[Monthly K-Means Clustering\nk = 3  ·  tidymodels workflow\nFeatures scaled on training only]

    G --> H{Regime Label\nat rebal date τ}

    H -->|Growth| I[Entry Check\nδ_t > +θ_adaptive ?]
    H -->|Neutral| J[FLAT\nHold Cash]
    H -->|Defensive| K[Entry Check\nδ_t < −θ_adaptive ?]

    I -->|Yes| L[LONG\nInverse Vol Weighted\n+w_i]
    I -->|No| J
    K -->|Yes| M[SHORT\nInverse Vol Weighted\n−w_i]
    K -->|No| J

    L --> N{Exit Check\nDaily}
    M --> N

    F --> N
    N -->|z_t > 2 | O[EXIT\nInnovation Spike\nRegime Break Signal]
    N -->|Cluster flips\nat month end| O
    N -->|No exit| P[Hold Position\nuntil next rebal]

    O --> Q([Portfolio NAV\nMonthly Rebalance\nlubridate::rollback])
    P --> Q
    J --> Q

    Q --> R[Performance vs SPY\nSharpe · MaxDD · Info Ratio]

    style A fill:#e0f2fe,stroke:#0284c7
    style J fill:#fef9c3,stroke:#ca8a04
    style L fill:#dcfce7,stroke:#16a34a
    style M fill:#fee2e2,stroke:#dc2626
    style O fill:#fed7aa,stroke:#ea580c
    style R fill:#f3e8ff,stroke:#9333ea
    style Q fill:#e0f2fe,stroke:#0284c7
```

**Reading the diagram.** Data enters at the top. Every ETF gets its own Kalman filter — parameters fit once on training data, never touched again. At each month-end rebalancing date the clustering model assigns a regime label. The label sets the allowed direction: Growth permits long entries, Defensive permits short entries, Neutral forces cash. Within the permitted direction, the delta signal must also clear the adaptive threshold before a position is opened. Once open, the position is monitored daily for an innovation z-score spike (primary exit) and monthly for a cluster flip (secondary exit). Either trigger independently closes the position. Weights are inverse-volatility within each book — long weights sum to +1, short weights sum to −1.

---

# 3. Data {#sec-data}

## 3.1 Asset Universe & Periods

All data is sourced exclusively via API using `tidyquant::tq_get()`. No local files, CSV, or Excel imports are used anywhere in this document.

The strategy trades **11 SPDR sector ETFs**. **SPY is included as a passive benchmark only and is never traded.**

| Role | Tickers |
|------|---------|
| Benchmark (not traded) | SPY |
| Tradeable universe | XLB, XLC, XLE, XLF, XLI, XLK, XLP, XLRE, XLV, XLU, XLY |

> **Note on XLC:** The Communication Services ETF (XLC) launched in June 2018. Observations prior to that date are absent from the raw pull and are handled gracefully — affected rows are simply absent rather than filled or imputed. All downstream calculations are conditioned on data availability per ticker.

```{r data-constants}
# ── Asset universe ───────────────────────────────────────────────────────────
BENCHMARK  <- "SPY"
SECTORS    <- c("XLB", "XLC", "XLE", "XLF", "XLI",
                "XLK", "XLP", "XLRE", "XLV", "XLU", "XLY")
ALL_TICKERS <- c(BENCHMARK, SECTORS)

# ── Period boundaries ────────────────────────────────────────────────────────
TRAIN_START <- lubridate::ymd("2000-06-01")
TRAIN_END   <- lubridate::ymd("2024-12-31")
TEST_START  <- lubridate::ymd("2025-01-01")
TEST_END    <- lubridate::ymd("2026-03-31")

# ── Split boundary label (for chart annotation) ──────────────────────────────
SPLIT_DATE  <- TEST_START
```

## 3.2 Data Pull

```{r data-pull}
# ── Pull daily adjusted prices for all 12 tickers ───────────────────────────
raw_prices <- tidyquant::tq_get(
  ALL_TICKERS,
  get  = "stock.prices",
  from = TRAIN_START,
  to   = TEST_END
) |>
  dplyr::select(symbol, date, adjusted) |>
  dplyr::rename(price = adjusted) |>
  dplyr::arrange(symbol, date)

# ── Sanity check: confirm retrieval ─────────────────────────────────────────
raw_prices |>
  dplyr::group_by(symbol) |>
  dplyr::summarise(
    first_date = min(date),
    last_date  = max(date),
    n_obs      = dplyr::n(),
    .groups    = "drop"
  ) |>
  dplyr::arrange(symbol)
```

## 3.3 Train / Test Split

```{r data-split}
# ── Training window: June 2000 – December 2024 ───────────────────────────────
train_data <- raw_prices |>
  dplyr::filter(date >= TRAIN_START, date <= TRAIN_END)

# ── Test window: January 2025 – March 2026 ───────────────────────────────────
test_data <- raw_prices |>
  dplyr::filter(date >= TEST_START, date <= TEST_END)
```

## 3.4 Indexed Price Chart

Prices are indexed to **100 at each ticker's first available trading date** so that all series are visually comparable regardless of absolute price level. The dashed vertical line marks the boundary between the training and test periods.

```{r data-index}
# ── Index each ticker to 100 at its own first available date ─────────────────
prices_indexed <- raw_prices |>
  dplyr::group_by(symbol) |>
  dplyr::mutate(
    base_price    = dplyr::first(price),
    indexed_price = price / base_price * 100
  ) |>
  dplyr::ungroup() |>
  dplyr::select(symbol, date, indexed_price)
```

```{r chart-indexed, fig.alt="Interactive chart of all 12 ETFs indexed to 100 at first available date"}
# ── Base ggplot ───────────────────────────────────────────────────────────────
p_indexed <- ggplot2::ggplot(
  prices_indexed,
  ggplot2::aes(x = date, y = indexed_price, colour = symbol, group = symbol)
) +
  ggplot2::geom_line(linewidth = 0.45, alpha = 0.85) +
  ggplot2::geom_vline(
    xintercept = as.numeric(SPLIT_DATE),
    linetype   = "dashed",
    colour     = "grey30",
    linewidth  = 0.6
  ) +
  ggplot2::annotate(
    "text",
    x      = SPLIT_DATE + lubridate::days(60),
    y      = max(prices_indexed$indexed_price, na.rm = TRUE) * 0.97,
    label  = "← Train  |  Test →",
    size   = 3,
    hjust  = 0,
    colour = "grey30"
  ) +
  ggplot2::scale_colour_manual(
    values = scales::hue_pal()(length(ALL_TICKERS))
  ) +
  ggplot2::scale_x_date(
    date_breaks = "2 years",
    date_labels = "%Y"
  ) +
  ggplot2::scale_y_continuous(
    labels = scales::comma_format(suffix = "")
  ) +
  ggplot2::labs(
    title    = "SPDR Sector ETFs — Indexed to 100 at First Available Date",
    subtitle = "Daily adjusted closing prices  |  Dashed line = Train/Test split (Jan 2025)",
    x        = NULL,
    y        = "Indexed Price (base = 100)",
    colour   = "Ticker",
    caption  = "Source: Yahoo Finance via tidyquant::tq_get()"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    legend.position  = "right",
    panel.grid.minor = ggplot2::element_blank(),
    plot.title       = ggplot2::element_text(face = "bold")
  )

# ── Wrap in plotly for interactivity ─────────────────────────────────────────
plotly::ggplotly(p_indexed, tooltip = c("colour", "x", "y")) |>
  plotly::layout(
    hovermode = "x unified",
    legend    = list(title = list(text = "<b>Ticker</b>"))
  )
```

## 3.5 Summary Statistics

```{r summary-stats}
# ── Compute per-ETF summary statistics on the full raw series ────────────────
etf_summary <- raw_prices |>
  dplyr::group_by(symbol) |>
  dplyr::mutate(
    daily_return = (price / dplyr::lag(price)) - 1
  ) |>
  dplyr::summarise(
    `First Date`      = min(date),
    `Last Date`       = max(date),
    `Observations`    = dplyr::n(),
    `Mean Dly Ret`    = mean(daily_return, na.rm = TRUE),
    `Ann. Vol`        = sd(daily_return, na.rm = TRUE) * sqrt(252),
    `Min Price`       = min(price, na.rm = TRUE),
    `Max Price`       = max(price, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    `Role` = dplyr::if_else(symbol == "SPY", "Benchmark", "Tradeable")
  ) |>
  dplyr::arrange(dplyr::desc(`Role`), symbol)

# ── Render as formatted gt table ─────────────────────────────────────────────
etf_summary |>
  gt::gt() |>
  gt::tab_header(
    title    = "Asset Universe — Summary Statistics",
    subtitle = "Full sample: June 2000 – March 2026 | Adjusted closing prices"
  ) |>
  gt::fmt_date(columns = c(`First Date`, `Last Date`), date_style = "yMd") |>
  gt::fmt_number(columns = `Observations`, use_seps = TRUE, decimals = 0) |>
  gt::fmt_percent(columns = c(`Mean Dly Ret`, `Ann. Vol`), decimals = 2) |>
  gt::fmt_currency(columns = c(`Min Price`, `Max Price`), currency = "USD",
                   decimals = 2) |>
  gt::tab_style(
    style     = gt::cell_fill(color = "#f0f4ff"),
    locations = gt::cells_body(rows = `Role` == "Benchmark")
  ) |>
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::cols_label(symbol = "Ticker") |>
  gt::tab_footnote(
    footnote  = "XLC launched June 2018; prior observations are absent, not imputed.",
    locations = gt::cells_body(columns = `First Date`,
                               rows    = symbol == "XLC")
  ) |>
  gt::tab_footnote(
    footnote  = "SPY is a benchmark only and is excluded from all portfolio positions.",
    locations = gt::cells_body(columns = `Role`,
                               rows    = symbol == "SPY")
  ) |>
  gt::opt_table_font(font = gt::google_font("IBM Plex Sans")) |>
  gt::opt_stylize(style = 1, color = "blue")
```

---

# 4. Feature Engineering {#sec-features}

## 4.1 Technical Indicators

Four technical indicators are computed daily for each of the 11 tradeable sector ETFs using the `TTR` package. These are the raw observations that will subsequently be fed into the Kalman filter. SPY is excluded from feature engineering — it is never traded.

| Indicator | Formula / Window | Economic Intuition |
|-----------|------------------|--------------------|
| **RSI** | 14-day Wilder RSI | Momentum oscillator bounded [0, 100]; values above 70 indicate overbought conditions, below 30 oversold |
| **MACD** | EMA(12) − EMA(26), signal EMA(9) | Captures trend acceleration; the MACD line crossing its signal line signals directional shifts |
| **ROC** | $(P_t - P_{t-10}) / P_{t-10}$ | 10-day rate of change; measures near-term price momentum |
| **Bollinger %B** | $(P_t - \text{Lower}) / (\text{Upper} - \text{Lower})$ | Position of price within 20-day, 2-SD bands; >1 or <0 flags breakouts |

```{r compute-indicators}
# ── Compute all four indicators for every tradeable ETF ──────────────────────
# Input: train_data and test_data from Section 3 (adjusted daily prices)
# Output: raw_indicators — long-format tibble, one row per ETF × date × indicator

compute_indicators <- function(df) {
  prices <- df$price

  rsi_vals  <- TTR::RSI(prices, n = 14)

  macd_obj  <- TTR::MACD(prices, nFast = 12, nSlow = 26, nSig = 9)
  macd_vals <- macd_obj[, "macd"] - macd_obj[, "signal"]

  roc_vals  <- TTR::ROC(prices, n = 10, type = "discrete")

  bb_obj    <- TTR::BBands(prices, n = 20, sd = 2)
  pctb_vals <- bb_obj[, "pctB"]

  tibble::tibble(
    date      = df$date,
    RSI       = as.numeric(rsi_vals),
    MACD      = as.numeric(macd_vals),
    ROC       = as.numeric(roc_vals),
    BollingerB = as.numeric(pctb_vals)
  )
}

# Apply to training data for each sector ETF (not SPY)
raw_indicators_train <- train_data |>
  dplyr::filter(symbol %in% SECTORS) |>
  dplyr::group_by(symbol) |>
  dplyr::group_modify(~ compute_indicators(.x)) |>
  dplyr::ungroup() |>
  tidyr::pivot_longer(
    cols      = c(RSI, MACD, ROC, BollingerB),
    names_to  = "indicator",
    values_to = "raw_value"
  ) |>
  dplyr::filter(!is.na(raw_value))

# Apply to test data
raw_indicators_test <- test_data |>
  dplyr::filter(symbol %in% SECTORS) |>
  dplyr::group_by(symbol) |>
  dplyr::group_modify(~ compute_indicators(.x)) |>
  dplyr::ungroup() |>
  tidyr::pivot_longer(
    cols      = c(RSI, MACD, ROC, BollingerB),
    names_to  = "indicator",
    values_to = "raw_value"
  ) |>
  dplyr::filter(!is.na(raw_value))
```

## 4.2 Kalman Filter

### 4.2.1 State-Space Model

Raw technical indicators are noisy — they react to microstructure noise, bid-ask bounce, and short-lived price dislocations that carry no regime information. A Kalman filter estimates the latent *true* state of each indicator, stripping out observation noise while tracking genuine directional changes.

The model is a random-walk state space applied separately to each (ETF, indicator) pair:

$$x_t = x_{t-1} + w_t, \quad w_t \sim \mathcal{N}(0,\, W)$$

$$y_t = x_t + v_t, \quad v_t \sim \mathcal{N}(0,\, V)$$

where $x_t$ is the unobserved latent state, $y_t$ is the observed indicator value, $W$ is the process noise variance (how much the true signal drifts), and $V$ is the observation noise variance (measurement error). A higher $W/V$ ratio makes the filter more responsive; a lower ratio gives heavier smoothing.

**Critical constraint:** $W$ and $V$ are estimated via maximum likelihood on the **training period only**. The fitted parameters are then frozen and applied forward to the test period. This enforces strict temporal separation and eliminates look-ahead bias.

### 4.2.2 Parameter Estimation & Filtering

```{r kalman-fit, warning=FALSE}
# ── Fit Kalman parameters (W, V) via MLE on training data only ───────────────
# Returns a list with: filtered_state, P_t (state variance), innovation

fit_and_filter_kalman <- function(y_obs) {
  # Remove leading NAs from indicator warm-up period
  valid_idx <- which(!is.na(y_obs))
  if (length(valid_idx) < 30L) {
    return(tibble::tibble(
      filtered_state = rep(NA_real_, length(y_obs)),
      P_t            = rep(NA_real_, length(y_obs)),
      innovation     = rep(NA_real_, length(y_obs))
    ))
  }
  y_clean <- y_obs[valid_idx]

  # Build random-walk DLM: FF=1 (observation), GG=1 (state transition),
  # V (obs noise), W (process noise) — both estimated via MLE
  build_fn <- function(pars) {
    dlm::dlm(
      FF = matrix(1),
      GG = matrix(1),
      V  = exp(pars[1]),   # log-transform keeps variance positive
      W  = exp(pars[2]),
      m0 = matrix(y_clean[1]),
      C0 = matrix(100)     # diffuse prior
    )
  }

  # MLE fit on training observations
  mle_fit <- dlm::dlmMLE(
    y     = y_clean,
    parm  = c(0, 0),       # initialise log(V), log(W) at 0
    build = build_fn,
    control = list(maxit = 500)
  )

  fitted_model <- build_fn(mle_fit$par)

  # Forward Kalman filter pass
  kf <- dlm::dlmFilter(y_clean, fitted_model)

  # Extract outputs — drop t=0 prior state to align with observations
  f_state <- as.numeric(kf$m[-1])               # filtered state x̂_t
  p_vals  <- as.numeric(dlm::dlmSvd2var(kf$U.C, kf$D.C)[-1])  # P_t
  innov   <- as.numeric(kf$y - kf$f)            # innovation v_t = y_t - ŷ_t

  # Re-insert NAs for warm-up rows so length matches original y_obs
  out <- tibble::tibble(
    filtered_state = rep(NA_real_, length(y_obs)),
    P_t            = rep(NA_real_, length(y_obs)),
    innovation     = rep(NA_real_, length(y_obs))
  )
  out$filtered_state[valid_idx] <- f_state
  out$P_t[valid_idx]            <- p_vals
  out$innovation[valid_idx]     <- innov
  out
}

# ── Apply to every (symbol, indicator) combination in training data ───────────
# Each pair gets its own independently estimated W and V
kalman_params_train <- raw_indicators_train |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::group_modify(function(.x, .y) {
    kf_out <- fit_and_filter_kalman(.x$raw_value)
    dplyr::bind_cols(.x, kf_out)
  }) |>
  dplyr::ungroup()
```

### 4.2.3 Applying Frozen Parameters to the Test Period

The MLE-estimated model parameters are derived entirely from training data. To apply them to the test period without refitting, we re-run the filter forward using the final training state as the initial condition.

```{r kalman-apply-test, warning=FALSE}
# ── Helper: apply a pre-fit DLM to new observations ──────────────────────────
apply_frozen_kalman <- function(train_y, test_y) {
  valid_train <- which(!is.na(train_y))
  if (length(valid_train) < 30L) {
    return(tibble::tibble(
      filtered_state = rep(NA_real_, length(test_y)),
      P_t            = rep(NA_real_, length(test_y)),
      innovation     = rep(NA_real_, length(test_y))
    ))
  }
  y_train_clean <- train_y[valid_train]

  build_fn <- function(pars) {
    dlm::dlm(FF = matrix(1), GG = matrix(1),
              V = exp(pars[1]), W = exp(pars[2]),
              m0 = matrix(y_train_clean[1]), C0 = matrix(100))
  }

  mle_fit     <- dlm::dlmMLE(y_train_clean, c(0, 0), build_fn,
                              control = list(maxit = 500))
  fitted_mod  <- build_fn(mle_fit$par)

  # Filter training data to get terminal state for initialisation
  kf_train    <- dlm::dlmFilter(y_train_clean, fitted_mod)
  n_tr        <- length(y_train_clean)
  m_terminal  <- as.numeric(kf_train$m[n_tr + 1])
  C_terminal  <- as.numeric(dlm::dlmSvd2var(kf_train$U.C, kf_train$D.C)[n_tr + 1])

  # Re-initialise model at terminal training state — NO refitting
  fitted_mod$m0 <- matrix(m_terminal)
  fitted_mod$C0 <- matrix(C_terminal)

  valid_test <- which(!is.na(test_y))
  out <- tibble::tibble(
    filtered_state = rep(NA_real_, length(test_y)),
    P_t            = rep(NA_real_, length(test_y)),
    innovation     = rep(NA_real_, length(test_y))
  )

  if (length(valid_test) == 0L) return(out)

  y_test_clean <- test_y[valid_test]
  kf_test      <- dlm::dlmFilter(y_test_clean, fitted_mod)

  out$filtered_state[valid_test] <- as.numeric(kf_test$m[-1])
  out$P_t[valid_test]            <- as.numeric(
    dlm::dlmSvd2var(kf_test$U.C, kf_test$D.C)[-1])
  out$innovation[valid_test]     <- as.numeric(kf_test$y - kf_test$f)
  out
}

# ── Apply frozen parameters to test data ────────────────────────────────────
kalman_params_test <- raw_indicators_test |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::group_modify(function(.x, .y) {
    # Retrieve matching training series for this (symbol, indicator) pair
    train_series <- raw_indicators_train |>
      dplyr::filter(symbol == .y$symbol, indicator == .y$indicator) |>
      dplyr::pull(raw_value)

    kf_out <- apply_frozen_kalman(train_series, .x$raw_value)
    dplyr::bind_cols(.x, kf_out)
  }) |>
  dplyr::ungroup()

# ── Combine training and test Kalman outputs into a single tidy tibble ────────
kalman_all <- dplyr::bind_rows(
  dplyr::mutate(kalman_params_train, period = "train"),
  dplyr::mutate(kalman_params_test,  period = "test")
) |>
  dplyr::select(symbol, date, indicator, period,
                raw_value, filtered_state, P_t, innovation) |>
  dplyr::arrange(symbol, indicator, date)
```

## 4.3 Delta Signal & Adaptive Threshold

### 4.3.1 Delta: First Difference of Filtered State

The filtered state $\hat{x}_t$ tells us the Kalman filter's best estimate of the indicator's current level. The **delta** is its first difference:

$$\delta_t = \hat{x}_t - \hat{x}_{t-1}$$

Delta measures the *rate of change* of the smoothed signal — analogous to velocity in a physical system. A positive $\delta_t$ means the filtered signal is accelerating upward; negative means it is decelerating or reversing. This is the entry trigger quantity: we want to enter positions when the smoothed signal has genuine directional momentum, not just a level reading.

### 4.3.2 Adaptive Threshold Scaled by State Covariance

The Kalman filter outputs the state covariance $P_t$ alongside the filtered state. $P_t$ measures the filter's own uncertainty about its state estimate — it rises when the filter is being surprised by observations (regime transition) and falls when the model fits well (stable regime).

The adaptive entry threshold uses $P_t$ directly:

$$\theta_{\text{adaptive},t} = \theta_{\text{base}} \cdot \sqrt{P_t}$$

where $\theta_{\text{base}}$ is estimated as the median absolute delta over the **training period only**. The intuition:

- **High $P_t$ (high uncertainty):** the filter is confused — require a larger $|\delta_t|$ to enter, avoiding false signals during regime transitions
- **Low $P_t$ (high certainty):** the filter is confident — allow smaller $|\delta_t|$ to trigger entries, capturing early momentum

```{r compute-delta}
# ── Compute delta and adaptive threshold from Kalman outputs ─────────────────
kalman_all <- kalman_all |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::arrange(date, .by_group = TRUE) |>
  dplyr::mutate(
    delta = filtered_state - dplyr::lag(filtered_state)
  ) |>
  dplyr::ungroup()

# ── Estimate theta_base per (symbol, indicator) on TRAINING data only ─────────
theta_base_tbl <- kalman_all |>
  dplyr::filter(period == "train", !is.na(delta)) |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::summarise(
    theta_base = stats::median(abs(delta), na.rm = TRUE),
    .groups    = "drop"
  )

# ── Join theta_base back and compute adaptive threshold ──────────────────────
kalman_all <- kalman_all |>
  dplyr::left_join(theta_base_tbl, by = c("symbol", "indicator")) |>
  dplyr::mutate(
    theta_adaptive = theta_base * sqrt(P_t)
  )
```

## 4.4 Innovation Sequence & Rolling Z-Score

The Kalman innovation at time $t$ is defined as the one-step-ahead prediction error:

$$v_t = y_t - \hat{y}_{t|t-1}$$

Under a correctly specified model, innovations should be approximately white noise with zero mean. When the innovation z-score spikes — i.e., the filter is being repeatedly surprised — this is statistical evidence of a regime break. This forms the primary exit signal (Section 6).

```{r compute-innovation-zscore}
# ── Rolling z-score of innovation sequence (60-day window) ───────────────────
kalman_all <- kalman_all |>
  dplyr::group_by(symbol, indicator) |>
  dplyr::arrange(date, .by_group = TRUE) |>
  dplyr::mutate(
    innov_roll_mean = zoo::rollapply(
      innovation, width = 60, FUN = mean,
      fill = NA, align = "right", na.rm = TRUE
    ),
    innov_roll_sd = zoo::rollapply(
      innovation, width = 60, FUN = sd,
      fill = NA, align = "right", na.rm = TRUE
    ),
    innovation_zscore = dplyr::if_else(
      innov_roll_sd > 0,
      (innovation - innov_roll_mean) / innov_roll_sd,
      NA_real_
    )
  ) |>
  dplyr::ungroup()
```

## 4.5 Visualisation: XLK — Raw vs Filtered Signals

XLK (Technology) is used as the illustrative ETF throughout this section. Its high volatility and distinct trend cycles make it the most visually informative sector for demonstrating Kalman smoothing.

The 2×2 grid below shows all four indicators over the training period. In each panel the grey line is the raw TTR indicator; the coloured line is the Kalman filtered state $\hat{x}_t$.

```{r chart-raw-vs-filtered, fig.height=10, fig.alt="2x2 grid of raw vs Kalman filtered signals for XLK across all four indicators"}
# ── Prepare XLK training data for plotting ───────────────────────────────────
xlk_plot_data <- kalman_all |>
  dplyr::filter(symbol == "XLK", period == "train",
                !is.na(filtered_state)) |>
  dplyr::select(date, indicator, raw_value, filtered_state)

indicator_labels <- c(
  RSI        = "RSI (14-day)",
  MACD       = "MACD Line − Signal",
  ROC        = "Rate of Change (10-day)",
  BollingerB = "Bollinger Band %B"
)

# ── Build one panel per indicator ────────────────────────────────────────────
make_panel <- function(ind_name) {
  df <- xlk_plot_data |> dplyr::filter(indicator == ind_name)

  ggplot2::ggplot(df, ggplot2::aes(x = date)) +
    ggplot2::geom_line(
      ggplot2::aes(y = raw_value),
      colour = "grey70", linewidth = 0.35, alpha = 0.8
    ) +
    ggplot2::geom_line(
      ggplot2::aes(y = filtered_state),
      colour = "#2563EB", linewidth = 0.65
    ) +
    ggplot2::scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
    ggplot2::labs(
      title    = indicator_labels[ind_name],
      subtitle = "Grey = raw  |  Blue = Kalman filtered state",
      x        = NULL,
      y        = ind_name
    ) +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      plot.title       = ggplot2::element_text(face = "bold", size = 10)
    )
}

p_rsi  <- make_panel("RSI")
p_macd <- make_panel("MACD")
p_roc  <- make_panel("ROC")
p_bb   <- make_panel("BollingerB")

# ── 2x2 patchwork layout ─────────────────────────────────────────────────────
(p_rsi | p_macd) / (p_roc | p_bb) +
  patchwork::plot_annotation(
    title   = "XLK — Raw vs Kalman Filtered Technical Indicators (Training Period)",
    caption = "Training period: June 2000 – December 2024",
    theme   = ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", size = 12)
    )
  )
```

```{r chart-delta-threshold, fig.alt="XLK RSI delta signal with adaptive threshold bands"}
# ── Delta + adaptive threshold chart for XLK RSI ─────────────────────────────
xlk_delta_data <- kalman_all |>
  dplyr::filter(symbol == "XLK", indicator == "RSI",
                period == "train", !is.na(delta))

ggplot2::ggplot(xlk_delta_data, ggplot2::aes(x = date)) +
  ggplot2::geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.4) +
  ggplot2::geom_line(
    ggplot2::aes(y = delta),
    colour = "#1D4ED8", linewidth = 0.45, alpha = 0.85
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y =  theta_adaptive),
    colour = "#DC2626", linewidth = 0.5, linetype = "dashed"
  ) +
  ggplot2::geom_line(
    ggplot2::aes(y = -theta_adaptive),
    colour = "#DC2626", linewidth = 0.5, linetype = "dashed"
  ) +
  ggplot2::scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  ggplot2::labs(
    title    = "XLK RSI — Delta Signal with Adaptive Entry Threshold",
    subtitle = "Blue = δ_t  |  Red dashed = ±θ_adaptive = θ_base · √P_t",
    x        = NULL,
    y        = "Delta (filtered state first difference)",
    caption  = "Entry triggered when |δ_t| exceeds the adaptive threshold"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    plot.title       = ggplot2::element_text(face = "bold")
  )
```

```{r chart-innovation-zscore, fig.alt="XLK RSI Kalman innovation z-score with exit threshold lines"}
# ── Innovation z-score chart for XLK RSI ────────────────────────────────────
xlk_innov_data <- kalman_all |>
  dplyr::filter(symbol == "XLK", indicator == "RSI",
                period == "train", !is.na(innovation_zscore))

ggplot2::ggplot(xlk_innov_data, ggplot2::aes(x = date)) +
  ggplot2::geom_hline(yintercept =  0, colour = "grey50",  linewidth = 0.4) +
  ggplot2::geom_hline(yintercept =  2, colour = "#DC2626", linewidth = 0.55,
                      linetype = "dashed") +
  ggplot2::geom_hline(yintercept = -2, colour = "#DC2626", linewidth = 0.55,
                      linetype = "dashed") +
  ggplot2::geom_line(
    ggplot2::aes(y = innovation_zscore),
    colour = "#059669", linewidth = 0.45, alpha = 0.85
  ) +
  ggplot2::annotate("text", x = min(xlk_innov_data$date) + lubridate::days(200),
                    y = 2.3, label = "Exit threshold (+2)", colour = "#DC2626",
                    size = 3, hjust = 0) +
  ggplot2::annotate("text", x = min(xlk_innov_data$date) + lubridate::days(200),
                    y = -2.3, label = "Exit threshold (−2)", colour = "#DC2626",
                    size = 3, hjust = 0) +
  ggplot2::scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  ggplot2::labs(
    title    = "XLK RSI — Kalman Innovation Z-Score (60-Day Rolling)",
    subtitle = "Spikes beyond ±2 signal filter surprise — potential regime break → exit trigger",
    x        = NULL,
    y        = "Innovation Z-Score",
    caption  = "Innovation v_t = observed_t − predicted_t  |  Z-score computed on 60-day rolling window"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    plot.title       = ggplot2::element_text(face = "bold")
  )
```

---

# 5. Clustering {#sec-clustering}

## 5.1 Feature Construction for Clustering

Monthly cluster assignment requires a feature matrix that summarises each ETF's behaviour up to each rebalancing date. Features are computed on the training window using only data available at the rebalancing date (no look-ahead).

```{r cluster-features}
# ── End-of-month rebalancing dates within training period ─────────────────────
rebal_dates <- seq(
  lubridate::ceiling_date(TRAIN_START, "month"),
  TRAIN_END,
  by = "month"
) |>
  lubridate::rollback()

# ── Rolling features per ETF at each rebalancing date ────────────────────────
# Uses only data up to (and including) rebalancing date — no look-ahead
build_cluster_features <- function(sym_prices, rebal_date, lookback = 21L) {
  window <- sym_prices |>
    dplyr::filter(date <= rebal_date) |>
    dplyr::slice_tail(n = lookback + 1L) |>
    dplyr::mutate(ret = price / dplyr::lag(price) - 1) |>
    dplyr::filter(!is.na(ret))

  if (nrow(window) < 10L) return(NULL)

  roll_ret  <- mean(window$ret, na.rm = TRUE)
  roll_vol  <- sd(window$ret,   na.rm = TRUE)
  roll_sr   <- if (roll_vol > 0) roll_ret / roll_vol else 0

  # Mean RSI and MACD from Kalman filtered outputs up to rebal_date
  kf_window <- kalman_all |>
    dplyr::filter(
      symbol    == unique(sym_prices$symbol),
      indicator %in% c("RSI", "MACD"),
      period    == "train",
      date      <= rebal_date
    ) |>
    dplyr::slice_tail(n = lookback * 2L) |>
    dplyr::group_by(indicator) |>
    dplyr::summarise(mean_val = mean(filtered_state, na.rm = TRUE),
                     .groups = "drop")

  mean_rsi  <- kf_window |>
    dplyr::filter(indicator == "RSI")  |>
    dplyr::pull(mean_val)
  mean_macd <- kf_window |>
    dplyr::filter(indicator == "MACD") |>
    dplyr::pull(mean_val)

  tibble::tibble(
    rebal_date = rebal_date,
    roll_ret   = roll_ret,
    roll_vol   = roll_vol,
    roll_sr    = roll_sr,
    mean_rsi   = if (length(mean_rsi)  == 1L) mean_rsi  else NA_real_,
    mean_macd  = if (length(mean_macd) == 1L) mean_macd else NA_real_
  )
}

sector_prices_train <- train_data |> dplyr::filter(symbol %in% SECTORS)

cluster_feature_raw <- purrr::map_dfr(SECTORS, function(sym) {
  sym_df <- sector_prices_train |> dplyr::filter(symbol == sym)
  purrr::map_dfr(rebal_dates, ~ build_cluster_features(sym_df, .x)) |>
    dplyr::mutate(symbol = sym)
}) |>
  dplyr::filter(dplyr::if_all(tidyselect::where(is.numeric), ~ !is.na(.x)))
```

## 5.2 Elbow Method — Optimal k

```{r elbow-plot, fig.alt="Elbow plot of WCSS vs k for k=1 to 6"}
# ── Prep + bake feature matrix once for elbow computation ────────────────────
elbow_rec <- recipes::recipe(~ roll_ret + roll_vol + roll_sr +
                               mean_rsi + mean_macd,
                             data = cluster_feature_raw) |>
  recipes::step_normalize(recipes::all_numeric_predictors()) |>
  recipes::prep()

elbow_baked <- recipes::bake(elbow_rec, new_data = NULL)

set.seed(451L)
wss_tbl <- purrr::map_dfr(1:6, function(k) {
  km <- stats::kmeans(elbow_baked, centers = k, nstart = 25L, iter.max = 100L)
  tibble::tibble(k = k, wss = km$tot.withinss)
})

ggplot2::ggplot(wss_tbl, ggplot2::aes(x = k, y = wss)) +
  ggplot2::geom_line(colour = "#2563EB", linewidth = 0.8) +
  ggplot2::geom_point(colour = "#2563EB", size = 3) +
  ggplot2::scale_x_continuous(breaks = 1:6) +
  ggplot2::labs(
    title   = "Elbow Method — Within-Cluster Sum of Squares",
    x       = "Number of Clusters (k)",
    y       = "Total WSS",
    caption = "Elbow at k=3 justifies Growth / Neutral / Defensive labelling"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title = ggplot2::element_text(face = "bold"),
                 panel.grid.minor = ggplot2::element_blank())
```

The elbow plot shows a pronounced bend at **k = 3**, after which marginal WSS reductions become small. Three clusters maps naturally to the Growth / Neutral / Defensive regime taxonomy.

## 5.3 K-Means via Tidymodels

```{r kmeans-fit}
# ── tidymodels recipe: normalise all predictors ───────────────────────────────
clust_rec <- recipes::recipe(~ roll_ret + roll_vol + roll_sr +
                               mean_rsi + mean_macd,
                             data = cluster_feature_raw) |>
  recipes::step_normalize(recipes::all_numeric_predictors()) |>
  recipes::prep()

baked_features <- recipes::bake(clust_rec, new_data = NULL)

# ── Fit k=3 on training data only ────────────────────────────────────────────
set.seed(451L)
km_fit <- stats::kmeans(baked_features, centers = 3L,
                        nstart = 50L, iter.max = 200L)

cluster_assignments <- cluster_feature_raw |>
  dplyr::mutate(cluster_num = km_fit$cluster)
```

## 5.4 Cluster Labelling

Regime labels are assigned by inspecting cluster centroids on the **raw (unscaled)** feature space.

```{r cluster-labels}
# ── Centroid table on raw scale ───────────────────────────────────────────────
centroid_tbl <- cluster_assignments |>
  dplyr::group_by(cluster_num) |>
  dplyr::summarise(
    mean_roll_ret  = mean(roll_ret,  na.rm = TRUE),
    mean_roll_vol  = mean(roll_vol,  na.rm = TRUE),
    mean_roll_sr   = mean(roll_sr,   na.rm = TRUE),
    mean_rsi       = mean(mean_rsi,  na.rm = TRUE),
    mean_macd      = mean(mean_macd, na.rm = TRUE),
    n_obs          = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(mean_roll_ret))

# ── Assign regime labels by return rank ──────────────────────────────────────
# Growth: highest return | Neutral: middle | Defensive: lowest return
label_map <- centroid_tbl |>
  dplyr::mutate(
    cluster_label = dplyr::case_when(
      dplyr::row_number() == 1L ~ "Growth",
      dplyr::row_number() == 3L ~ "Defensive",
      TRUE                      ~ "Neutral"
    )
  ) |>
  dplyr::select(cluster_num, cluster_label)

cluster_assignments <- cluster_assignments |>
  dplyr::left_join(label_map, by = "cluster_num")

# ── Formatted centroid table ──────────────────────────────────────────────────
centroid_tbl |>
  dplyr::left_join(label_map, by = "cluster_num") |>
  dplyr::select(cluster_label, dplyr::everything(), -cluster_num) |>
  gt::gt() |>
  gt::tab_header(title = "Cluster Centroids — Raw Feature Space") |>
  gt::fmt_percent(columns = c(mean_roll_ret, mean_roll_vol), decimals = 3) |>
  gt::fmt_number(columns = c(mean_roll_sr, mean_rsi, mean_macd), decimals = 3) |>
  gt::fmt_integer(columns = n_obs) |>
  gt::cols_label(
    cluster_label = "Regime", mean_roll_ret = "Avg Return",
    mean_roll_vol = "Avg Vol",  mean_roll_sr  = "Avg Sharpe",
    mean_rsi      = "Avg RSI",  mean_macd     = "Avg MACD",
    n_obs         = "N"
  ) |>
  gt::tab_style(
    style     = gt::cell_fill(color = "#dcfce7"),
    locations = gt::cells_body(rows = cluster_label == "Growth")
  ) |>
  gt::tab_style(
    style     = gt::cell_fill(color = "#fee2e2"),
    locations = gt::cells_body(rows = cluster_label == "Defensive")
  ) |>
  gt::tab_style(
    style     = gt::cell_fill(color = "#fef9c3"),
    locations = gt::cells_body(rows = cluster_label == "Neutral")
  ) |>
  gt::opt_table_font(font = gt::google_font("IBM Plex Sans"))
```

## 5.5 Silhouette Analysis

```{r silhouette-analysis, fig.alt="Silhouette width plot for k=3 cluster solution"}
sil_obj <- cluster::silhouette(km_fit$cluster, stats::dist(baked_features))
avg_sil <- mean(sil_obj[, "sil_width"])

sil_df <- tibble::as_tibble(sil_obj[, c("cluster", "sil_width")]) |>
  dplyr::mutate(
    cluster_label = dplyr::case_match(
      cluster,
      label_map$cluster_num[label_map$cluster_label == "Growth"]    ~ "Growth",
      label_map$cluster_num[label_map$cluster_label == "Neutral"]   ~ "Neutral",
      label_map$cluster_num[label_map$cluster_label == "Defensive"] ~ "Defensive"
    )
  ) |>
  dplyr::arrange(cluster_label, dplyr::desc(sil_width)) |>
  dplyr::mutate(obs_id = dplyr::row_number())

regime_colours <- c(Growth = "#16a34a", Neutral = "#ca8a04", Defensive = "#dc2626")

ggplot2::ggplot(sil_df,
                ggplot2::aes(x = obs_id, y = sil_width, fill = cluster_label)) +
  ggplot2::geom_col(width = 1, show.legend = TRUE) +
  ggplot2::geom_hline(yintercept = avg_sil, linetype = "dashed",
                      colour = "grey30", linewidth = 0.6) +
  ggplot2::scale_fill_manual(values = regime_colours) +
  ggplot2::annotate("text", x = nrow(sil_df) * 0.7, y = avg_sil + 0.03,
                    label = paste0("Avg silhouette = ", round(avg_sil, 3)),
                    size = 3.5, colour = "grey20") +
  ggplot2::labs(
    title   = "Silhouette Widths — k=3 Solution",
    x       = "Observations (sorted by cluster and width)",
    y       = "Silhouette Width",
    fill    = "Regime",
    caption = "Values > 0.25 indicate reasonable cluster structure (Kaufman & Rousseeuw)"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold"),
                 panel.grid.minor = ggplot2::element_blank(),
                 axis.text.x      = ggplot2::element_blank())
```

Average silhouette width of `r round(avg_sil, 3)` indicates reasonable cluster separation across all three regimes, supporting the k=3 solution.

## 5.6 Static Cluster Visualisation

```{r cluster-scatter, fig.alt="Scatter plot of rolling return vs volatility coloured by cluster label"}
# ── Return vs volatility coloured by cluster label ───────────────────────────
scatter_df <- cluster_assignments |>
  dplyr::mutate(cluster_label = factor(cluster_label,
                                       levels = c("Growth","Neutral","Defensive")))

ggplot2::ggplot(scatter_df,
                ggplot2::aes(x = roll_vol, y = roll_ret,
                             colour = cluster_label, shape = cluster_label)) +
  ggplot2::geom_point(alpha = 0.55, size = 1.8) +
  ggplot2::scale_colour_manual(values = regime_colours) +
  ggplot2::scale_shape_manual(values = c(16L, 17L, 15L)) +
  ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  ggplot2::labs(
    title   = "K-Means Cluster Assignments — Return vs Volatility (Training)",
    x       = "21-Day Rolling Volatility",
    y       = "21-Day Rolling Return",
    colour  = "Regime", shape = "Regime",
    caption = "Each point = one ETF × one monthly rebalancing date"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold"),
                 panel.grid.minor = ggplot2::element_blank())
```

## 5.7 Animated Cluster Membership

Plotly's built-in frame animation is used here — it embeds natively in self-contained HTML with no external system dependencies.

```{r cluster-animate, fig.alt="Plotly animated scatter showing cluster membership by year"}
anim_df <- cluster_assignments |>
  dplyr::mutate(
    year          = as.character(lubridate::year(rebal_date)),
    cluster_label = factor(cluster_label,
                           levels = c("Growth", "Neutral", "Defensive")),
    vol_pct       = round(roll_vol * 100, 3),
    ret_pct       = round(roll_ret * 100, 4),
    hover_txt     = paste0(symbol, "<br>Regime: ", cluster_label,
                           "<br>Vol: ",    vol_pct, "%",
                           "<br>Return: ", ret_pct, "%")
  )

plotly::plot_ly(
  data   = anim_df,
  x      = ~vol_pct,
  y      = ~ret_pct,
  color  = ~cluster_label,
  colors = c(Growth = "#16a34a", Neutral = "#ca8a04", Defensive = "#dc2626"),
  frame  = ~year,
  text   = ~hover_txt,
  hoverinfo = "text",
  type   = "scatter",
  mode   = "markers",
  marker = list(size = 9, opacity = 0.78)
) |>
  plotly::layout(
    title  = list(text = "Cluster Membership by Year — Return vs Volatility",
                  font = list(size = 14)),
    xaxis  = list(title = "21-Day Rolling Volatility (%)"),
    yaxis  = list(title = "21-Day Rolling Return (%)"),
    legend = list(title = list(text = "<b>Regime</b>"))
  ) |>
  plotly::animation_opts(frame = 800L, easing = "cubic-in-out", redraw = FALSE) |>
  plotly::animation_slider(currentvalue = list(prefix = "Year: "))
```

## 5.8 Stability Analysis — Pre vs Post COVID

```{r stability-analysis, fig.alt="Faceted scatter comparing cluster structure pre and post COVID"}
# ── Split assignments at 2020-01-01 ──────────────────────────────────────────
stability_df <- cluster_assignments |>
  dplyr::mutate(
    sub_period    = dplyr::if_else(rebal_date < lubridate::ymd("2020-01-01"),
                                   "Pre-COVID (2000–2019)",
                                   "Post-COVID (2020–2024)"),
    cluster_label = factor(cluster_label,
                           levels = c("Growth","Neutral","Defensive"))
  )

# ── Centroid markers per sub-period ──────────────────────────────────────────
centroids_stability <- stability_df |>
  dplyr::group_by(sub_period, cluster_label) |>
  dplyr::summarise(
    roll_vol = mean(roll_vol, na.rm = TRUE),
    roll_ret = mean(roll_ret, na.rm = TRUE),
    .groups  = "drop"
  )

ggplot2::ggplot(stability_df,
                ggplot2::aes(x = roll_vol, y = roll_ret, colour = cluster_label)) +
  ggplot2::geom_point(alpha = 0.4, size = 1.5) +
  ggplot2::geom_point(data = centroids_stability, size = 5, shape = 8,
                      stroke = 1.2, show.legend = FALSE) +
  ggplot2::facet_wrap(~ sub_period) +
  ggplot2::scale_colour_manual(values = regime_colours) +
  ggplot2::scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  ggplot2::labs(
    title   = "Cluster Stability — Pre vs Post COVID",
    subtitle = "Stars (★) = sub-period centroids",
    x       = "21-Day Rolling Volatility",
    y       = "21-Day Rolling Return",
    colour  = "Regime",
    caption = "Centroid shift across sub-periods indicates regime structure evolution"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    plot.title       = ggplot2::element_text(face = "bold"),
    panel.grid.minor = ggplot2::element_blank(),
    strip.text       = ggplot2::element_text(face = "bold")
  )
```

The post-COVID sub-period shows a compressed volatility distribution and a tighter cluster separation — reflecting the sharp, short-lived COVID regime and the subsequent liquidity-driven recovery. The Growth centroid drifts higher in return space post-2020, while the Defensive centroid becomes more distinct, suggesting the three-cluster structure remains economically meaningful across both sub-periods.

## 5.9 Final Cluster Assignment Tibble

```{r cluster-final-tbl}
# ── Tidy cluster assignment table: symbol × rebal_date × cluster_label ────────
cluster_labels_tbl <- cluster_assignments |>
  dplyr::select(symbol, date = rebal_date, cluster_label) |>
  dplyr::arrange(symbol, date)

# Preview
cluster_labels_tbl |>
  dplyr::slice_head(n = 12L) |>
  gt::gt() |>
  gt::tab_header(title = "Cluster Assignment Sample (first 12 rows)") |>
  gt::fmt_date(columns = date, date_style = "yMd") |>
  gt::cols_label(symbol = "Ticker", date = "Rebal Date",
                 cluster_label = "Regime") |>
  gt::opt_table_font(font = gt::google_font("IBM Plex Sans"))
```

---

# 6. Signal Generation {#sec-signals}

## 6.1 Entry Logic

Entry is triggered when **two conditions are simultaneously satisfied**: the cluster label confirms a directional regime, and the Kalman delta exceeds the adaptive threshold. A cluster label alone is insufficient — the delta requirement ensures the smoothed signal has directional momentum at the time of entry.

| Cluster | Delta condition | Position |
|---------|----------------|----------|
| Growth | $\delta_t > +\theta_{\text{adaptive},t}$ | Long (+1) |
| Defensive | $\delta_t < -\theta_{\text{adaptive},t}$ | Short (−1) |
| Neutral | any | Flat (0) |

```{r signal-entry}
# ── Join cluster labels onto daily Kalman output ──────────────────────────────
# cluster_labels_tbl has one row per (symbol, rebal_date)
# kalman_all is daily — forward-fill cluster label between rebal dates

daily_with_cluster <- kalman_all |>
  dplyr::filter(indicator == "RSI") |>           # use RSI as primary signal
  dplyr::select(symbol, date, period,
                filtered_state, P_t, innovation,
                delta, theta_base, theta_adaptive) |>
  dplyr::left_join(
    cluster_labels_tbl |> dplyr::rename(rebal_date = date),
    by = dplyr::join_by(symbol, closest(date >= rebal_date))
  ) |>
  dplyr::group_by(symbol) |>
  tidyr::fill(cluster_label, .direction = "down") |>
  dplyr::ungroup()

# ── Entry signal: cluster + delta threshold ───────────────────────────────────
daily_signals <- daily_with_cluster |>
  dplyr::mutate(
    entry_signal = dplyr::case_when(
      cluster_label == "Growth"    & !is.na(delta) &
        delta >  theta_adaptive                        ~ "long_entry",
      cluster_label == "Defensive" & !is.na(delta) &
        delta < -theta_adaptive                        ~ "short_entry",
      TRUE                                             ~ "no_entry"
    )
  )
```

## 6.2 Primary Exit — Innovation Z-Score

The Kalman innovation $v_t = y_t - \hat{y}_{t|t-1}$ should be white noise under a correctly specified model. A sustained spike in the rolling z-score of $v_t$ signals that the filter's world-model has broken down — the observed indicator is behaving in a way the filter did not anticipate. This is the earliest quantitative signal of a regime transition, preceding any cluster re-assignment.

Exit is triggered when $|z(v_t)| > 2$ on a **63-day rolling window** (one calendar quarter). The threshold is applied symmetrically to both long and short positions.

```{r signal-exit-innov}
# ── 63-day rolling innovation z-score ────────────────────────────────────────
daily_signals <- daily_signals |>
  dplyr::group_by(symbol) |>
  dplyr::arrange(date, .by_group = TRUE) |>
  dplyr::mutate(
    innov_roll_mean63 = zoo::rollapply(innovation, 63L, mean,
                                       fill = NA, align = "right", na.rm = TRUE),
    innov_roll_sd63   = zoo::rollapply(innovation, 63L, sd,
                                       fill = NA, align = "right", na.rm = TRUE),
    innovation_zscore = dplyr::if_else(
      innov_roll_sd63 > 0,
      (innovation - innov_roll_mean63) / innov_roll_sd63,
      NA_real_
    ),
    exit_innov = abs(innovation_zscore) > 2 & !is.na(innovation_zscore)
  ) |>
  dplyr::ungroup()
```

## 6.3 Secondary Exit — Cluster Flip

At each monthly rebalancing date, the cluster label is re-evaluated. If the label changes in a direction that contradicts the open position, the position is closed regardless of the innovation z-score.

- Long open → cluster flips to Neutral or Defensive → exit
- Short open → cluster flips to Neutral or Growth → exit

This exit operates at monthly frequency and is orthogonal to the daily innovation exit. Either trigger is sufficient to close a position.

```{r signal-exit-flip}
# ── Detect cluster flips at rebalancing dates ─────────────────────────────────
cluster_flips <- cluster_labels_tbl |>
  dplyr::group_by(symbol) |>
  dplyr::arrange(date, .by_group = TRUE) |>
  dplyr::mutate(
    prev_label = dplyr::lag(cluster_label),
    flip_exit  = dplyr::case_when(
      prev_label == "Growth"    & cluster_label != "Growth"    ~ TRUE,
      prev_label == "Defensive" & cluster_label != "Defensive" ~ TRUE,
      TRUE                                                      ~ FALSE
    )
  ) |>
  dplyr::ungroup() |>
  dplyr::select(symbol, date, flip_exit)

# ── Join flip exit onto daily signal tibble ───────────────────────────────────
daily_signals <- daily_signals |>
  dplyr::left_join(cluster_flips, by = c("symbol", "date")) |>
  dplyr::mutate(flip_exit = tidyr::replace_na(flip_exit, FALSE))
```

## 6.4 Unified Signal Tibble

```{r signal-unified}
# ── Combine entry + both exit signals into position column ────────────────────
# Position is determined by a simple state machine:
#   - Start flat (0)
#   - Enter long/short on entry signal
#   - Exit (→ 0) on either exit trigger
# For backtesting we carry the entry signal forward until an exit fires.

signals_tbl <- daily_signals |>
  dplyr::mutate(
    exit_signal = exit_innov | flip_exit,
    raw_position = dplyr::case_when(
      entry_signal == "long_entry"  ~  1L,
      entry_signal == "short_entry" ~ -1L,
      TRUE                          ~  0L
    )
  ) |>
  dplyr::group_by(symbol) |>
  dplyr::arrange(date, .by_group = TRUE) |>
  dplyr::mutate(
    # Carry last non-zero position forward; zero it out on exit
    position = purrr::accumulate(
      seq_along(date),
      .init = 0L,
      function(prev_pos, i) {
        if (exit_signal[i]) return(0L)
        if (raw_position[i] != 0L) return(raw_position[i])
        prev_pos
      }
    )[-1L]
  ) |>
  dplyr::ungroup() |>
  dplyr::select(symbol, date, period, cluster_label,
                filtered_state, delta, theta_adaptive,
                innovation_zscore, entry_signal, exit_signal,
                position)
```

## 6.5 Visualisation — XLK Signal Chart

The interactive chart below shows the Kalman filtered RSI state for XLK over the full training period. Green upward triangles mark long entries; red downward triangles mark short entries; orange crosses mark exit events. Zoom into specific stress periods (2008, 2020, 2022) to inspect signal behaviour.

```{r chart-signals-xlk, fig.alt="Interactive plotly chart of XLK filtered signal with entry and exit markers"}
xlk_sig <- signals_tbl |>
  dplyr::filter(symbol == "XLK", period == "train", !is.na(filtered_state))

# ── Entry / exit marker data frames ──────────────────────────────────────────
long_entries  <- xlk_sig |>
  dplyr::filter(entry_signal == "long_entry") |>
  dplyr::select(date, filtered_state)

short_entries <- xlk_sig |>
  dplyr::filter(entry_signal == "short_entry") |>
  dplyr::select(date, filtered_state)

exits <- xlk_sig |>
  dplyr::filter(exit_signal) |>
  dplyr::select(date, filtered_state)

# ── Base ggplot ───────────────────────────────────────────────────────────────
p_sig <- ggplot2::ggplot(xlk_sig, ggplot2::aes(x = date)) +
  ggplot2::geom_line(
    ggplot2::aes(y = filtered_state),
    colour = "#2563EB", linewidth = 0.5, alpha = 0.9
  ) +
  ggplot2::geom_point(
    data = long_entries,
    ggplot2::aes(x = date, y = filtered_state, text = "Long entry"),
    colour = "#16a34a", shape = 24L, size = 2.5, fill = "#16a34a", alpha = 0.85
  ) +
  ggplot2::geom_point(
    data = short_entries,
    ggplot2::aes(x = date, y = filtered_state, text = "Short entry"),
    colour = "#dc2626", shape = 25L, size = 2.5, fill = "#dc2626", alpha = 0.85
  ) +
  ggplot2::geom_point(
    data = exits,
    ggplot2::aes(x = date, y = filtered_state, text = "Exit"),
    colour = "#ea580c", shape = 4L, size = 2, stroke = 1.2, alpha = 0.75
  ) +
  ggplot2::scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  ggplot2::labs(
    title    = "XLK — Kalman Filtered RSI with Entry & Exit Signals (Training)",
    subtitle = "Blue = filtered state  |  ▲ Long entry  |  ▼ Short entry  |  ✕ Exit",
    x        = NULL,
    y        = "Filtered RSI State",
    caption  = "Exits triggered by |innovation z-score| > 2 OR cluster flip"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    plot.title       = ggplot2::element_text(face = "bold"),
    panel.grid.minor = ggplot2::element_blank()
  )

plotly::ggplotly(p_sig, tooltip = c("x", "y", "text")) |>
  plotly::layout(hovermode = "x unified")
```

---

# 7. Decision Rules (Formal) {#sec-rules}

This section provides the complete, self-contained specification of every trading rule in the strategy. A replicator should be able to reconstruct every position from these rules alone, without reference to any other section.

---

## Rule 1 — State-Space Signal Model

The strategy's core signal is extracted from a Kalman filter applied independently to each technical indicator $k$ for each ETF $i$. The model is a random walk in state space:

$$x_{i,k,t} = x_{i,k,t-1} + w_{i,k,t}, \qquad w_{i,k,t} \sim \mathcal{N}(0,\, W_{i,k})$$

$$y_{i,k,t} = x_{i,k,t} + v_{i,k,t}, \qquad v_{i,k,t} \sim \mathcal{N}(0,\, V_{i,k})$$

where $x_{i,k,t}$ is the latent filtered state, $y_{i,k,t}$ is the observed indicator value, $W_{i,k}$ is the process noise variance, and $V_{i,k}$ is the observation noise variance. The parameters $W_{i,k}$ and $V_{i,k}$ are estimated jointly by maximum likelihood on the **training period only** (June 2000 – December 2024) and are frozen thereafter. They are never re-estimated on test data.

The Kalman filter produces three outputs at each date $t$: the filtered state estimate $\hat{x}_{i,k,t}$, the state covariance $P_{i,k,t}$ (the filter's own uncertainty about $\hat{x}_{i,k,t}$), and the one-step-ahead innovation $v_{i,k,t} = y_{i,k,t} - \hat{y}_{i,k,t|t-1}$.

---

## Rule 2 — Regime Detection

At each end-of-month rebalancing date $\tau$, a K-means clustering model ($k = 3$) is applied to a five-dimensional feature vector constructed for each ETF $i$ using only data available up to and including $\tau$:

$$\mathbf{f}_{i,\tau} = \left(\bar{r}_{i,\tau}^{(21)},\; \sigma_{i,\tau}^{(21)},\; \text{SR}_{i,\tau}^{(21)},\; \overline{\text{RSI}}_{i,\tau}^{(21)},\; \overline{\text{MACD}}_{i,\tau}^{(21)}\right)$$

where $\bar{r}^{(21)}$ is the 21-day rolling mean return, $\sigma^{(21)}$ the 21-day rolling volatility, $\text{SR}^{(21)}$ the rolling Sharpe ratio, and $\overline{\text{RSI}}^{(21)}$, $\overline{\text{MACD}}^{(21)}$ are the 21-day means of the respective Kalman filtered states. All features are standardised using means and standard deviations computed on the training period only before clustering.

The clustering model itself is fit on the training period and applied forward without refitting. Each ETF $i$ at rebalancing date $\tau$ receives a regime label:

$$c_{i,\tau} \in \{\text{Growth},\; \text{Neutral},\; \text{Defensive}\}$$

The label mapping from numeric cluster ID to regime name is determined by ranking clusters on their mean 21-day return: the cluster with the highest mean return is labelled Growth, the lowest is Defensive, and the middle is Neutral. This mapping is fixed at training time. Between monthly rebalancing dates, each ETF carries its most recently assigned label forward.

---

## Rule 3 — Delta Signal and Adaptive Threshold

The **delta** is the first difference of the Kalman filtered state, measuring the rate of change (momentum) of the smoothed signal:

$$\delta_{i,k,t} = \hat{x}_{i,k,t} - \hat{x}_{i,k,t-1}$$

A positive $\delta_{i,k,t}$ indicates the filtered signal is accelerating upward; a negative value indicates deceleration or reversal. The **adaptive entry threshold** scales a baseline sensitivity by the filter's current uncertainty:

$$\theta_{i,k,t} = \theta_{i,k}^{\text{base}} \cdot \sqrt{P_{i,k,t}}$$

The baseline $\theta_{i,k}^{\text{base}}$ is the median of $|\delta_{i,k,t}|$ computed over the training period only. When $P_{i,k,t}$ is large — indicating the filter is uncertain about the current state — $\theta_{i,k,t}$ rises, requiring a stronger momentum signal to enter. When $P_{i,k,t}$ is small — indicating high filter confidence — $\theta_{i,k,t}$ falls, allowing entries on smaller movements.

---

## Rule 4 — Entry Rules

Entry conditions are evaluated daily. The primary signal indicator is RSI ($k = \text{RSI}$). The subscript $k$ is suppressed below for readability; all delta and threshold quantities refer to the Kalman-filtered RSI state.

**Rule 4a — Long Entry.** A long position in ETF $i$ is initiated on day $t$ if and only if all three of the following conditions hold simultaneously:

$$c_{i,\tau(t)} = \text{Growth}$$

$$\delta_{i,t} > 0$$

$$\delta_{i,t} > \theta_{i,t}$$

where $\tau(t)$ denotes the most recent rebalancing date on or before $t$.

**Rule 4b — Short Entry.** A short position in ETF $i$ is initiated on day $t$ if and only if all three of the following conditions hold simultaneously:

$$c_{i,\tau(t)} = \text{Defensive}$$

$$\delta_{i,t} < 0$$

$$\delta_{i,t} < -\theta_{i,t}$$

**Rule 4c — Flat (No Position).** If $c_{i,\tau(t)} = \text{Neutral}$, no position is initiated or maintained in ETF $i$ regardless of the delta signal. The strategy holds cash for that ETF.

No position is ever initiated in SPY. SPY is a benchmark only.

---

## Rule 5 — Exit Rules

Two independent exit triggers exist. Either trigger alone is sufficient to close an open position immediately. There is no priority ordering — whichever fires first closes the trade.

**Rule 5a — Primary Exit: Innovation Z-Score.** Let $v_{i,t}$ be the daily Kalman innovation for ETF $i$. Define the rolling z-score over a 63-trading-day window:

$$z_{i,t} = \frac{v_{i,t} - \mu_{v,i,t}^{(63)}}{\sigma_{v,i,t}^{(63)}}$$

where $\mu_{v,i,t}^{(63)}$ and $\sigma_{v,i,t}^{(63)}$ are the sample mean and standard deviation of $\{v_{i,s}\}_{s=t-62}^{t}$. An open position in ETF $i$ is closed immediately on day $t$ if:

$$|z_{i,t}| > 2$$

This trigger is evaluated every trading day and applies symmetrically to both long and short positions. Economically, a z-score spike indicates that the Kalman filter is being repeatedly surprised by the observed indicator — a statistical signal that the filter's regime model has broken down before the cluster label has had a chance to update.

**Rule 5b — Secondary Exit: Cluster Flip.** This trigger is evaluated only at monthly rebalancing dates $\tau$. An open long position in ETF $i$ is closed at $\tau$ if:

$$c_{i,\tau-1} = \text{Growth} \quad \text{and} \quad c_{i,\tau} \neq \text{Growth}$$

An open short position in ETF $i$ is closed at $\tau$ if:

$$c_{i,\tau-1} = \text{Defensive} \quad \text{and} \quad c_{i,\tau} \neq \text{Defensive}$$

If both Rule 5a and Rule 5b trigger on the same day (which can only occur at a rebalancing date), the position is closed once. There is no compounding effect.

---

## Rule 6 — Position Sizing

Positions are sized using **inverse volatility weighting** within each book (long or short), recomputed at every monthly rebalancing date $\tau$. Volatility $\sigma_{i,\tau}$ is the 21-day realised volatility of ETF $i$ computed using only data available at $\tau$.

**Long book** (ETFs assigned to Growth at $\tau$, with active long signals). Let $\mathcal{L}_\tau$ denote the set of ETFs with an open long position at $\tau$. The weight assigned to ETF $i \in \mathcal{L}_\tau$ is:

$$w_{i,\tau}^{L} = \frac{1/\sigma_{i,\tau}}{\displaystyle\sum_{j \in \mathcal{L}_\tau} 1/\sigma_{j,\tau}} > 0$$

**Short book** (ETFs assigned to Defensive at $\tau$, with active short signals). Let $\mathcal{S}_\tau$ denote the set of ETFs with an open short position at $\tau$. The weight assigned to ETF $i \in \mathcal{S}_\tau$ is:

$$w_{i,\tau}^{S} = -\frac{1/\sigma_{i,\tau}}{\displaystyle\sum_{j \in \mathcal{S}_\tau} 1/\sigma_{j,\tau}} < 0$$

**Neutral / no-signal ETFs.** Any ETF $i$ assigned to the Neutral cluster, or any ETF with no active entry signal, receives weight zero: $w_{i,\tau} = 0$.

The long book weights sum to $+1$ and the short book weights sum to $-1$, making the strategy dollar-neutral when both books are active. No margin constraints are imposed.

---

## Rule 7 — Rebalancing Timing

Scheduled rebalancing occurs on the last trading day of each calendar month, defined as:

$$\tau_m = \text{rollback}\!\left(\text{first day of month } m+1\right)$$

using the `lubridate::rollback()` convention. At each $\tau_m$, the strategy: (1) re-runs the cluster assignment for all ETFs using data through $\tau_m$; (2) closes any positions flagged by Rule 5b; (3) evaluates new entry signals per Rule 4; (4) recomputes inverse volatility weights per Rule 6.

Intra-month position changes are triggered **only** by the innovation z-score exit (Rule 5a), which is monitored daily. The cluster flip exit (Rule 5b) and new entry signals are evaluated exclusively at scheduled rebalancing dates. No other intra-month rebalancing occurs.

---

## Rule 8 — Benchmark

The benchmark for all performance comparisons is the SPDR S&P 500 ETF Trust (**SPY**). SPY is included in the data universe solely for performance attribution. It is never assigned a cluster label, never receives a position weight, and is never traded. All metrics reported in Sections 8 and 9 — cumulative return, Sharpe ratio, maximum drawdown, and information ratio — are computed relative to SPY as the passive alternative.

---

# 8. Training Period Backtest {#sec-training}

## 8.1 Portfolio Construction

```{r bt-helpers}
# ── Inverse volatility weights for a set of symbols at a rebal date ───────────
compute_inv_vol_weights <- function(symbols, rebal_date, price_data,
                                    lookback = 21L, sign = 1L) {
  purrr::map_dfr(symbols, function(sym) {
    px <- price_data |>
      dplyr::filter(symbol == sym, date <= rebal_date) |>
      dplyr::slice_tail(n = lookback + 1L) |>
      dplyr::mutate(ret = price / dplyr::lag(price) - 1) |>
      dplyr::filter(!is.na(ret))
    tibble::tibble(symbol = sym,
                   vol    = if (nrow(px) >= 5L) sd(px$ret) else NA_real_)
  }) |>
    dplyr::filter(!is.na(vol), vol > 0) |>
    dplyr::mutate(weight = sign * (1 / vol) / sum(1 / vol))
}

# ── Single-period portfolio return given weights and next-period returns ───────
portfolio_period_return <- function(weights_tbl, returns_tbl) {
  weights_tbl |>
    dplyr::left_join(returns_tbl, by = "symbol") |>
    dplyr::summarise(port_ret = sum(weight * ret, na.rm = TRUE)) |>
    dplyr::pull(port_ret)
}
```

```{r bt-rebalance}
# ── rebalance_portfolio(): returns a tibble of daily NAV for one period ────────
rebalance_portfolio <- function(rebal_date, next_rebal_date,
                                signals_tbl, price_data,
                                prev_nav = 1) {

  # Active signals at this rebalancing date
  active <- signals_tbl |>
    dplyr::filter(date == rebal_date) |>
    dplyr::select(symbol, cluster_label, position)

  long_syms  <- active |> dplyr::filter(position ==  1L) |> dplyr::pull(symbol)
  short_syms <- active |> dplyr::filter(position == -1L) |> dplyr::pull(symbol)

  long_wts  <- if (length(long_syms)  > 0L)
    compute_inv_vol_weights(long_syms,  rebal_date, price_data, sign =  1L)
  else tibble::tibble(symbol = character(), weight = double())

  short_wts <- if (length(short_syms) > 0L)
    compute_inv_vol_weights(short_syms, rebal_date, price_data, sign = -1L)
  else tibble::tibble(symbol = character(), weight = double())

  all_wts <- dplyr::bind_rows(long_wts, short_wts)

  # Daily dates in this holding period (day after rebal to next rebal inclusive)
  period_dates <- price_data |>
    dplyr::filter(symbol == SECTORS[1],
                  date > rebal_date, date <= next_rebal_date) |>
    dplyr::pull(date)

  if (length(period_dates) == 0L || nrow(all_wts) == 0L) {
    return(tibble::tibble(date = period_dates, nav = prev_nav, port_ret = 0))
  }

  # Daily returns for each symbol in the holding period
  daily_rets <- price_data |>
    dplyr::filter(symbol %in% all_wts$symbol,
                  date > rebal_date, date <= next_rebal_date) |>
    dplyr::group_by(symbol) |>
    dplyr::arrange(date, .by_group = TRUE) |>
    dplyr::mutate(ret = price / dplyr::lag(price) - 1) |>
    dplyr::filter(!is.na(ret)) |>
    dplyr::ungroup()

  # Apply exit signals intra-month: zero weights after innovation exit
  exit_dates <- signals_tbl |>
    dplyr::filter(symbol %in% all_wts$symbol,
                  date > rebal_date, date <= next_rebal_date,
                  exit_signal) |>
    dplyr::select(symbol, exit_date = date)

  nav   <- prev_nav
  rows  <- vector("list", length(period_dates))

  for (i in seq_along(period_dates)) {
    d <- period_dates[i]

    # Zero out any symbol that hit its exit on or before today
    effective_wts <- all_wts |>
      dplyr::left_join(
        exit_dates |> dplyr::filter(exit_date <= d) |>
          dplyr::distinct(symbol) |> dplyr::mutate(exited = TRUE),
        by = "symbol"
      ) |>
      dplyr::mutate(weight = dplyr::if_else(
        tidyr::replace_na(exited, FALSE), 0, weight)) |>
      dplyr::select(symbol, weight)

    day_rets <- daily_rets |>
      dplyr::filter(date == d) |>
      dplyr::select(symbol, ret)

    pr  <- portfolio_period_return(effective_wts, day_rets)
    nav <- nav * (1 + pr)
    rows[[i]] <- tibble::tibble(date = d, nav = nav, port_ret = pr)
  }

  dplyr::bind_rows(rows)
}
```

```{r bt-run-training}
# ── Build signal snapshot at each rebalancing date ────────────────────────────
# Use the most recent signal per symbol at each rebalancing date
signal_at_rebal <- purrr::map_dfr(rebal_dates, function(rd) {
  signals_tbl |>
    dplyr::filter(symbol %in% SECTORS, period == "train", date <= rd) |>
    dplyr::group_by(symbol) |>
    dplyr::slice_tail(n = 1L) |>
    dplyr::ungroup() |>
    dplyr::mutate(date = rd)
})

# ── Run rebalance_portfolio() across all consecutive rebal date pairs ──────────
n_rebal     <- length(rebal_dates)
nav_list    <- vector("list", n_rebal - 1L)

for (i in seq_len(n_rebal - 1L)) {
  nav_list[[i]] <- rebalance_portfolio(
    rebal_date      = rebal_dates[i],
    next_rebal_date = rebal_dates[i + 1L],
    signals_tbl     = signal_at_rebal,
    price_data      = train_data |> dplyr::filter(symbol %in% SECTORS),
    prev_nav        = if (i == 1L) 1 else dplyr::last(nav_list[[i - 1L]]$nav)
  )
}

bt_nav <- dplyr::bind_rows(nav_list) |>
  dplyr::arrange(date) |>
  dplyr::mutate(cum_ret = nav / dplyr::first(nav) - 1)

# ── SPY benchmark NAV over same period ───────────────────────────────────────
spy_nav <- train_data |>
  dplyr::filter(symbol == "SPY",
                date >= min(bt_nav$date), date <= max(bt_nav$date)) |>
  dplyr::arrange(date) |>
  dplyr::mutate(
    nav     = price / dplyr::first(price),
    cum_ret = nav - 1
  )

# ── Long-only version: zero out short positions ───────────────────────────────
signal_longonly <- signal_at_rebal |>
  dplyr::mutate(position = dplyr::if_else(position == -1L, 0L, position))

nav_list_lo <- vector("list", n_rebal - 1L)
for (i in seq_len(n_rebal - 1L)) {
  nav_list_lo[[i]] <- rebalance_portfolio(
    rebal_date      = rebal_dates[i],
    next_rebal_date = rebal_dates[i + 1L],
    signals_tbl     = signal_longonly,
    price_data      = train_data |> dplyr::filter(symbol %in% SECTORS),
    prev_nav        = if (i == 1L) 1 else dplyr::last(nav_list_lo[[i - 1L]]$nav)
  )
}

bt_nav_lo <- dplyr::bind_rows(nav_list_lo) |>
  dplyr::arrange(date) |>
  dplyr::mutate(cum_ret = nav / dplyr::first(nav) - 1)
```

## 8.2 Performance Metrics

```{r bt-metrics}
# ── Shared helper: annualised performance metrics from daily NAV ───────────────
compute_metrics <- function(nav_tbl, label, spy_rets = NULL) {
  rets <- nav_tbl |>
    dplyr::arrange(date) |>
    dplyr::mutate(daily_ret = nav / dplyr::lag(nav) - 1) |>
    dplyr::filter(!is.na(daily_ret)) |>
    dplyr::pull(daily_ret)

  ann_ret  <- prod(1 + rets)^(252 / length(rets)) - 1
  ann_vol  <- sd(rets) * sqrt(252)
  sharpe   <- ann_ret / ann_vol

  running_max <- cummax(nav_tbl$nav)
  drawdowns   <- nav_tbl$nav / running_max - 1
  max_dd      <- min(drawdowns)

  monthly_rets <- nav_tbl |>
    dplyr::mutate(ym = lubridate::floor_date(date, "month")) |>
    dplyr::group_by(ym) |>
    dplyr::summarise(m_ret = dplyr::last(nav) / dplyr::first(nav) - 1,
                     .groups = "drop")
  win_rate <- mean(monthly_rets$m_ret > 0, na.rm = TRUE)

  ir <- NA_real_
  if (!is.null(spy_rets)) {
    common_dates <- intersect(nav_tbl$date, spy_rets$date)
    strat_r <- nav_tbl |> dplyr::filter(date %in% common_dates) |>
      dplyr::arrange(date) |>
      dplyr::mutate(r = nav / dplyr::lag(nav) - 1) |>
      dplyr::filter(!is.na(r)) |> dplyr::pull(r)
    bench_r <- spy_rets |> dplyr::filter(date %in% common_dates) |>
      dplyr::arrange(date) |>
      dplyr::mutate(r = nav / dplyr::lag(nav) - 1) |>
      dplyr::filter(!is.na(r)) |> dplyr::pull(r)
    n_common <- min(length(strat_r), length(bench_r))
    if (n_common > 20L) {
      active_r <- strat_r[seq_len(n_common)] - bench_r[seq_len(n_common)]
      ir <- mean(active_r) / sd(active_r) * sqrt(252)
    }
  }

  tibble::tibble(
    Strategy        = label,
    `Ann. Return`   = ann_ret,
    `Ann. Vol`      = ann_vol,
    `Sharpe`        = sharpe,
    `Max Drawdown`  = max_dd,
    `Win Rate`      = win_rate,
    `Info Ratio`    = ir
  )
}

metrics_tbl <- dplyr::bind_rows(
  compute_metrics(bt_nav,    "Long-Short Strategy", spy_nav),
  compute_metrics(bt_nav_lo, "Long-Only Strategy",  spy_nav),
  compute_metrics(spy_nav,   "SPY Benchmark",        spy_nav)
)

metrics_tbl |>
  gt::gt() |>
  gt::tab_header(
    title    = "Training Period Performance Summary",
    subtitle = "June 2000 – December 2024"
  ) |>
  gt::fmt_percent(columns = c(`Ann. Return`, `Ann. Vol`,
                               `Max Drawdown`, `Win Rate`), decimals = 2) |>
  gt::fmt_number(columns  = c(`Sharpe`, `Info Ratio`), decimals = 3) |>
  gt::tab_style(
    style     = gt::cell_fill(color = "#dcfce7"),
    locations = gt::cells_body(rows = Strategy == "Long-Short Strategy")
  ) |>
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::sub_missing(missing_text = "—") |>
  gt::opt_table_font(font = gt::google_font("IBM Plex Sans")) |>
  gt::opt_stylize(style = 1, color = "blue")
```

## 8.3 Cumulative Returns — Strategy vs SPY

```{r bt-cumret-chart, fig.alt="Interactive plotly cumulative return chart with stress period shading"}
# ── Stress period boundaries ──────────────────────────────────────────────────
stress_periods <- tibble::tribble(
  ~label,                   ~start,               ~end,
  "GFC (2008–2009)",        "2008-09-01",          "2009-03-31",
  "COVID (2020)",           "2020-02-01",          "2020-06-30",
  "Rate Hikes (2022)",      "2022-01-01",          "2022-12-31"
) |>
  dplyr::mutate(
    start = lubridate::ymd(start),
    end   = lubridate::ymd(end)
  )

# ── Long data for plotly ──────────────────────────────────────────────────────
cumret_long <- dplyr::bind_rows(
  bt_nav    |> dplyr::transmute(date, cum_ret, series = "Long-Short"),
  bt_nav_lo |> dplyr::transmute(date, cum_ret, series = "Long-Only"),
  spy_nav   |> dplyr::transmute(date, cum_ret, series = "SPY")
)

series_colours <- c(`Long-Short` = "#2563EB", `Long-Only` = "#16a34a",
                    SPY = "#6b7280")

p_cum <- ggplot2::ggplot(
  cumret_long,
  ggplot2::aes(x = date, y = cum_ret, colour = series)
) +
  # Stress period shading
  purrr::pmap(stress_periods, function(label, start, end) {
    ggplot2::annotate("rect", xmin = start, xmax = end,
                      ymin = -Inf, ymax = Inf,
                      fill = "#fef08a", alpha = 0.35)
  }) +
  purrr::pmap(stress_periods, function(label, start, end) {
    ggplot2::annotate("text", x = start + (end - start) / 2,
                      y = max(cumret_long$cum_ret, na.rm = TRUE) * 0.95,
                      label = label, size = 2.8, colour = "#854d0e",
                      angle = 90, vjust = 0.5)
  }) +
  ggplot2::geom_line(linewidth = 0.6, alpha = 0.9) +
  ggplot2::scale_colour_manual(values = series_colours) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  ggplot2::labs(
    title   = "Cumulative Returns — Training Period (June 2000 – Dec 2024)",
    x       = NULL, y = "Cumulative Return",
    colour  = NULL,
    caption = "Yellow shading = stress periods: GFC, COVID, 2022 rate hikes"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold"),
                 panel.grid.minor = ggplot2::element_blank(),
                 legend.position  = "top")

plotly::ggplotly(p_cum, tooltip = c("colour", "x", "y")) |>
  plotly::layout(hovermode = "x unified",
                 legend = list(orientation = "h", y = 1.08))
```

## 8.4 Monthly Returns Heatmap

```{r bt-heatmap, fig.height=7, fig.alt="Monthly returns heatmap year by month"}
monthly_ret_tbl <- bt_nav |>
  dplyr::mutate(
    year  = lubridate::year(date),
    month = lubridate::month(date, label = TRUE, abbr = TRUE)
  ) |>
  dplyr::group_by(year, month) |>
  dplyr::summarise(
    m_ret = dplyr::last(nav) / dplyr::first(nav) - 1,
    .groups = "drop"
  )

ggplot2::ggplot(monthly_ret_tbl,
                ggplot2::aes(x = month, y = factor(year), fill = m_ret)) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.4) +
  ggplot2::geom_text(
    ggplot2::aes(
      label  = scales::percent(m_ret, accuracy = 0.1),
      colour = abs(m_ret) > 0.06
    ),
    size = 2.6, show.legend = FALSE
  ) +
  ggplot2::scale_fill_gradient2(
    low      = "#dc2626", mid = "white", high = "#16a34a",
    midpoint = 0,
    labels   = scales::percent_format(accuracy = 1)
  ) +
  ggplot2::scale_colour_manual(values = c(`TRUE` = "white", `FALSE` = "grey20")) +
  ggplot2::scale_y_discrete(limits = rev) +
  ggplot2::labs(
    title = "Monthly Returns Heatmap — Long-Short Strategy (Training)",
    x = NULL, y = NULL, fill = "Monthly\nReturn"
  ) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(
    plot.title       = ggplot2::element_text(face = "bold"),
    panel.grid       = ggplot2::element_blank(),
    axis.text.y      = ggplot2::element_text(size = 8),
    legend.position  = "right"
  )
```

## 8.5 Drawdown Chart & Long/Short Contribution

```{r bt-drawdown-contrib, fig.height=8, fig.alt="Patchwork of drawdown and long/short contribution charts"}
# ── Drawdown ──────────────────────────────────────────────────────────────────
drawdown_tbl <- bt_nav |>
  dplyr::arrange(date) |>
  dplyr::mutate(
    running_max = cummax(nav),
    drawdown    = nav / running_max - 1
  )

p_dd <- ggplot2::ggplot(drawdown_tbl,
                         ggplot2::aes(x = date, y = drawdown)) +
  ggplot2::geom_area(fill = "#fca5a5", alpha = 0.6) +
  ggplot2::geom_line(colour = "#dc2626", linewidth = 0.5) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  ggplot2::labs(
    title = "Strategy Drawdown — Training Period",
    x = NULL, y = "Drawdown from Peak"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold"),
                 panel.grid.minor = ggplot2::element_blank())

# ── Long / Short monthly contribution breakdown ───────────────────────────────
contrib_tbl <- signal_at_rebal |>
  dplyr::left_join(
    train_data |> dplyr::filter(symbol %in% SECTORS),
    by = c("symbol", "date")
  ) |>
  dplyr::group_by(date, position) |>
  dplyr::summarise(
    contrib = mean(
      dplyr::if_else(position != 0L,
        sign(position) * (price / dplyr::lag(price, default = NA) - 1),
        0),
      na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::filter(!is.na(contrib)) |>
  dplyr::mutate(
    book = dplyr::case_when(
      position ==  1L ~ "Long",
      position == -1L ~ "Short",
      TRUE            ~ "Cash"
    )
  ) |>
  dplyr::filter(book != "Cash")

p_contrib <- ggplot2::ggplot(
  contrib_tbl,
  ggplot2::aes(x = date, y = contrib, fill = book)
) +
  ggplot2::geom_col(position = "stack", width = 20) +
  ggplot2::scale_fill_manual(values = c(Long = "#16a34a", Short = "#dc2626")) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  ggplot2::scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  ggplot2::labs(
    title = "Long / Short Monthly Contribution — Training Period",
    x = NULL, y = "Average Daily Return Contribution",
    fill = "Book"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold"),
                 panel.grid.minor = ggplot2::element_blank(),
                 legend.position  = "top")

# ── patchwork layout ──────────────────────────────────────────────────────────
p_dd / p_contrib +
  patchwork::plot_annotation(
    title = "Risk & Attribution — Training Period Backtest",
    theme = ggplot2::theme(plot.title = ggplot2::element_text(face = "bold",
                                                               size = 13))
  )
```

---

# 9. Out-of-Sample Test {#sec-oos}

All model parameters — Kalman noise variances $W_{i,k}$ and $V_{i,k}$, the baseline threshold $\theta_{i,k}^{\text{base}}$, the feature scaling statistics, and the cluster centroid mapping — were estimated exclusively on the training period (June 2000 – December 2024) and are carried forward without modification. No re-estimation occurs on test data.

## 9.1 OOS Signal Generation

```{r oos-rebal-dates}
# ── End-of-month rebalancing dates within test period ─────────────────────────
rebal_dates_oos <- seq(
  lubridate::ceiling_date(TEST_START, "month"),
  TEST_END,
  by = "month"
) |>
  lubridate::rollback()

# ── Re-build feature matrix at each OOS rebalancing date ─────────────────────
# Uses all data up to each rd (training + OOS so far) — no look-ahead.
# Kalman W/V, theta_base, scaling params, and cluster centroids are all frozen
# from the training period; only the feature window advances.
oos_cluster_features <- purrr::map_dfr(SECTORS, function(sym) {
  sym_df <- test_data |> dplyr::filter(symbol == sym)
  purrr::map_dfr(rebal_dates_oos, function(rd) {
    # Use all data up to rd (training + test available so far) — no look-ahead
    full_sym <- dplyr::bind_rows(
      train_data |> dplyr::filter(symbol == sym),
      test_data  |> dplyr::filter(symbol == sym, date <= rd)
    )
    build_cluster_features(full_sym, rd)
  }) |>
    dplyr::mutate(symbol = sym)
}) |>
  dplyr::filter(dplyr::if_all(tidyselect::where(is.numeric), ~ !is.na(.x)))

# ── Predict cluster using frozen training centroids ───────────────────────────
# Bake OOS features with training recipe (no re-prepping)
oos_baked <- recipes::bake(clust_rec, new_data = oos_cluster_features |>
                             dplyr::select(roll_ret, roll_vol, roll_sr,
                                           mean_rsi, mean_macd))

# Assign to nearest centroid (Euclidean distance to km_fit$centers)
oos_cluster_num <- apply(oos_baked, 1L, function(row) {
  dists <- apply(km_fit$centers, 1L, function(ctr) sqrt(sum((row - ctr)^2)))
  which.min(dists)
})

oos_cluster_labels <- oos_cluster_features |>
  dplyr::mutate(
    cluster_num   = oos_cluster_num,
    cluster_label = dplyr::recode(
      as.character(cluster_num),
      !!!stats::setNames(label_map$cluster_label,
                         as.character(label_map$cluster_num))
    ),
    date = rebal_date
  ) |>
  dplyr::select(symbol, date, cluster_label)

# ── Build OOS signal snapshot at each rebalancing date ───────────────────────
signal_at_rebal_oos <- purrr::map_dfr(rebal_dates_oos, function(rd) {
  signals_tbl |>
    dplyr::filter(symbol %in% SECTORS, period == "test", date <= rd) |>
    dplyr::group_by(symbol) |>
    dplyr::slice_tail(n = 1L) |>
    dplyr::ungroup() |>
    dplyr::select(-cluster_label) |>
    dplyr::left_join(
      oos_cluster_labels |> dplyr::filter(date == rd),
      by = "symbol"
    ) |>
    dplyr::mutate(
      date         = rd,
      entry_signal = dplyr::case_when(
        cluster_label == "Growth"    & !is.na(delta) &
          delta >  theta_adaptive  ~ "long_entry",
        cluster_label == "Defensive" & !is.na(delta) &
          delta < -theta_adaptive  ~ "short_entry",
        TRUE                       ~ "no_entry"
      ),
      position = dplyr::case_when(
        entry_signal == "long_entry"  ~  1L,
        entry_signal == "short_entry" ~ -1L,
        TRUE                          ~  0L
      )
    )
})
```

## 9.2 OOS NAV Computation

```{r oos-nav}
# ── Run rebalance_portfolio() over OOS period ─────────────────────────────────
n_rebal_oos  <- length(rebal_dates_oos)
nav_list_oos <- vector("list", n_rebal_oos - 1L)

for (i in seq_len(n_rebal_oos - 1L)) {
  nav_list_oos[[i]] <- rebalance_portfolio(
    rebal_date      = rebal_dates_oos[i],
    next_rebal_date = rebal_dates_oos[i + 1L],
    signals_tbl     = signal_at_rebal_oos,
    price_data      = test_data |> dplyr::filter(symbol %in% SECTORS),
    prev_nav        = if (i == 1L) 1
                      else dplyr::last(nav_list_oos[[i - 1L]]$nav)
  )
}

oos_nav <- dplyr::bind_rows(nav_list_oos) |>
  dplyr::arrange(date) |>
  dplyr::mutate(cum_ret = nav / dplyr::first(nav) - 1)

# ── Long-only OOS ─────────────────────────────────────────────────────────────
signal_lo_oos <- signal_at_rebal_oos |>
  dplyr::mutate(position = dplyr::if_else(position == -1L, 0L, position))

nav_list_oos_lo <- vector("list", n_rebal_oos - 1L)
for (i in seq_len(n_rebal_oos - 1L)) {
  nav_list_oos_lo[[i]] <- rebalance_portfolio(
    rebal_date      = rebal_dates_oos[i],
    next_rebal_date = rebal_dates_oos[i + 1L],
    signals_tbl     = signal_lo_oos,
    price_data      = test_data |> dplyr::filter(symbol %in% SECTORS),
    prev_nav        = if (i == 1L) 1
                      else dplyr::last(nav_list_oos_lo[[i - 1L]]$nav)
  )
}

oos_nav_lo <- dplyr::bind_rows(nav_list_oos_lo) |>
  dplyr::arrange(date) |>
  dplyr::mutate(cum_ret = nav / dplyr::first(nav) - 1)

# ── SPY OOS benchmark ─────────────────────────────────────────────────────────
spy_nav_oos <- test_data |>
  dplyr::filter(symbol == "SPY",
                date >= min(oos_nav$date), date <= max(oos_nav$date)) |>
  dplyr::arrange(date) |>
  dplyr::mutate(nav = price / dplyr::first(price), cum_ret = nav - 1)
```

## 9.3 Performance Metrics

```{r oos-metrics}
# ── OOS metrics using same compute_metrics() helper from Section 8 ─────────────
oos_metrics <- dplyr::bind_rows(
  compute_metrics(oos_nav,    "Long-Short (OOS)", spy_nav_oos),
  compute_metrics(oos_nav_lo, "Long-Only (OOS)",  spy_nav_oos),
  compute_metrics(spy_nav_oos,"SPY (OOS)",         spy_nav_oos)
)

# ── Side-by-side training vs OOS ─────────────────────────────────────────────
train_metrics_compare <- dplyr::bind_rows(
  compute_metrics(bt_nav,    "Long-Short", spy_nav),
  compute_metrics(spy_nav,   "SPY",        spy_nav)
) |>
  dplyr::rename_with(~ paste0(.x, " (Train)"), -Strategy)

oos_metrics_compare <- dplyr::bind_rows(
  compute_metrics(oos_nav,    "Long-Short", spy_nav_oos),
  compute_metrics(spy_nav_oos,"SPY",        spy_nav_oos)
) |>
  dplyr::rename_with(~ paste0(.x, " (OOS)"), -Strategy)

dplyr::left_join(train_metrics_compare, oos_metrics_compare, by = "Strategy") |>
  gt::gt() |>
  gt::tab_header(
    title    = "Training vs Out-of-Sample Performance Comparison",
    subtitle = "Train: Jun 2000–Dec 2024  |  OOS: Jan 2025–Mar 2026"
  ) |>
  gt::fmt_percent(
    columns = tidyselect::contains(c("Return", "Vol", "Drawdown", "Win")),
    decimals = 2
  ) |>
  gt::fmt_number(
    columns = tidyselect::contains(c("Sharpe", "Info")),
    decimals = 3
  ) |>
  gt::tab_style(
    style     = gt::cell_fill(color = "#dcfce7"),
    locations = gt::cells_body(rows = Strategy == "Long-Short")
  ) |>
  gt::tab_spanner(label = "Training Period",
                  columns = tidyselect::ends_with("(Train)")) |>
  gt::tab_spanner(label = "Out-of-Sample Period",
                  columns = tidyselect::ends_with("(OOS)")) |>
  gt::sub_missing(missing_text = "—") |>
  gt::opt_table_font(font = gt::google_font("IBM Plex Sans")) |>
  gt::opt_stylize(style = 1, color = "blue")
```

## 9.4 OOS Cumulative Returns

```{r oos-cumret, fig.alt="Interactive OOS cumulative return chart vs SPY"}
oos_long <- dplyr::bind_rows(
  oos_nav    |> dplyr::transmute(date, cum_ret, series = "Long-Short (OOS)"),
  oos_nav_lo |> dplyr::transmute(date, cum_ret, series = "Long-Only (OOS)"),
  spy_nav_oos|> dplyr::transmute(date, cum_ret, series = "SPY (OOS)")
)

p_oos_cum <- ggplot2::ggplot(
  oos_long,
  ggplot2::aes(x = date, y = cum_ret, colour = series)
) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::scale_colour_manual(
    values = c(`Long-Short (OOS)` = "#2563EB",
               `Long-Only (OOS)`  = "#16a34a",
               `SPY (OOS)`        = "#6b7280")
  ) +
  ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggplot2::scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  ggplot2::labs(
    title   = "OOS Cumulative Returns — Jan 2025 to Mar 2026",
    x = NULL, y = "Cumulative Return", colour = NULL
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold"),
                 panel.grid.minor = ggplot2::element_blank(),
                 legend.position  = "top")

plotly::ggplotly(p_oos_cum, tooltip = c("colour", "x", "y")) |>
  plotly::layout(hovermode = "x unified",
                 legend = list(orientation = "h", y = 1.1))
```

## 9.5 Training vs OOS Side-by-Side

```{r oos-vs-train, fig.height=5, fig.alt="Patchwork of training and OOS cumulative return charts side by side"}
# ── Normalise training returns for visual comparability ───────────────────────
train_long <- dplyr::bind_rows(
  bt_nav  |> dplyr::transmute(date, cum_ret, series = "Long-Short"),
  spy_nav |> dplyr::transmute(date, cum_ret, series = "SPY")
)

make_cumret_panel <- function(df, title_str) {
  ggplot2::ggplot(df, ggplot2::aes(x = date, y = cum_ret, colour = series)) +
    ggplot2::geom_line(linewidth = 0.6) +
    ggplot2::scale_colour_manual(
      values = c(`Long-Short` = "#2563EB", SPY = "#6b7280",
                 `Long-Short (OOS)` = "#2563EB", `SPY (OOS)` = "#6b7280")
    ) +
    ggplot2::scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    ggplot2::scale_x_date(date_labels = "%Y") +
    ggplot2::labs(title = title_str, x = NULL, y = "Cumulative Return",
                  colour = NULL) +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::theme(plot.title       = ggplot2::element_text(face = "bold"),
                   panel.grid.minor = ggplot2::element_blank(),
                   legend.position  = "bottom")
}

p_train_panel <- make_cumret_panel(train_long,
                                   "Training (Jun 2000 – Dec 2024)")
p_oos_panel   <- make_cumret_panel(
  oos_long |>
    dplyr::filter(series %in% c("Long-Short (OOS)", "SPY (OOS)")) |>
    dplyr::mutate(series = stringr::str_remove(series, " \\(OOS\\)")),
  "Out-of-Sample (Jan 2025 – Mar 2026)"
)

p_train_panel | p_oos_panel
```

## 9.6 OOS Monthly Returns Heatmap

```{r oos-heatmap, fig.alt="OOS monthly returns heatmap"}
oos_monthly <- oos_nav |>
  dplyr::mutate(
    year  = lubridate::year(date),
    month = lubridate::month(date, label = TRUE, abbr = TRUE)
  ) |>
  dplyr::group_by(year, month) |>
  dplyr::summarise(m_ret = dplyr::last(nav) / dplyr::first(nav) - 1,
                   .groups = "drop")

ggplot2::ggplot(oos_monthly,
                ggplot2::aes(x = month, y = factor(year), fill = m_ret)) +
  ggplot2::geom_tile(colour = "white", linewidth = 0.5) +
  ggplot2::geom_text(
    ggplot2::aes(
      label  = scales::percent(m_ret, accuracy = 0.1),
      colour = abs(m_ret) > 0.04
    ),
    size = 3.2, show.legend = FALSE
  ) +
  ggplot2::scale_fill_gradient2(
    low = "#dc2626", mid = "white", high = "#16a34a", midpoint = 0,
    labels = scales::percent_format(accuracy = 1)
  ) +
  ggplot2::scale_colour_manual(values = c(`TRUE` = "white",
                                          `FALSE` = "grey20")) +
  ggplot2::scale_y_discrete(limits = rev) +
  ggplot2::labs(
    title = "Monthly Returns Heatmap — OOS Period",
    x = NULL, y = NULL, fill = "Monthly\nReturn"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(plot.title    = ggplot2::element_text(face = "bold"),
                 panel.grid    = ggplot2::element_blank())
```

## 9.7 Honest Assessment

The out-of-sample period (January 2025 – March 2026) presents a genuinely short window — fifteen months — which limits the statistical power of any performance comparison. Several structural factors are worth examining honestly.

**Cluster stability in the OOS period.** The k-means centroids were frozen at end-2024. If the cross-sectional dispersion of sector returns compressed or reversed in 2025 — as might occur in a late-cycle or policy-driven environment — the frozen centroids will mis-classify regimes. The stability analysis in Section 5 showed that post-COVID centroids already drifted relative to the pre-2020 period; a further drift in 2025 would compound this. Any month where the average silhouette width falls below 0.25 should be treated as a low-confidence cluster assignment, and positions taken in those months should be interpreted cautiously.

**Monthly rebalancing lag.** The strategy is structurally slow to react. Entry signals are only evaluated at month-end; exit signals via the innovation z-score operate daily but require 63 consecutive days of innovation data to produce a reliable z-score. In fast-moving markets — a sudden Fed pivot, a geopolitical shock — a position can be held through a significant drawdown before the exit triggers. The 2025 environment, characterised by elevated rate uncertainty and occasional sharp sector rotations, is precisely the regime where this lag is most costly.

**Value added by the short book.** The comparison between the long-short and long-only versions in the OOS period directly tests whether shorting the Defensive cluster added or subtracted value. If Defensive sectors (XLU, XLP, XLRE) performed well in 2025 — which is plausible in a risk-off or rate-cut environment — the short book would have detracted. The metrics table above shows the directional result; the long-only version is the appropriate counterfactual.

**What would need to be true for the strategy to work OOS.** The strategy's positive expectation relies on three conditions holding: (1) sector return dispersion remains wide enough for K-means to find distinct clusters, (2) the Kalman filtered signal leads price moves rather than lagging them, and (3) the short book's Defensive ETFs genuinely underperform during Growth regimes. All three are empirical claims that should be re-examined with each year of new data — not maintained by assumption.

---

# 10. Circle Back {#sec-circle-back}

## 10.1 What the Strategy Learned

The document opened with a single question: why should an investor pay for active management instead of simply holding SPY? Nine sections later, the honest answer is that the strategy earns that justification in some market conditions and not others — and understanding the boundary between those conditions is itself the most valuable output of the analysis.

The clustering exercise revealed something non-trivial about the structure of sector returns. The Growth / Neutral / Defensive taxonomy was not imposed arbitrarily; it emerged from the data. The k-means algorithm consistently separated a high-return, moderate-volatility group of sectors (technology, consumer discretionary, communication services) from a low-return, low-volatility group (utilities, real estate, consumer staples) with an intermediate cluster in between. This structure proved reasonably stable across the pre-COVID sub-period (2000–2019) and the post-COVID sub-period (2020–2024), though the centroids drifted — the post-COVID Growth cluster moved to higher return space, and the Defensive cluster became more tightly defined. The three-cluster solution found a genuine economic signal: sectors do rotate across regimes, and that rotation is partially predictable from lagged price behaviour.

The Kalman filter served a purpose beyond mere smoothing. The innovation sequence — the filter's one-step-ahead prediction error — carried information that the raw technical indicators did not. When the z-score of the innovation sequence spiked beyond two standard deviations, the filter was encountering observations that its learned model of the world did not anticipate. In the training period, the majority of these spikes preceded or coincided with cluster flips, suggesting that the innovation exit captured regime transitions earlier than the monthly cluster re-assignment could. This is the strategy's most technically differentiated feature: it uses the internal mechanics of the filter, not just its output, as a risk signal.

## 10.2 When Clustering Fails

No statistical model is unconditionally valid, and k-means clustering on sector ETFs is particularly fragile under three identifiable conditions.

The first is extreme volatility. When the CBOE VIX index exceeds approximately 40 — as it did in September 2008, March 2020, and briefly in early 2022 — cross-sector correlations converge sharply toward 1.0. Every sector falls simultaneously, and the five-dimensional feature space that normally separates Growth from Defensive collapses to a single dimension: magnitude of loss. In this environment, the average silhouette width typically falls below 0.25, signalling that cluster assignments are essentially random. A strategy that enters short positions in Defensive ETFs during a market panic will find that those ETFs also collapse, destroying the thesis of the short book.

The second failure condition is structural breaks in sector relationships. The AI revolution that accelerated after 2022 fundamentally altered the return dynamics of the technology sector and, by extension, its correlations with every other sector. A clustering model trained on data through 2024 implicitly assumes that XLK's relationship to XLU in 2025 resembles its historical relationship. If that assumption fails — because technology has become a macro factor rather than a sector — the cluster centroids are anchored to a world that no longer exists.

The third condition is regime speed. The strategy rebalances monthly and evaluates cluster flips monthly. A regime transition that unfolds over days rather than weeks — the March 2020 COVID crash took approximately 23 trading days from peak to trough — will generate a cluster flip signal only after the damage is done. The innovation z-score exit provides some daily-frequency protection, but it requires 63 days of innovation history to compute a reliable z-score, meaning the first two months of any new regime are partially blind.

## 10.3 Limitations

Several limitations constrain the practical interpretation of the results and would need to be addressed before any live implementation.

Transaction costs and market impact are absent from the backtest. The strategy rebalances eleven ETF positions monthly and may exit individual positions intra-month on innovation signals. In liquid ETFs this cost is small per trade, but across a 24-year backtest with frequent rebalancing it is not negligible. A conservative estimate of 5 basis points per one-way trade — low given the liquidity of SPDR sector ETFs — would meaningfully reduce the reported Sharpe ratio, particularly in periods with high turnover.

Short selling is treated as costless. In practice, the borrow cost for sector ETFs varies from negligible to prohibitive depending on short interest at any given time. During the periods when the strategy most wants to be short Defensive ETFs — specifically during bear markets — those ETFs may carry elevated borrow costs precisely because other investors have the same idea. The strategy's information ratio relative to SPY should be understood as an upper bound that will compress in any realistic implementation.

The single-model structure creates concentration risk. K-means with $k = 3$ is one of many possible clustering approaches. Different distance metrics, different initialisation seeds (though controlled here with `set.seed(451)`), or different numbers of clusters would produce different regime assignments. The strategy does not quantify uncertainty across clustering models — it commits to a single solution and acts on it. A more robust implementation would average signals across multiple cluster specifications and reduce position size in months where the solutions disagree.

Finally, the look-ahead audit: the most dangerous form of bias in backtesting is subtle. The Kalman parameters $W_{i,k}$ and $V_{i,k}$ were estimated on training data and frozen — this is correct. The feature scaling statistics and cluster centroids were similarly frozen — this is correct. The one remaining risk is the `theta_base` calculation, which uses the median absolute delta from the training period and is applied to test-period deltas. This is a valid use of training-period statistics applied to test data, but it assumes that the distribution of deltas is stationary across the two periods. If the test-period filtered signals are substantially more volatile than the training period — possible in a high-uncertainty macro environment — `theta_base` will be set too low and generate excessive false entries.

## 10.4 Inversion: Five Ways This Strategy Fails

Inversion is more revealing than optimisation. Rather than asking how to maximise returns, the more useful question is: what would have to be true for this strategy to reliably destroy capital? Five conditions are sufficient.

First, a persistent low-dispersion regime in which all eleven sector ETFs move together for an extended period. In such an environment the cross-sectional variance of the feature matrix collapses, silhouette widths approach zero, and the three-cluster solution becomes arbitrary. The strategy would generate random signals, and transaction costs would erode capital steadily.

Second, a sustained inverted-return relationship between the Growth and Defensive clusters. If utilities and consumer staples deliver higher returns than technology and consumer discretionary over a multi-year period — which occurred partially during the 2022 rate-hike cycle — the short book becomes the return source while the long book destroys value. The strategy would lose on both sides if the cluster labels fail to update fast enough to reflect the inversion.

Third, a structural break in the Kalman filter's learned model. If the statistical relationship between indicator levels and their filtered states changes — because high-frequency trading algorithms have permanently altered the autocorrelation structure of RSI or MACD — the filter's innovation sequence will be persistently non-zero and will trigger false exits constantly, churning the portfolio and incurring transaction costs without protective benefit.

Fourth, concentrated crowding in sector ETF trades. If a sufficiently large number of quantitative managers implement sector rotation strategies on the same eleven ETFs using similar signals, the entry and exit points the strategy identifies will already be priced in. The strategy profits on the existence of a systematic relationship between signals and subsequent returns; if that relationship is arbitraged away by crowding, the edge disappears even if the signals remain technically correct.

Fifth, regulatory or structural changes to the short-selling market. A sustained restriction on borrowing Defensive ETFs — analogous to the short-sale restrictions imposed on financial stocks in 2008 — would prevent the strategy from expressing its Defensive-cluster thesis, leaving it holding only a long book with no hedge. This is not a theoretical risk; it has happened in precisely the regime type that triggers the short-book signals.

## 10.5 What Would Change in a Rebuild

If building this strategy from scratch with the benefit of hindsight, three changes would improve robustness without sacrificing the core thesis.

The first is regime uncertainty quantification. Instead of a single k-means solution, a Gaussian Mixture Model would provide cluster membership probabilities rather than hard assignments. A month in which XLK has 0.9 probability of belonging to the Growth cluster is fundamentally different from a month in which the probability is 0.6. Position sizes should scale with this confidence. Scott Page's concept of diversity in forecasting models applies directly here: averaging across multiple clustering approaches — k-means, GMM, hierarchical clustering — would reduce the variance of the regime signal without necessarily reducing its mean.

The second change is signal frequency. Monthly rebalancing is appropriate for a strategy whose primary driver is regime-level behaviour, but the intra-month innovation exit points to the value of a mixed-frequency architecture. The entry and exit rules could be evaluated at weekly frequency — retaining the monthly cluster assignment as the regime prior but allowing weekly reassessment of whether the delta signal still supports the position. This would reduce the lag cost identified in Sections 9 and 10.2 without abandoning the deliberate, low-turnover character of the strategy.

The third change is an explicit regime-quality filter. A silhouette width below 0.25 should suppress all trading signals for that month, returning the strategy to a cash position regardless of what the cluster label says. Many models agree: if the clustering is ambiguous, the signal is uninformative, and the correct action is to do nothing. The course framework's emphasis on knowing when your model is not applicable is more valuable than any specific model improvement.

## 10.6 Connection to the Course Framework

This document followed the Frame → Sketch → Implement → Circle Back structure throughout. Section 1 framed the problem — why active management, and under what conditions does it create value. Sections 2 and 7 sketched the decision architecture before any code was written, forcing a commitment to the rules in plain language and mathematical notation before seeing any results. Sections 3 through 6 implemented the strategy sequentially, with each section producing outputs that fed the next without crossing the look-ahead boundary. Sections 8 and 9 evaluated the implementation honestly against an observable benchmark, reporting both the successes and the failures. This section circles back to the original frame.

The course's many-models philosophy is visible throughout the strategy's architecture. The Kalman filter is one model of how indicators evolve; the k-means clustering is a separate model of how regimes partition the sector space; the innovation z-score is a third model of when the first model is failing. No single model is trusted unconditionally. When they agree — delta above threshold, cluster in Growth, innovation z-score within normal range — the signal is strong. When they disagree — cluster says Growth but innovation z-score is spiking — the exit overrides the entry, which is the correct priority ordering under uncertainty.

The final answer to the opening question is this: the strategy justifies active management over holding SPY during periods of high cross-sector dispersion, stable regime structure, and moderate volatility — broadly, the kind of market that prevailed for much of the 2003–2007 and 2013–2019 periods. It does not justify active management during crises, during structural breaks, or during the first fifteen months of a new macro regime when the cluster model is most likely to be stale. An intellectually honest investment mandate would state this explicitly, suspend trading when silhouette widths fall below the quality threshold, and measure performance only over the regimes in which the strategy was designed to operate.

---

# References {#sec-references}

<!-- ─────────────────────────────────────────────────────────────────────────
     REFERENCES
     Required citations:
       • Kalman (1960) — original Kalman filter paper
       • Harvey (1989) — Time Series Models, state space in econometrics
       • Durbin & Koopman (2012) — State Space Methods by State Space
       • Hartigan & Wong (1979) — K-means algorithm
       • Avellaneda & Lee (2010) — stat arb, state space in finance
     ──────────────────────────────────────────────────────────────────────── -->

<!-- TODO: Create references.bib with all required citations -->
<!-- TODO: Re-add  bibliography: references.bib  to YAML header once .bib file exists -->
<!-- TODO: Restore ::: {#refs} ::: div once bibliography is wired up -->
